\chapter[THE TENTH STAGE]{}

Then Christian addressed himself thus to his fellow:

Christian: Well, come, my good Hopeful, I perceive that thou and I must walk by ourselves again.

So I saw in my dream, that they went on apace before, and Ignorance he came hobbling after. Then said Christian to his companion, I much pity this poor man: it will certainly go ill with him at last.

Hopeful: Alas! there are abundance in our town in his condition, whole families, yea, whole streets, and that of pilgrims too; and if there be so many in our parts, how many, think you, must there be in the place where he was born?

Christian: Indeed, the word saith, ``He hath blinded their eyes, lest they should see," etc.

But, now we are by ourselves, what do you think of such men? Have they at no time, think you, convictions of sin, and so, consequently, fears that their state is dangerous?

Hopeful: Nay, do you answer that question yourself, for you are the elder man.

Christian: Then I say, sometimes (as I think) they may; but they being naturally ignorant, understand not that such convictions tend to their good; and therefore they do desperately seek to stifle them, and presumptuously continue to flatter themselves in the way of their own hearts.

Hopeful: I do believe, as you say, that fear tends much to men's good, and to make them right at their beginning to go on pilgrimage.

Christian: Without all doubt it doth, if it be right; for so says the word, ``The fear of the Lord is the beginning of wisdom." Job 28:28; Psalm 111:10; Prov. 1:7; 9:10.

Hopeful: How will you describe right fear?

Christian: True or right fear is discovered by three things:

1. By its rise; it is caused by saving convictions for sin.

2. It driveth the soul to lay fast hold of Christ for salvation.

3. It begetteth and continueth in the soul a great reverence of God, his word, and ways; keeping it tender, and making it afraid to turn from them, to the right hand or to the left, to any thing that may dishonor God, break its peace, grieve the Spirit, or cause the enemy to speak reproachfully.

Hopeful: Well said; I believe you have said the truth. Are we now almost got past the Enchanted Ground?

Christian: Why? are you weary of this discourse?

Hopeful: No, verily, but that I would know where we are.

Christian: We have not now above two miles further to go thereon. But let us return to our matter.

Now, the ignorant know not that such conviction as tend to put them in fear, are for their good, and therefore they seek to stifle them.

Hopeful: How do they seek to stifle them?

Christian: 1. They think that those fears are wrought by the devil, (though indeed they are wrought of God,) and thinking so, they resist them, as things that directly tend to their overthrow. 2. They also think that these fears tend to the spoiling of their faith; when, alas for them, poor men that they are, they have none at all; and therefore they harden their hearts against them. 3. They presume they ought not to fear, and therefore, in despite of them, wax presumptuously confident. 4. They see that those fears tend to take away from them their pitiful old self-holiness, and therefore they resist them with all their might.

Hopeful: I know something of this myself; for before I knew myself it was so with me.

Christian: Well, we will leave, at this time, our neighbor Ignorance by himself, and fall upon another profitable question.

Hopeful: With all my heart; but you shall still begin.

Christian: Well then, did you not know, about ten years ago, one Temporary in your parts, who was a forward man in religion then?

Hopeful: Know him! yes; he dwelt in Graceless, a town about two miles off of Honesty, and he dwelt next door to one Turnback.

Christian: Right; he dwelt under the same roof with him. Well, that man was much awakened once: I believe that then he had some sight of his sins, and of the wages that were due thereto.

Hopeful: I am of your mind, for (my house not being above three miles from him) he would oft-times come to me, and that with many tears. Truly I pitied the man, and was not altogether without hope of him; but one may see, it is not every one that cries, ``Lord, Lord!"

Christian: He told me once that he was resolved to go on pilgrimage, as we go now; but all of a sudden he grew acquainted with one Save-self, and then he became a stranger to me.

Hopeful: Now, since we are talking about him, let us a little inquire into the reason of the sudden backsliding of him and such others.

Christian: It may be very profitable; but do you begin.

Hopeful: Well, then, there are, in my judgment, four reasons for it:

1. Though the consciences of such men are awakened, yet their minds are not changed: therefore, when the power of guilt weareth away, that which provoked them to be religious ceaseth; wherefore they naturally turn to their own course again; even as we see the dog that is sick of what he hath eaten, so long as his sickness prevails, he vomits and casts up all; not that he doth this of a free mind, (if we may say a dog has a mind,) but because it troubleth his stomach: but now, when his sickness is over, and so his stomach eased, his desires being not at all alienated from his vomit, he turns him about, and licks up all; and so it is true which is written, ``The dog is turned to his own vomit again." 2 Pet. 2:22. Thus, I say, being hot for heaven, by virtue only of the sense and fear of the torments of hell, as their sense and fear of damnation chills and cools, so their desires for heaven and salvation cool also. So then it comes to pass, that when their guilt and fear is gone, their desires for heaven and happiness die, and they return to their course again.

2. Another reason is, they have slavish fears that do overmaster them: I speak now of the fears that they have of men; ``For the fear of man bringeth a snare." Prov. 29:25. So then, though they seem to be hot for heaven so long as the flames of hell are about their ears, yet, when that terror is a little over, they betake themselves to second thoughts, namely, that it is good to be wise and not to run (for they know not what) the hazard of losing all, or at least of bringing themselves into unavoidable and unnecessary troubles; and so they fall in with the world again.

3. The shame that attends religion lies also as a block in their way: they are proud and haughty, and religion in their eye is low and contemptible: therefore when they have lost their sense of hell and the wrath to come, they return again to their former course.

4. Guilt, and to meditate terror, are grievous to them; they like not to see their misery before they come into it; though perhaps the sight of at it first, if they loved that sight, might make them fly whither the righteous fly and are safe; but because they do, as I hinted before, even shun the thoughts of guilt and terror, therefore, when once they are rid of their awakenings about the terrors and wrath of God, they harden their hearts gladly, and choose such ways as will harden them more and more.

Christian: You are pretty near the business, for the bottom of all is for want of a change in their mind and will. And therefore they are but like the felon that standeth before the judge: he quakes and trembles, and seems to repent most heartily, but the bottom of all is the fear of the halter: not that he hath any detestation of the offence, as it is evident; because, let but this man have his liberty, and he will be a thief, and so a rogue still; whereas, if his mind was changed, he would be otherwise.

Hopeful: Now I have showed you the reason of their going back, do you show me the manner thereof.

Christian: So I will willingly.

1. They draw off their thoughts, all that they may, from the remembrance of God, death, and judgment to come.

2. Then they cast off by degrees private duties, as closet prayer, curbing their lusts, watching, sorrow for sin, and the like.

3. Then they shun the company of lively and warm Christians.

4. After that, they grow cold to public duty, as hearing, reading, godly conference, and the like.

5. They then begin to pick holes, as we say, in the coats of some of the godly, and that devilishly, that they may have a seeming color to throw religion (for the sake of some infirmities they have espied in them) behind their backs.

6. Then they begin to adhere to, and associate themselves with, carnal, loose, and wanton men.

7. Then they give way to carnal and wanton discourses in secret; and glad are they if they can see such things in any that are counted honest, that they may the more boldly do it through their example.

8. After this they begin to play with little sins openly.

9. And then, being hardened, they show themselves as they are. Thus, being launched again into the gulf of misery, unless a miracle of grace prevent it, they everlastingly perish in their own deceivings.

Now I saw in my dream, that by this time the pilgrims were got over the Enchanted Ground, and entering into the country of Beulah, whose air was very sweet and pleasant, Isaiah 62:4-12; Song 2:10-12; the way lying directly through it, they solaced themselves there for a season. Yea, here they heard continually the singing of birds, and saw every day the flowers appear in the earth, and heard the voice of the turtle in the land. In this country the sun shineth night and day: wherefore this was beyond the Valley of the Shadow of Death, and also out of the reach of Giant Despair; neither could they from this place so much as see Doubting Castle. Here they were within sight of the city they were going to; also here met them some of the inhabitants thereof; for in this land the shining ones commonly walked, because it was upon the borders of heaven. In this land also the contract between the Bride and the Bridegroom was renewed; yea, here, ``as the bridegroom rejoiceth over the bride, so doth God rejoice over them." Here they had no want of corn and wine; for in this place they met with abundance of what they had sought for in all their pilgrimage. Here they heard voices from out of the city, loud voices, saying, ``Say ye to the daughter of Zion, Behold, thy salvation cometh! Behold, his reward is with him!" Here all the inhabitants of the country called them ``the holy People, the redeemed of the Lord, sought out," etc.

Now, as they walked in this land, they had more rejoicing than in parts more remote from the kingdom to which they were bound; and drawing near to the city, they had yet a more perfect view thereof: It was builded of pearls and precious stones, also the streets thereof were paved with gold; so that, by reason of the natural glory of the city, and the reflection of the sunbeams upon it, Christian with desire fell sick; Hopeful also had a fit or two of the same disease: wherefore here they lay by it a while, crying out because of their pangs, ``If you see my Beloved, tell him that I am sick of love."

But, being a little strengthened, and better able to bear their sickness, they walked on their way, and came yet nearer and nearer, where were orchards, vineyards, and gardens, and their gates opened into the highway. Now, as they came up to these places, behold the gardener stood in the way; to whom the pilgrims said, Whose goodly vineyards and gardens are these? He answered, they are the King's, and are planted here for his own delight, and also for the solace of pilgrims. So the gardener had them into the vineyards, and bid them refresh themselves with the dainties, Deut. 23:24; he also showed them there the King's walks and arbors where he delighted to be: And here they tarried and slept.

Now I beheld in my dream, that they talked more in their sleep at this time than ever they did in all their journey; and, being in a muse thereabout, the gardener said even to me, Wherefore musest thou at the matter? It is the nature of the fruit of the grapes of these vineyards, ``to go down so sweetly as to cause the lips of them that are asleep to speak." Song 7:9.

So I saw that when they awoke, they addressed themselves to go up to the city. But, as I said, the reflection of the sun upon the city (for the city was pure gold, Rev. 21:18,) was so extremely glorious, that they could not as yet with open face behold it, but through an instrument made for that purpose. 2 Cor. 3:18. So I saw, that as they went on, there met them two men in raiment that shone like gold, also their faces shone as the light.

These men asked the pilgrims whence they came; and they told them. They also asked them where they had lodged, what difficulties and dangers, what comforts and pleasures, they had met with in the way; and they told them. Then said the men that met them, You have but two difficulties more to meet with, and then you are in the City.

Christian then and his companion asked the men to go along with them: so they told them that they would; But, said they, you must obtain it by your own faith. So I saw in my dream, that they went on together till they came in sight of the gate.

Now I further saw, that betwixt them and the gate was a river; but there was no bridge to go over, and the river was very deep. At the sight, therefore, of this river the pilgrims were much stunned; but the men that went with them said, You must go through, or you cannot come at the gate.

The pilgrims then began to inquire if there was no other way to the gate. To which they answered, Yes; but there hath not any, save two, to wit, Enoch and Elijah, been permitted to tread that path since the foundation of the world, nor shall until the last trumpet shall sound. The pilgrims then, especially Christian, began to despond in their mind, and looked this way and that, but no way could be found by them by which they might escape the river. Then they asked the men if the waters were all of a depth. They said, No; yet they could not help them in that case; for, said they, you shall find it deeper or shallower as you believe in the King of the place.

Then they addressed themselves to the water, and entering, Christian began to sink, and crying out to his good friend Hopeful, he said, I sink in deep waters; the billows go over my head; all his waves go over me. Selah.

Then said the other, Be of good cheer, my brother: I feel the bottom, and it is good. Then said Christian, Ah! my friend, the sorrows of death have compassed me about, I shall not see the land that flows with milk and honey. And with that a great darkness and horror fell upon Christian, so that he could not see before him. Also here he in a great measure lost his senses, so that he could neither remember nor orderly talk of any of those sweet refreshments that he had met with in the way of his pilgrimage. But all the words that he spoke still tended to discover that he had horror of mind, and heart-fears that he should die in that river, and never obtain entrance in at the gate. Here also, as they that stood by perceived, he was much in the troublesome thoughts of the sins that he had committed, both since and before he began to be a pilgrim. It was also observed that he was troubled with apparitions of hobgoblins and evil spirits; for ever and anon he would intimate so much by words.

Hopeful therefore here had much ado to keep his brother's head above water; yea, sometimes he would be quite gone down, and then, ere a while, he would rise up again half dead. Hopeful did also endeavor to comfort him, saying, Brother, I see the gate, and men standing by to receive us; but Christian would answer, It is you, it is you they wait for; for you have been hopeful ever since I knew you. And so have you, said he to Christian. Ah, brother, (said he,) surely if I was right he would now arise to help me; but for my sins he hath brought me into the snare, and hath left me. Then said Hopeful, My brother, you have quite forgot the text where it is said of the wicked, ``There are no bands in their death, but their strength is firm; they are not troubled as other men, neither are they plagued like other men." Psa. 73:4,5. These troubles and distresses that you go through in these waters, are no sign that God hath forsaken you; but are sent to try you, whether you will call to mind that which heretofore you have received of his goodness, and live upon him in your distresses.

Then I saw in my dream, that Christian was in a muse a while. To whom also Hopeful added these words, Be of good cheer, Jesus Christ maketh thee whole. And with that Christian brake out with a loud voice, Oh, I see him again; and he tells me, ``When thou passest through the waters, I will be with thee; and through the rivers, they shall not overflow thee." Isa. 43:2. Then they both took courage, and the enemy was after that as still as a stone, until they were gone over. Christian, therefore, presently found ground to stand upon, and so it followed that the rest of the river was but shallow. Thus they got over.

Now, upon the bank of the river, on the other side, they saw the two shining men again, who there waited for them. Wherefore, being come out of the river, they saluted them, saying, We are ministering spirits, sent forth to minister for those that shall be the heirs of salvation. Thus they went along towards the gate.

Now you must note, that the city stood upon a mighty hill; but the pilgrims went up that hill with ease, because they had these two men to lead them up by the arms: they had likewise left their mortal garments behind them in the river; for though they went in with them, they came out without them. They therefore went up here with much agility and speed, though the foundation upon which the city was framed was higher than the clouds; they therefore went up through the region of the air, sweetly talking as they went, being comforted because they safely got over the river, and had such glorious companions to attend them.

The talk that they had with the shining ones was about the glory of the place; who told them that the beauty and glory of it was inexpressible. There, said they, is ``Mount Sion, the heavenly Jerusalem, the innumerable company of angels, and the spirits of just men made perfect." Heb. 12:22-24. You are going now, said they, to the paradise of God, wherein you shall see the tree of life, and eat of the never-fading fruits thereof: and when you come there you shall have white robes given you, and your walk and talk shall be every day with the King, even all the days of eternity. Rev. 2:7; 3:4,5; 22:5. There you shall not see again such things as you saw when you were in the lower region upon earth; to wit, sorrow, sickness, affliction, and death; ``For the former things are passed away." Rev. 21:4. You are going now to Abraham, to Isaac, and Jacob, and to the prophets, men that God hath taken away from the evil to come, and that are now ``resting upon their beds, each one walking in his righteousness." The men then asked, What must we do in the holy place? To whom it was answered, You must there receive the comfort of all your toil, and have joy for all your sorrow; you must reap what you have sown, even the fruit of all your prayers, and tears, and sufferings for the King by the way. Gal. 6:7,8. In that place you must wear crowns of gold, and enjoy the perpetual sight and vision of the Holy One; for ``there you shall see him as he is." 1 John, 3:2. There also you shall serve him continually with praise, with shouting and thanksgiving, whom you desired to serve in the world, though with much difficulty, because of the infirmity of your flesh. There your eyes shall be delighted with seeing, and your ears with hearing the pleasant voice of the Mighty One. There you shall enjoy your friends again that are gone thither before you; and there you shall with joy receive even every one that follows into the holy place after you. There also you shall be clothed with glory and majesty, and put into an equipage fit to ride out with the King of Glory. When he shall come with sound of trumpet in the clouds, as upon the wings of the wind, you shall come with him; and when he shall sit upon the throne of judgment, you shall sit by him; yea, and when he shall pass sentence upon all the workers of iniquity, let them be angels or men, you also shall have a voice in that judgment, because they were his and your enemies. Also, when he shall again return to the city, you shall go too with sound of trumpet, and be ever with him. 1 Thess. 4:14-17; Jude 14,15; Dan. 7:9,10; 1 Cor. 6:2,3.

Now, while they were thus drawing towards the gate, behold a company of the heavenly host came out to meet them: to whom it was said by the other two shining ones, These are the men that have loved our Lord when they were in the world, and that have left all for his holy name; and he hath sent us to fetch them, and we have brought them thus far on their desired journey, that they may go in and look their Redeemer in the face with joy. Then the heavenly host gave a great shout, saying, ``Blessed are they that are called to the marriage-supper of the Lamb." Rev. 19:9. There came out also at this time to meet them several of the King's trumpeters, clothed in white and shining raiment, who, with melodious noises and loud, made even the heavens to echo with their sound. These trumpeters saluted Christian and his fellow with ten thousand welcomes from the world; and this they did with shouting and sound of trumpet.

This done, they compassed them round on every side; some went before, some behind, and some on the right hand, and some on the left, (as it were to guard them through the upper regions,) continually sounding as they went, with melodious noise, in notes on high; so that the very sight was to them that could behold it as if heaven itself was come down to meet them. Thus, therefore, they walked on together; and, as they walked, ever and anon these trumpeters, even with joyful sound, would, by mixing their music with looks and gestures, still signify to Christian and his brother how welcome they were into their company, and with what gladness they came to meet them. And now were these two men, as it were, in heaven, before they came to it, being swallowed up with the sight of angels, and with hearing of their melodious notes. Here also they had the city itself in view; and they thought they heard all the bells therein to ring, to welcome them thereto. But, above all, the warm and joyful thoughts that they had about their own dwelling there with such company, and that for ever and ever; oh, by what tongue or pen can their glorious joy be expressed! Thus they came up to the gate.

Now when they were come up to the gate, there was written over it, in letters of gold,

``blessed are they that do his commandments, that they may have right to the tree of life, and may enter in through the gates into the city."

Then I saw in my dream, that the shining men bid them call at the gate: the which when they did, some from above looked over the gate, to wit, Enoch, Moses, and Elijah, etc., to whom it was said, These pilgrims are come from the City of Destruction, for the love that they bear to the King of this place; and then the pilgrims gave in unto them each man his certificate, which they had received in the beginning: those therefore were carried in unto the King, who, when he had read them, said, Where are the men? To whom it was answered, They are standing without the gate. The King then commanded to open the gate, ``That the righteous nation (said he) that keepeth the truth may enter in."

Isa. 26:2.

Now I saw in my dream, that these two men went in at the gate; and lo, as they entered, they were transfigured; and they had raiment put on that shone like gold. There were also that met them with harps and crowns, and gave them to them; the harps to praise withal, and the crowns in token of honor. Then I heard in my dream, that all the bells in the city rang again for joy, and that it was said unto them,

``enter ye into the joy of your lord."

I also heard the men themselves, that they sang with a loud voice, saying,

``blessing, and honor, and glory, and power, be unto him that sitteth upon the throne, and unto the lamb, for ever and ever."

Now, just as the gates were opened to let in the men, I looked in after them, and behold the city shone like the sun; the streets also were paved with gold; and in them walked many men, with crowns on their heads, palms in their hands, and golden harps, to sing praises withal.

There were also of them that had wings, and they answered one another without intermission, saying, Holy, holy, holy is the Lord. And after that they shut up the gates; which, when I had seen, I wished myself among them.

Now, while I was gazing upon all these things, I turned my head to look back, and saw Ignorance come up to the river side; but he soon got over, and that without half the difficulty which the other two men met with. For it happened that there was then in that place one Vain-Hope, a ferryman, that with his boat helped him over; so he, as the other I saw, did ascend the hill, to come up to the gate; only he came alone, neither did any man meet him with the least encouragement. When he was come up to the gate, he looked up to the writing that was above, and then began to knock, supposing that entrance should have been quickly administered to him; but he was asked by the men that looked over the top of the gate, Whence come you? and what would you have? He answered, I have ate and drank in the presence of the King, and he has taught in our streets. Then they asked him for his certificate, that they might go in and show it to the King: so he fumbled in his bosom for one, and found none. Then said they, Have you none? but the man answered never a word. So they told the King, but he would not come down to see him, but commanded the two shining ones, that conducted Christian and Hopeful to the city, to go out and take Ignorance, and bind him hand and foot, and have him away. Then they took him up, and carried him through the air to the door that I saw in the side of the hill, and put him in there. Then I saw that there was a way to hell, even from the gate of heaven, as well as from the City of Destruction. So I awoke, and behold it was a dream. 
