\chapter[THE SECOND STAGE]{}

So I saw in my dream, that they walked on their way, and had the weather very comfortable to them.

Then Christiana began to sing, saying,
\begin{verse}
Blessed be the day that I began\\
A pilgrim for to be;\\
And blessed also be the man\\
That thereto moved me.\\
\end{verse}
\begin{verse}
'Tis true, t was long ere I began\\
To seek to live for ever;\\
But now I run fast as I can:\\
Tis better late than never.\\
\end{verse}
\begin{verse}
Our tears to joy, our fears to faith,\\
Are turned, as we see;\\
Thus our beginning (as one saith)\\
Shows what our end will be.\\
\end{verse}
Now there was, on the other side of the wall that fenced in the way up which Christiana and her companions were to go, a garden, and that garden belonged to him whose was that barking dog, of whom mention was made before. And some of the fruit-trees that grew in that garden shot their branches over the wall; and being mellow, they that found them did gather them up, and eat of them to their hurt. So Christiana's boys, as boys are apt to do, being pleased with the trees, and with the fruit that hung thereon, did pluck them, and began to eat. Their mother did also chide them for so doing, But still the boys went on.

Well, said she, my sons, you transgress, for that fruit is none of ours; but she did not know that it belonged to the enemy: I'll warrant you, if she had she would have been ready to die for fear. But that passed, and they went on their way. Now, by that they were gone about two bow-shots from the place that led them into the way, they espied two very ill-favored ones coming down apace to meet them. With that, Christiana and Mercy her friend covered themselves with their veils, and so kept on their journey: the children also went on before; so that at last they met together. Then they that came down to meet them, came just up to the women, as if they would embrace them; but Christiana said, stand back, or go peaceably as you should. Yet these two, as men that are deaf, regarded not Christiana's words, but began to lay hands upon them: at that Christiana waxing very wroth, spurned at them with her feet. Mercy also, as well as she could, did what she could to shift them. Christiana again said to them, Stand back, and be gone, for we have no money to lose, being pilgrims, as you see, and such too as live upon the charity of our friends.

Ill-Favored Ones: Then said one of the two men, We make no assault upon you for money, but are come out to tell you, that if you will but grant one small request which we shall ask, we will make women of you for ever.

Christiana: Now Christiana, imagining what they should mean, made answer again, We will neither hear, nor regard, nor yield to what you shall ask. We are in haste, and cannot stay; our business is a business of life and death. So again she and her companion made a fresh essay to go past them; but they letted them in their way.

Ill-Favored Ones: And they said, We intend no hurt to your lives; it is another thing we would have.

Christiana: Aye, quoth Christiana, you would have us body and soul, for I know it is for that you are come; but we will die rather upon the spot, than to suffer ourselves to be brought into such snares as shall hazard our well-being hereafter. And with that they both shrieked out, and cried, Murder! murder! and so put themselves under those laws that are provided for the protection of women. Deut. 22:25-27. But the men still made their approach upon them, with design to prevail against them. They therefore cried out again.

Now they being, as I said, not far from the gate in at which they came, their voice was heard from whence they were, thither: wherefore some of the house came out, and knowing that it was Christiana's tongue, they made haste to her relief. But by that they were got within sight of them, the women were in a very great scuffle; the children also stood crying by. Then did he that came in for their relief call out to the ruffians, saying, What is that thing you do? Would you make my Lord's people to transgress? He also attempted to take them, but they did make their escape over the wall into the garden of the man to whom the great dog belonged; so the dog became their protector. This Reliever then came up to the women, and asked them how they did. So they answered, We thank thy Prince, pretty well, only we have been somewhat affrighted: we thank thee also for that thou camest in to our help, otherwise we had been overcome.

Reliever: So, after a few more words, this Reliever said as followeth: I marveled much, when you were entertained at the gate above, seeing ye knew that ye were but weak women, that you petitioned not the Lord for a conductor; then might you have avoided these troubles and dangers; for he would have granted you one.

Christiana: Alas! said Christiana, we were so taken with our present blessing, that dangers to come were forgotten by us. Besides, who could have thought, that so near the King's palace there could have lurked such naughty ones? Indeed, it had been well for us had we asked our Lord for one; but since our Lord knew it would be for our profit, I wonder he sent not one along with us.

Reliever: It is not always necessary to grant things not asked for, lest by so doing they become of little esteem; but when the want of a thing is felt, it then comes under, in the eyes of him that feels it, that estimate that properly is its due, and so consequently will be thereafter used. Had my Lord granted you a conductor, you would not either so have bewailed that oversight of yours, in not asking for one, as now you have occasion to do. So all things work for good, and tend to make you more wary.

Christiana: Shall we go back again to my Lord, and confess our folly, and ask one?

Reliever: Your confession of your folly I will present him with. To go back again, you need not, for in all places where you shall come, you will find no want at all; for in every one of my Lord's lodgings, which he has prepared for the reception of his pilgrims, there is sufficient to furnish them against all attempts whatsoever. But, as I said, He will be inquired of by them, to do it for them. Ezek. 36:37. And tis a poor thing that is not worth asking for. When he had thus said, he went back to his place, and the pilgrims went on their way.

Mercy: Then said Mercy, What a sudden blank is here! I made account that we had been past all danger, and that we should never see sorrow more.

Christiana: Thy innocency, my sister, said Christiana to Mercy, may excuse thee much; but as for me, my fault is so much the greater, for that I saw this danger before I came out of the doors, and yet did not provide for it when provision might have been had. I am much to be blamed.

Mercy: Then said Mercy, How knew you this before you came from home? Pray open to me this riddle.

Christiana: Why, I will tell you. Before I set foot out of doors, one night as I lay in my bed I had a dream about this; for methought I saw two men, as like these as ever any in the world could look, stand at my bed's feet, plotting how they might prevent my salvation. I will tell you their very words. They said, (it was when I was in my troubles,) What shall we do with this woman? for she cries out, waking and sleeping, for forgiveness: if she be sufferet do go on as she begins, we shall lose her as we have lost her husband. This you know might have made me take heed, and have provided when provision might have been had.

Mercy: Well, said Mercy, as by this neglect we have an occasion ministered unto us to behold our own imperfections, so our Lord has taken occasion thereby to make manifest the riches of his grace; for he, as we see, has followed us with unasked kindness, and has delivered us from their hands that were stronger than we, of his mere good pleasure.

Thus now, when they had talked away a little more time, they drew near to a house which stood in the way, which house was built for the relief of pilgrims, as you will find more fully related in the first part of these records of the Pilgrim's Progress. So they drew on towards the house, (the house of the Interpreter;) and when they came to the door, they heard a great talk in the house. Then they gave ear, and heard, as they thought, Christiana mentioned by name; for you must know that there went along, even before her, a talk of her and her children's going on pilgrimage. And this was the most pleasing to them, because they had heard that she was Christian's wife, that woman who was some time ago so unwilling to hear of going on pilgrimage. Thus, therefore, they stood still, and heard the good people within commending her who they little thought stood at the door. At last Christiana knocked, as she had done at the gate before. Now, when she had knocked, there came to the door a young damsel, and opened the door, and looked, and behold, two women were there.

The Damsel: Then said the damsel to them, With whom would you speak in this place?

Christiana: Christiana answered, We understand that this is a privileged place for those that are become pilgrims, and we now at this door are such: wherefore we pray that we may be partakers of that for which we at this time are come; for the day, as thou seest, is very far spent, and we are loth to-night to go any further.

The Damsel: Pray, what may I call your name, that I may tell it to my Lord within.

Christiana: My name is Christiana; I was the wife of that pilgrim that some years ago did travel this way, and these be his four children. This maiden also is my companion, and is going on pilgrimage too.

Innocent: Then Innocent ran in, (for that was her name,) and said to those within, Can you think who is at the door? There is Christiana and her children, and her companion, all waiting for entertainment here. Then they leaped for joy, and went and told their Master. So he came to the door and looking upon her, he said, Art thou that Christiana whom Christian the good man left behind him when he betook himself to a pilgrim's life.

Christiana: I am that woman that was so hard-hearted as to slight my husband's troubles, and that left him to go on in his journey alone, and these are his four children; but now I also am come, for I am convinced that no way is right but this.

Interpreter: Then is fulfilled that which is written of the man that said to his son, ``Go work to-day in my vineyard; and he said to his father, I will not: but afterwards repented and went." Matt. 21:29.

Christiana: Then said Christiana, So be it: Amen. God made it a true saying upon me, and grant that I may be found at the last of him in peace, without spot, and blameless.

Interpreter: But why standest thou thus at the door? Come in, thou daughter of Abraham; we were talking of thee but now, for tidings have come to us before how thou art become a pilgrim. Come, children, come in; come, maiden, come in. So he had them all into the house.

So when they were within, they were bidden to sit down and rest them; the which when they had done, those that attended upon the pilgrims in the house came into the room to see them. And one smiled, and another smiled, and they all smiled for joy that Christiana was become a pilgrim: They also looked upon the boys; they stroked them over their faces with the hand, in token of their kind reception of them: they also carried it lovingly to Mercy, and bid them all welcome into their Master's house.

After a while, because supper was not ready, the Interpreter took them into his Significant Rooms, and showed them what Christian, Christiana's husband, had seen some time before. Here, therefore, they saw the man in the cage, the man and his dream, the man that cut his way through his enemies, and the picture of the biggest of them all, together with the rest of those things that were then so profitable to Christian.

This done, and after those things had been somewhat digested by Christiana and her company, the Interpreter takes them apart again, and has them first into a room where was a man that could look no way but downwards, with a muck-rake in his hand. There stood also one over his head with a celestial crown in his hand, and proffered him that crown for his muck-rake; but the man did neither look up nor regard, but raked to himself the straws, the small sticks, and dust of the floor.

Then said Christiana, I persuade myself that I know somewhat the meaning of this; for this is a figure of a man of this world: is it not, good sir?

Interpreter: Thou hast said right, said he; and his muck-rake doth show his carnal mind. And whereas thou seest him rather give heed to rake up straws and sticks, and the dust of the floor, than to do what He says that calls to him from above with the celestial crown in his hand; it is to show, that heaven is but as a fable to some, and that things here are counted the only things substantial. Now, whereas it was also showed thee that the man could look no way but downwards, it is to let thee know that earthly things, when they are with power upon men's minds, quite carry their hearts away from God.

Christiana: Then said Christiana, O deliver me from this muck-rake. Prov. 30:8.

Interpreter: That prayer, said the Interpreter, has lain by till it is almost rusty: ``Give me not riches," is scarce the prayer of one in ten thousand. Straws, and sticks, and dust, with most, are the great things now looked after.

With that Christiana and Mercy wept, and said, It is, alas! too true.

When the Interpreter had shown them this, he had them into the very best room in the house; a very brave room it was. So he bid them look round about, and see if they could find any thing profitable there. Then they looked round and round; for there was nothing to be seen but a very great spider on the wall, and that they overlooked.

Mercy: Then said Mercy, Sir, I see nothing; but Christiana held her peace.

Interpreter: But, said the Interpreter, look again. She therefore looked again, and said, Here is not any thing but an ugly spider, who hangs by her hands upon the wall. Then said he, Is there but one spider in all this spacious room? Then the water stood in Christiana's eyes, for she was a woman quick of apprehension; and she said, Yea, Lord, there are more here than one; yea, and spiders whose venom is far more destructive than that which is in her. The Interpreter then looked pleasantly on her, and said, Thou hast said the truth. This made Mercy to blush, and the boys to cover their faces; for they all began now to understand the riddle.

Then said the Interpreter again, ``The spider taketh hold with her hands," as you see, ``and is in kings' palaces." Prov. 30:28. And wherefore is this recorded, but to show you, that, how full of the venom of sin soever you be, yet you may, by the hand of Faith, lay hold of and dwell in the best room that belongs to the King's house above?

Christiana: I thought, said Christiana, of something of this; but I could not imagine it at all. I thought that we were like spiders, and that we looked like ugly creatures, in what fine room soever we were: but that by this spider, that venomous and ill-favored creature, we were to learn how to act faith, that came not into my thoughts; and yet she had taken hold with her hands, and, as I see, dwelleth in the best room in the house. God has made nothing in vain.

Then they seemed all to be glad; but the water stood in their eyes; yet they looked one upon another, and also bowed before the Interpreter.

He had them into another room, where were a hen and chickens, and bid them observe a while. So one of the chickens went to the trough to drink, and every time she drank she lifted up her head and her eyes towards heaven. See, said he, what this little chick doth, and learn of her to acknowledge whence your mercies come, by receiving them with looking up. Yet again, said he, observe and look: so they gave heed, and perceived that the hen did walk in a fourfold method towards her chickens: 1. She had a common call, and that she hath all the day long. 2. She had a special call, and that she had but sometimes. 3. She had a brooding note. Matt. 23:37. And, 4. She had an outcry.

Now, said he, compare this hen to your King and these chickens to his obedient ones; for, answerable to her, he himself hath his methods which he walketh in towards his people. By his common call, he gives nothing; by his special call, he always has something to give; he has also a brooding voice, for them that are under his wing; and he has an outcry, to give the alarm when he seeth the enemy come. I choose, my darlings, to lead you into the room where such things are, because you are women, and they are easy for you.

Christiana: And, sir, said Christiana, pray let us see some more. So he had them into the slaughter-house, where was a butcher killing a sheep; and behold, the sheep was quiet, and took her death patiently. Then said the Interpreter, You must learn of this sheep to suffer and to put up with wrongs without murmurings and complaints. Behold how quietly she takes her death, and, without objecting, she suffereth her skin to be pulled over her ears. Your King doth call you his sheep.

After this he led them into his garden, where was great variety of flowers; and he, said, Do you see all these? So Christiana said, Yes. Then said he again, Behold, the flowers are diverse in stature, in quality, and color, and smell, and virtue; and some are better than others; also, where the gardener has set them, there they stand, and quarrel not one with another.

Again, he had them into his field, which he had sown with wheat and corn: but when they beheld, the tops of all were cut off, and only the straw remained. He said again, This ground was dunged, and ploughed, and sowed, but what shall we do with the crop? Then said Christiana, Burn some, and make muck of the rest. Then said the Interpreter again, Fruit, you see, is that thing you look for; and for want of that you condemn it to the fire, and to be trodden under foot of men: beware that in this you condemn not yourselves.

Then, as they were coming in from abroad, they espied a little robin with a great spider in his mouth. So the Interpreter said, Look here. So they looked, and Mercy wondered, but Christiana said, What a disparagement is it to such a pretty little bird as the robin-red-breast; he being also a bird above many, that loveth to maintain a kind of sociableness with men! I had thought they had lived upon crumbs of bread, or upon other such harmless matter: I like him worse than I did.

The Interpreter then replied, This robin is an emblem, very apt to set forth some professors by; for to sight they are, as this robin, pretty of note, color, and carriage. They seem also to have a very great love for professors that are sincere; and, above all others, to desire to associate with them, and to be in their company, as if they could live upon the good man's crumbs. They pretend also, that therefore it is that they frequent the house of the godly, and the appointments of the Lord: but when they are by themselves, as the robin, they can catch and gobble up spiders; they can change their diet, drink iniquity, and swallow down sin like water.

So, when they were come again into the house, because supper as yet was not ready, Christiana again desired that the Interpreter would either show or tell some other things that are profitable.

Then the Interpreter began, and said, The fatter the sow is, the more she desires the mire; the fatter the ox is, the more gamesomely he goes to the slaughter; and the more healthy the lustful man is, the more prone he is unto evil. There is a desire in women to go neat and find; and it is a comely thing to be adorned with that which in God's sight is of great price. T is easier watching a night or two, than to sit up a whole year together: so t is easier for one to begin to profess well, than to hold out as he should to the end. Every shipmaster, when in a storm, will willingly cast that overboard which is of the smallest value in the vessel; but who will throw the best out first? None but he that feareth not God. One leak will sink a ship, and one sin will destroy a sinner. He that forgets his friend is ungrateful unto him; but he that forgets his Saviour is unmerciful to himself. He that lives in sin, and looks for happiness hereafter, is like him that soweth cockle, and thinks to fill his barn with wheat or barley. If a man would live well, let him fetch his last day to him, and make it always his company-keeper. Whispering, and change of thoughts, prove that sin is in the world. If the world, which God sets light by, is counted a thing of that worth with men, what is heaven, that God commendeth? If the life that is attended with so many troubles, is so loth to be let go by us, what is the life above? Every body will cry up the goodness of men; but who is there that is, as he should be, affected with the goodness of God? We seldom sit down to meat, but we eat, and leave. So there is in Jesus Christ more merit and righteousness than the whole world has need of.

When the Interpreter had done, he takes them out into his garden again, and had them to a tree whose inside was all rotten and gone, and yet it grew and had leaves. Then said Mercy, What means this? This tree, said he, whose outside is fair, and whose inside is rotten, is that to which many may be compared that are in the garden of God; who with their mouths speak high in behalf of God, but indeed will do nothing for him; whose leaves are fair, but their heart good for nothing but to be tinder for the devil's tinder-box.

Now supper was ready, the table spread, and all things set on the board: so they sat down, and did eat, when one had given thanks. And the Interpreter did usually entertain those that lodged with him with music at meals; so the minstrels played. There was also one that did sing, and a very fine voice he had. His song was this:
\begin{verse}
``The Lord is only my support,\\
And he that doth me feed;\\
How can I then want any thing\\
Whereof I stand in need?"\\
\end{verse}

When the song and music were ended, the Interpreter asked Christiana what it was that at first did move her thus to betake herself to a pilgrim's life. Christiana answered, First, the loss of my husband came into my mind, at which I was heartily grieved; but all that was but natural affection. Then after that came the troubles and pilgrimage of my husband into my mind, and also how like a churl I had carried it to him as to that. So guilt took hold of my mind, and would have drawn me into the pond, but that opportunely I had a dream of the well-being of my husband, and a letter sent me by the King of that country where my husband dwells, to come to him. The dream and the letter together so wrought upon my mind that they forced me to this way.

Interpreter: But met you with no opposition before you set out of doors?

Christiana: Yes, a neighbor of mine, one Mrs. Timorous: she was akin to him that would have persuaded my husband to go back, for fear of the lions. She also befooled me, for, as she called it, my intended desperate adventure; she also urged what she could to dishearten me from it, the hardships and troubles that my husband met with in the way; but all this I got over pretty well. But a dream that I had of two ill-looking ones, that I thought did plot how to make me miscarry in my journey, that hath troubled me much: yea, it still runs in my mind, and makes me afraid of every one that I meet, lest they should meet me to do me a mischief, and to turn me out of my way. Yea, I may tell my Lord, though I would not have every body know of it, that between this and the gate by which we got into the way, we were both so sorely assaulted that we were made to cry out murder; and the two that made this assault upon us, were like the two that I saw in my dream.

Then said the Interpreter, Thy beginning is good; thy latter end shall greatly increase. So he addressed himself to Mercy, and said unto her, And what moved thee to come hither, sweet heart?

Mercy: Then Mercy blushed and trembled, and for a while continued silent.

Interpreter: Then said he, Be not afraid; only believe, and speak thy mind.

Mercy: So she began, and said, Truly, sir, my want of experience is that which makes me covet to be in silence, and that also that fills me with fears of coming short at last. I cannot tell of visions and dreams, as my friend Christiana can; nor know I what it is to mourn for my refusing the counsel of those that were good relations.

Interpreter: What was it, then, dear heart, that hath prevailed with thee to do as thou hast done?

Mercy: Why, when our friend here was packing up to be gone from our town, I and another went accidentally to see her. So we knocked at the door and went in. When we were within, and seeing what she was doing, we asked her what was her meaning. She said she was sent for to go to her husband; and then she up and told us how she had seen him in a dream, dwelling in a curious place, among immortals, wearing a crown, playing upon a harp, eating and drinking at his Prince's table, and singing praises to him for bringing him thither, etc. Now, methought, while she was telling these things unto us, my heart burned within me. And I said in my heart, If this be true, I will leave my father and my mother, and the land of my nativity, and will, if I may, go along with Christiana. So I asked her further of the truth of these things, and if she would let me go with her; for I saw now that there was no dwelling, but with the danger of ruin, any longer in our town. But yet I came away with a heavy heart; not for that I was unwilling to come away, but for that so many of my relations were left behind. And I am come with all the desire of my heart, and will go, if I may, with Christiana unto her husband and his King.

Interpreter: Thy setting out is good, for thou hast given credit to the truth; thou art a Ruth, who did, for the love she bare to Naomi and to the Lord her God, leave father and mother, and the land of her nativity, to come out and go with a people she knew not heretofore. ``The Lord recompense thy work, and a full reward be given thee of the Lord God of Israel, under whose wings thou art come to trust." Ruth 2:11,12.

Now supper was ended, and preparation was made for bed; the women were laid singly alone, and the boys by themselves. Now when Mercy was in bed, she could not sleep for joy, for that now her doubts of missing at last were removed further from her than ever they were before. So she lay blessing and praising God, who had such favor for her.

In the morning they arose with the sun, and prepared themselves for their departure; but the Interpreter would have them tarry a while; For, said he, you must orderly go from hence. Then said he to the damsel that first opened unto them, Take them and have them into the garden to the bath, and there wash them and make them clean from the soil which they had gathered by traveling. Then Innocent the damsel took them and led them into the garden, and brought them to the bath; so she told them that there they must wash and be clean, for so her Master would have the women to do that called at his house as they were going on pilgrimage. Then they went in and washed, yea, they and the boys, and all; and they came out of that bath, not only sweet and clean, but also much enlivened and strengthened in their joints. So when they came in, they looked fairer a deal than when they went out to the washing.

When they were returned out of the garden from the bath, the Interpreter took them and looked upon them, and said unto them, ``Fair as the moon." Then he called for the seal wherewith they used to be sealed that were washed in his bath. So the seal was brought, and he set his mark upon them, that they might be known in the places whither they were yet to go. Now the seal was the contents and sum of the passover which the children of Israel did eat, Exod. 13: 8-10, when they came out of the land of Egypt; and the mark was set between their eyes. This seal greatly added to their beauty, for it was an ornament to their faces. It also added to their gravity, and made their countenance more like those of angels.

Then said the Interpreter again to the damsel that waited upon these women, Go into the vestry, and fetch out garments for these people. So she went and fetched out white raiment, and laid it down before him; so he commanded them to put it on: it was fine linen, white and clean. When the women were thus adorned, they seemed to be a terror one to the other; for that they could not see that glory each one had in herself, which they could see in each other. Now therefore they began to esteem each other better than themselves. For, You are fairer than I am, said one; and, You are more comely than I am, said another. The children also stood amazed, to see into what fashion they were brought. 
