\chapter[THE SEVENTH STAGE]{}

Well, the time grew on that the pilgrims must go on their way; wherefore they prepared for their journey. They sent for their friends; they conferred with them; they had some time set apart therein to commit each other to the protection of their Prince. There were again that brought them of such things as they had, that were fit for the weak and the strong, for the women and the men, and so laded them with such things as were necessary. Acts 28:10. Then they set forward on their way; and their friends accompanying them so far as was convenient, they again committed each other to the protection of their King, and parted.

They therefore that were of the pilgrims' company went on, and Mr. Great-Heart went before them. Now, the women and children being weakly, they were forced to go as they could bear; by which means Mr. Ready-to-halt and Mr. Feeble-mind, had more to sympathize with their condition.

When they were gone from the townsmen, and when their friends had bid them farewell, they quickly came to the place where Faithful was put to death. Therefore they made a stand, and thanked him that had enabled him to bear his cross so well; and the rather, because they now found that they had a benefit by such a manly suffering as his was.

They went on therefore after this a good way further, talking of Christian and Faithful, and how Hopeful joined himself to Christian after that Faithful was dead.

Now they were come up with the hill Lucre, where the silver mine was which took Demas off from his pilgrimage, and into which, as some think, By-ends fell and perished; wherefore they considered that. But when they were come to the old monument that stood over against the hill Lucre, to wit, to the pillar of salt, that stood also within view of Sodom and its stinking lake, they marvelled, as did Christian before, that men of such knowledge and ripeness of wit as they were, should be so blinded as to turn aside here. Only they considered again, that nature is not affected with the harms that others have met with, especially if that thing upon which they look has an attracting virtue upon the foolish eye.

I saw now, that they went on till they came to the river that was on this side of the Delectable Mountains; to the river where the fine trees grow on both sides, and whose leaves, if taken inwardly, are good against surfeits; where the meadows are green all the year long, and where they might lie down safely. Psa. 23:2.

By this river-side, in the meadows, there were cotes and folds for sheep, a house built for the nourishing and bringing up of those lambs, the babes of those women that go on pilgrimage. Also there was here one that was intrusted with them, who could have compassion; and that could gather these lambs with his arm, and carry them in his bosom, and gently lead those that were with young. Heb. 5:2; Isa. 40:11. Now, to the care of this man Christiana admonished her four daughters to commit their little ones, that by these waters they might be housed, harbored, succored, and nourished, and that none of them might be lacking in time to come. This man, if any of them go astray, or be lost, will bring them again; he will also bind up that which was broken, and will strengthen them that are sick. Jer. 23:4; Ezek. 34:11-16. Here they will never want meat, drink, and clothing; here they will be kept from thieves and robbers; for this man will die before one of those committed to his trust shall be lost. Besides, here they shall be sure to have good nurture and admonition, and shall be taught to walk in right paths, and that you know is a favor of no small account. Also here, as you see, are delicate waters, pleasant meadows, dainty flowers, variety of trees, and such as bear wholesome fruit: fruit, not like that which Matthew ate of, that fell over the wall out of Beelzebub's garden; but fruit that procureth health where there is none, and that continueth and increaseth it where it is. So they were content to commit their little ones to him; and that which was also an encouragement to them so to do, was, for that all this was to be at the charge of the King, and so was as an hospital to young children and orphans.

Now they went on. And when they were come to By-path Meadow, to the stile over which Christian went with his fellow Hopeful, when they were taken by Giant Despair and put into Doubting Castle, they sat down, and consulted what was best to be done: to wit, now they were so strong, and had got such a man as Mr. Great-Heart for their conductor, whether they had not best to make an attempt upon the giant, demolish his castle, and if there were any pilgrims in it, to set them at liberty before they went any further. So one said one thing, and another said the contrary. One questioned if it was lawful to go upon unconsecrated ground; another said they might, provided their end was good; but Mr. Great-Heart said, Though that assertion offered last cannot be universally true, yet I have a commandment to resist sin, to overcome evil, to fight the good fight of faith: and I pray, with whom should I fight this good fight, if not with Giant Despair? I will therefore attempt the taking away of his life, and the demolishing of Doubting Castle. Then said he, Who will go with me? Then said old Honest, I will. And so will we too, said Christiana's four sons, Matthew, Samuel, Joseph, and James; for they were young men and strong. 1 John 2:13,14. So they left the women in the road, and with them Mr. Feeble-mind, and Mr. Ready-to-halt with his crutches, to be their guard until they came back; for in that place the Giant Despair dwelt so near, they keeping in the road, a little child might lead them. Isa. 11:6.

So Mr. Great-Heart, old Honest, and the four young men, went to go up to Doubting Castle, to look for Giant Despair. When they came at the castle gate, they knocked for entrance with an unusual noise. At that the old Giant comes to the gate, and Diffidence his wife follows. Then said he, Who and what is he that is so hardy, as after this manner to molest the Giant Despair? Mr. Great-Heart replied, It is I, Great-Heart, one of the King of the Celestial country's conductors of pilgrims to their place; and I demand of thee that thou open thy gates for my entrance: prepare thyself also to fight, for I am come to take away thy head; and to demolish Doubting Castle.

Now Giant Despair, because he was a giant, thought no man could overcome him: and again thought he, Since heretofore I have made a conquest of angels, shall Great-Heart make me afraid? So he harnessed himself, and went out. He had a cap of steel upon his head, a breast-plate of fire girded to him, and he came out in iron shoes, with a great club in his hand. Then these six men made up to him, and beset him behind and before: also, when Diffidence the giantess came up to help him, old Mr. Honest cut her down at one blow. Then they fought for their lives, and Giant Despair was brought down to the ground, but was very loth die. He struggled hard, and had, as they say, as many lives as a cat; but Great-Heart was his death, for he left him not till he had severed his head from his shoulders.

Then they fell to demolishing Doubting Castle, and that you know might with ease be done, since Giant Despair was dead. They were seven days in destroying of that; and in it of pilgrims they found one Mr. Despondency, almost starved to death, and one Much-afraid, his daughter: these two they saved alive. But it would have made you wonder to have seen the dead bodies that lay here and there in the castle yard, and how full of dead men's bones the dungeon was.

When Mr. Great-Heart and his companions had performed this exploit, they took Mr. Despondency, and his daughter Much-afraid, into their protection; for they were honest people, though they were prisoners in Doubting Castle to that tyrant Giant Despair. They, therefore, I say, took with them the head of the giant, (for his body they had buried under a heap of stones,) and down to the road and to their companions they came, and showed them what they had done. Now, when Feeble-mind and Ready-to-halt saw that it was the head of Giant Despair indeed, they were very jocund and merry. Now Christiana, if need was, could play upon the viol, and her daughter Mercy upon the lute: so, since they were so merry disposed, she played them a lesson, and Ready-to-halt would dance. So he took Despondency's daughter, Much-afraid, by the hand, and to dancing they went in the road. True, he could not dance without one crutch in his hand, but I promise you he footed it well: also the girl was to be commended, for she answered the music handsomely.

As for Mr. Despondency, the music was not so much to him; he was for feeding rather than dancing, for that he was almost starved. So Christiana gave him some of her bottle of spirits for present relief, and then prepared him something to eat; and in a little time the old gentleman came to himself, and began to be finely revived.

Now I saw in my dream, when all these things were finished, Mr. Great-Heart took the head of Giant Despair, and set it upon a pole by the highway-side, right over against the pillar that Christian erected for a caution to pilgrims that came after, to take heed of entering into his grounds.

 Then he writ under it upon a marble stone these verses following:
\begin{verse}
``This is the head of him whose name only\\
In former times did pilgrims terrify.\\
His castle's down, and Diffidence his wife\\
Brave Mr. Great-Heart has bereft of life.\\
Despondency, his daughter Much-afraid,\\
Great-Heart for them also the man has play'd.\\
Who hereof doubts, if he'll but cast his eye\\
Up hither, may his scruples satisfy.\\
This head also, when doubting cripples dance,\\
Doth show from fears they have deliverance."\\
\end{verse}

When these men had thus bravely showed themselves against Doubting Castle, and had slain Giant Despair, they went forward, and went on till they came to the Delectable Mountains, where Christian and Hopeful refreshed themselves with the varieties of the place. They also acquainted themselves with the shepherds there, who welcomed them, as they had done Christian before, unto the Delectable Mountains.

Now the shepherds seeing so great a train follow Mr. Great-Heart, (for with him they were well acquainted,) they said unto him, Good sir, you have got a goodly company here; pray where did you find all these?

Then Mr. Great-Heart replied,
\begin{verse}
``First, here is Christiana and her train,\\
Her sons, and her sons' wives, who, like the wain,\\
Keep by the pole, and do by compass steer\\
From sin to grace, else they had not been here.\\
Next here's old Honest come on pilgrimage,\\
Ready-to-halt too, who I dare engage\\
True-hearted is, and so is Feeble-mind,\\
Who willing was not to be left behind.\\
Despondency, good man, is coming after,\\
And so also is Much-afraid, his daughter.\\
May we have entertainment here, or must\\
We further go? Let's knew whereon to trust."\\
\end{verse}

Then said the shepherds, This is a comfortable company. You are welcome to us; for we have for the feeble, as well as for the strong. Our Prince has an eye to what is done to the least of these; therefore Infirmity must not be a block to our entertainment. Matt. 25:40. So they had them to the palace door, and then said unto them, Come in, Mr. Feeble-Mind; come in Mr. Ready-to-halt; Come in, Mr. Despondency, and Mrs. Much-afraid his daughter. These, Mr. Great-Heart, said the shepherds to the guide, we call in by name, for that they are most subject to draw back; but as for you, and the rest that are strong, we leave you to your wonted liberty. Then said Mr. Great-Heart, This day I see that grace doth shine in your faces, and that you are my Lord's shepherds indeed; for that you have not pushed these diseased neither with side nor shoulder, but have rather strewed their way into the palace with flowers, as you should. Ezek. 34:21.

So the feeble and weak went in, and Mr. Great-Heart and the rest did follow. When they were also set down, the shepherds said to those of the weaker sort, What is it that you would have? for, said they, all things must be managed here to the supporting of the weak, as well as to the warning of the unruly. So they made them a feast of things easy of digestion, and that were pleasant to the palate and nourishing; the which when they had received, they went to their rest, each one respectively unto his proper place.

When morning was come, because the mountains were high and the day clear, and because it was the custom of the shepherds to show the pilgrims before their departure some rarities, therefore, after they were ready, and had refreshed themselves, the shepherds took them out into the fields, and showed them first what they had shown to Christian before.

Then they had them to some new places. The first was Mount Marvel, where they looked, and beheld a man at a distance that tumbled the hills about with words. Then they asked the shepherds what that should mean. So they told them, that that man was the son of one Mr. Great-grace, of whom you read in the first part of the records of the Pilgrim's Progress; and he is set there to teach pilgrims how to believe down, or to tumble out of their ways, what difficulties they should meet with, by faith. Mark 11:23,24. Then said Mr. Great-Heart, I know him; he is a man above many.

Then they had them to another place, called Mount Innocence. And there they saw a man clothed all in white; and two men, Prejudice and Ill-will, continually casting dirt upon him. Now behold, the dirt, whatsoever they cast at him, would in a little time fall off again, and his garment would look as clear as if no dirt had been cast thereat. Then said the pilgrims, What means this? The shepherds answered, This man is named Godlyman, and this garment is to show the innocency of his life. Now, those that throw dirt at him are such as hate his well-doing; but, as you see the dirt will not stick upon his clothes, so it shall be with him that liveth innocently in the world. Whoever they be that would make such men dirty, they labor all in vain; for God, by that a little time is spent, will cause that their innocence shall break forth as the light, and their righteousness as the noonday.

Then they took them, and had them to Mount Charity, where they showed them a man that had a bundle of cloth lying before him, out of which he cut coats and garments for the poor that stood about him; yet his bundle or roll of cloth was never the less. Then said they, What should this be? This is, said the shepherds, to show you, that he who has a heart to give of his labor to the poor, shall never want wherewithal. He that watereth shall be watered himself. And the cake that the widow gave to the prophet did not cause that she had the less in her barrel.

They had them also to the place where they saw one Fool and one Want-wit washing an Ethiopian, with intention to make him white; but the more they washed him, the blacker he was. Then they asked the shepherds what that should mean. So they told them, saying, Thus it is with the vile person; all means used to get such a one a good name, shall in conclusion tend but to make him more abominable. Thus it was with the pharisees; and so it shall be with all hypocrites.

Then said Mercy, the wife of Matthew, to Christiana her mother, Mother, I would, if it might be, see the hole in the hill, or that commonly called the By-way to hell. So her mother brake her mind to the shepherds. Then they went to the door; it was on the side of an hill; and they opened it, and bid Mercy hearken a while. So she hearkened, and heard one saying, Cursed be my father for holding of my feet back from the way of peace and life. Another said, Oh that I had been torn in pieces before I had, to save my life, lost my soul! And another said, If I were to live again, how would I deny myself, rather than to come to this place! Then there was as if the very earth groaned and quaked under the feet of this young woman for fear; so she looked white, and came trembling away, saying, Blessed be he and she that is delivered from this place!

Now, when the shepherds had shown them all these things, then they had them back to the palace, and entertained them with what the house would afford. But Mercy, being a young and married woman, longed for something that she saw there, but was ashamed to ask. Her mother-in-law then asked her what she ailed, for she looked as one not well. Then said Mercy, There is a looking-glass hangs up in the dining-room, off which I cannot take my mind; if, therefore, I have it not, I think I shall miscarry. Then said her mother, I will mention thy wants to the shepherds, and they will not deny thee. But she said, I am ashamed that these men should know that I longed. Nay, my daughter, said she, it is no shame, but a virtue, to long for such a thing as that. So Mercy said, Then mother, if you please, ask the shepherds if they are willing to sell it.

Now the glass was one of a thousand. It would present a man, one way, with his own features exactly; and turn it but another way, and it would show one the very face and similitude of the Prince of pilgrims himself. Yes, I have talked with them that can tell, and they have said that they have seen the very crown of thorns upon his head by looking in that glass; they have therein also seen the holes in his hands, his feet, and his side. Yea, such an excellency is there in this glass, that it will show him to one where they have a mind to see him, whether living or dead; whether in earth, or in heaven; whether in a state of humiliation, or in his exaltation; whether coming to suffer, or coming to reign. James 1:23; 1 Cor. 13:12; 2 Cor. 3:18.

Christiana therefore went to the shepherds apart, (now the names of the shepherds were Knowledge, Experience, Watchful, and Sincere,) and said unto them, There is one of my daughters, a breeding woman, that I think doth long for something that she hath seen in this house; and she thinks that she shall miscarry if she should by you be denied.

Experience: Call her, call her, she shall assuredly have what we can help her to. So they called her, and said to her, Mercy, what is that thing thou wouldst have? Then she blushed, and said, The great glass that hangs up in the dining-room. So Sincere ran and fetched it, and with a joyful consent it was given her. Then she bowed her head, and gave thanks, and said, By this I know that I have obtained favor in your eyes.

They also gave to the other young women such things as they desired, and to their husbands great commendations, for that they had joined with Mr. Great-Heart in the slaying of Giant Despair, and the demolishing of Doubting Castle.

About Christiana's neck the shepherds put a bracelet, and so did they about the necks of her four daughters; also they put ear-rings in their ears, and jewels on their foreheads.

When they were minded to go hence, they let them go in peace, but gave not to them those certain cautions which before were given to Christian and his companion. The reason was, for that these had Great-Heart to be their guide, who was one that was well acquainted with things, and so could give them their cautions more seasonably, to wit, even when the danger was nigh the approaching. What cautions Christian and his companion had received of the shepherds, they had also lost by that the time was come that they had need to put them in practice. Wherefore, here was the advantage that this company had over the other.

From thence they went on singing, and they said,
\begin{verse}
 ``Behold how fitly are the stages set\\
For their relief that pilgrims are become,\\
And how they us receive without one let,\\
That make the other life our mark and home!\\
 What novelties they have to us they give,\\
That we, though pilgrims, joyful lives may live;\\
They do upon us, too, such things bestow,\\
That show we pilgrims are, where'er we go."\\
\end{verse}
