\chapter[THE FIRST STAGE]{} 
As I walked through the wilderness of this world, I lighted on a certain place where was a den,\footnote{Bedford jail, in which the author was imprisoned for conscience’ sake} and laid me down in that place to sleep; and as I slept, I dreamed a dream. I dreamed, and behold, I saw a man clothed with rags, standing in a certain place, with his face from his own house, a book in his hand, and a great burden upon his back.Isa 64:6; Luke 14:33; Psalm 38:4. I looked and saw him open the book, and read therein; and as he read, he wept and trembled; and not being able longer to contain, he brake out with a lamentable cry, saying, ``What shall I do?"Acts 2:37; 16:30; Habak 1:2,3. 

In this plight, therefore, he went home, and restrained himself as long as he could, that his wife and children should not perceive his distress; but he could not be silent long, because that his trouble increased. Wherefore at length he brake his mind to his wife and children; and thus he began to talk to them: ``O, my dear wife," said he, ``and you the children of my bowels, I, your dear friend, am in myself undone by reason of a burden that lieth hard upon me; moreover, I am certainly informed that this our city will be burnt with fire from heaven; in which fearful overthrow, both myself, with thee my wife, and you my sweet babes, shall miserably come to ruin, except (the which yet I see not) some way of escape can be found whereby we may be delivered." At this his relations were sore amazed; not for that they believed that what he had said to them was true, but because they thought that some frenzy distemper had got into his head; therefore, it drawing towards night, and they hoping that sleep might settle his brains, with all haste they got him to bed. But the night was as troublesome to him as the day; wherefore, instead of sleeping, he spent it in sighs and tears. So when the morning was come, they would know how he did. He told them, ``Worse and worse:" he also set to talking to them again; but they began to be hardened. They also thought to drive away his distemper by harsh and surly carriage to him; sometimes they would deride, sometimes they would chide, and sometimes they would quite neglect him. Wherefore he began to retire himself to his chamber to pray for and pity them, and also to condole his own misery; he would also walk solitarily in the fields, sometimes reading, and sometimes praying: and thus for some days he spent his time. 

Now I saw, upon a time, when he was walking in the fields, that he was (as he was wont) reading in his book, and greatly distressed in his mind; and as he read, he burst out, as he had done before, crying, ``What shall I do to be saved?" Acts 16:30,31. 

I saw also that he looked this way, and that way, as if he would run; yet he stood still because (as I perceived) he could not tell which way to go. I looked then, and saw a man named Evangelist coming to him, and he asked, ``Wherefore dost thou cry?" 

He answered, ``Sir, I perceive, by the book in my hand, that I am condemned to die, and after that to come to judgment, Heb. 9:27; and I find that I am not willing to do the first, Job 10:21,22, nor able to do the second." Ezek. 22:14. 

Then said Evangelist, ``Why not willing to die, since this life is attended with so many evils?" The man answered, ``Because, I fear that this burden that is upon my back will sink me lower than the grave, and I shall fall into Tophet. Isa. 30:33. And Sir, if I be not fit to go to prison, I am not fit to go to judgment, and from thence to execution; and the thoughts of these things make me cry." 

Then said Evangelist, ``If this be thy condition, why standest thou still?" He answered, ``Because I know not whither to go." Then he gave him a parchment roll, and there was written within, ``Fly from the wrath to come." Matt. 3:7. 

The man therefore read it, and looking upon Evangelist very carefully, said, ``Whither must I fly?" Then said Evangelist, (pointing with his finger over a very wide field,) ``Do you see yonder wicket-gate?" Matt. 7:13,14. The man said, ``No." Then said the other, ``Do you see yonder shining light?" Psalm 119:105; 2 Pet. 1:19. He said, ``I think I do." Then said Evangelist, ``Keep that light in your eye, and go up directly thereto, so shalt thou see the gate; at which, when thou knockest, it shall be told thee what thou shalt do." So I saw in my dream that the man began to run. Now he had not run far from his own door when his wife and children, perceiving it, began to cry after him to return; but the man put his fingers in his ears, and ran on crying, Life! life! eternal life! Luke 14:26. So he looked not behind him, Gen. 19:17, but fled towards the middle of the plain. 

The neighbors also came out to see him run, Jer. 20:10; and as he ran, some mocked, others threatened, and some cried after him to return; and among those that did so, there were two that were resolved to fetch him back by force. The name of the one was Obstinate and the name of the other Pliable. Now by this time the man was got a good distance from them; but, however, they were resolved to pursue him, which they did, and in a little time they overtook him. Then said the man, ``Neighbors, wherefore are you come?" They said, ``To persuade you to go back with us." But he said, ``That can by no means be: you dwell," said he, ``in the city of Destruction, the place also where I was born: I see it to be so; and dying there, sooner or later, you will sink lower than the grave, into a place that burns with fire and brimstone: be content, good neighbors, and go along with me." 

Obstinate: What, said Obstinate, and leave our friends and our comforts behind us! 

Christian: Yes, said Christian, (for that was his name,) because that all which you forsake is not worthy to be compared with a little of that I am seeking to enjoy, 2 Cor. 4:18; and if you will go along with me, and hold it, you shall fare as I myself; for there, where I go, is enough and to spare. Luke 15:17. Come away, and prove my words. 

Obstinate: What are the things you seek, since you leave all the world to find them? 

Christian: I seek an inheritance incorruptible, undefiled, and that fadeth not away, 1 Peter 1:4; and it is laid up in heaven, and safe there, Heb. 11:16, to be bestowed, at the time appointed, on them that diligently seek it. Read it so, if you will, in my book. 

Obstinate: Tush, said Obstinate, away with your book; will you go back with us or no? 

Christian: No, not I, said the other, because I have laid my hand to the plough. Luke 9:62. 

Obstinate: Come then, neighbor Pliable, let us turn again, and go home without him: there is a company of these crazy-headed coxcombs, that when they take a fancy by the end, are wiser in their own eyes than seven men that can render a reason. 

Pliable: Then said Pliable, Don't revile; if what the good Christian says is true, the things he looks after are better than ours: my heart inclines to go with my neighbor. 

Obstinate: What, more fools still! Be ruled by me, and go back; who knows whither such a brain-sick fellow will lead you? Go back, go back, and be wise. 

Christian: Nay, but do thou come with thy neighbor Pliable; there are such things to be had which I spoke of, and many more glories besides. If you believe not me, read here in this book, and for the truth of what is expressed therein, behold, all is confirmed by the blood of Him that made it. Heb. 9: 17-21. 

Pliable: Well, neighbor Obstinate, said Pliable, I begin to come to a point; I intend to go along with this good man, and to cast in my lot with him: but, my good companion, do you know the way to this desired place? 

Christian: I am directed by a man whose name is Evangelist, to speed me to a little gate that is before us, where we shall receive instructions about the way. 

Pliable: Come then, good neighbor, let us be going. Then they went both together. 

Obstinate: And I will go back to my place, said Obstinate. I will be no companion of such misled, fantastical fellows. 

Now I saw in my dream, that when Obstinate was gone back, Christian and Pliable went talking over the plain; and thus they began their discourse. 

Christian: Come, neighbor Pliable, how do you do? I am glad you are persuaded to go along with me. Had even Obstinate himself but felt what I have felt of the powers and terrors of what is yet unseen, he would not thus lightly have given us the back. 

Pliable: Come, neighbor Christian, since there are none but us two here, tell me now farther, what the things are, and how to be enjoyed, whither we are going. 

Christian: I can better conceive of them with my mind, than speak of them with my tongue: but yet, since you are desirous to know, I will read of them in my book. 

Pliable: And do you think that the words of your book are certainly true? 

Christian: Yes, verily; for it was made by Him that cannot lie. Tit. 1:2. 

Pliable: Well said; what things are they? 

Christian: There is an endless kingdom to be inhabited, and everlasting life to be given us, that we may inhabit that kingdom for ever. Isa. 65:17; John 10: 27-29. 

Pliable: Well said; and what else? 

Christian: There are crowns of glory to be given us; and garments that will make us shine like the sun in the firmament of heaven. 2 Tim. 4:8; Rev. 22:5; Matt. 13:43. 

Pliable: This is very pleasant; and what else? 

Christian: There shall be no more crying, nor sorrow; for he that is owner of the place will wipe all tears from our eyes. Isa. 25:8; Rev 7:16, 17; 21:4. 

Pliable: And what company shall we have there? 

Christian: There we shall be with seraphims and cherubims, Isaiah 6:2; 1 Thess. 4:16,17; Rev. 5:11; creatures that will dazzle your eyes to look on them. There also you shall meet with thousands and ten thousands that have gone before us to that place; none of them are hurtful, but loving and holy; every one walking in the sight of God, and standing in his presence with acceptance for ever. In a word, there we shall see the elders with their golden crowns, Rev. 4:4; there we shall see the holy virgins with their golden harps, Rev. 14:1-5; there we shall see men, that by the world were cut in pieces, burnt in flames, eaten of beasts, drowned in the seas, for the love they bare to the Lord of the place, John 12:25; all well, and clothed with immortality as with a garment. 2 Cor. 5:2. 

Pliable: The hearing of this is enough to ravish one's heart. But are these things to be enjoyed? How shall we get to be sharers thereof? 

Christian: The Lord, the governor of the country, hath recorded that in this book, Isaiah 55:1,2; John 6:37; 7:37; Rev. 21:6; 22:17; the substance of which is, if we be truly willing to have it, he will bestow it upon us freely. 

Pliable: Well, my good companion, glad am I to hear of these things: come on, let us mend our pace. 

Christian: I cannot go as fast as I would, by reason of this burden that is on my back. 

Now I saw in my dream, that just as they had ended this talk, they drew nigh to a very miry slough that was in the midst of the plain: and they being heedless, did both fall suddenly into the bog. The name of the slough was Despond. Here, therefore, they wallowed for a time, being grievously bedaubed with the dirt; and Christian, because of the burden that was on his back, began to sink in the mire. 

Pliable: Then said Pliable, Ah, neighbor Christian, where are you now? 

Christian: Truly, said Christian, I do not know. 

Pliable: At this Pliable began to be offended, and angrily said to his fellow, Is this the happiness you have told me all this while of? If we have such ill speed at our first setting out, what may we expect between this and our journey's end? May I get out again with my life, you shall possess the brave country alone for me. And with that he gave a desperate struggle or two, and got out of the mire on that side of the slough which was next to his own house: so away he went, and Christian saw him no more. 

Wherefore Christian was left to tumble in the Slough of Despond alone; but still he endeavored to struggle to that side of the slough that was farthest from his own house, and next to the wicket-gate; the which he did, but could not get out because of the burden that was upon his back: but I beheld in my dream, that a man came to him, whose name was Help, and asked him what he did there. 

Christian: Sir, said Christian, I was bid to go this way by a man called Evangelist, who directed me also to yonder gate, that I might escape the wrath to come. And as I was going thither, I fell in here. 

Help: But why did not you look for the steps? 

Christian: Fear followed me so hard that I fled the next way, and fell in. 

Help: Then, said he, Give me thine hand: so he gave him his hand, and he drew him out, Psalm 40:2, and he set him upon sound ground, and bid him go on his way. 

Then I stepped to him that plucked him out, and said, ``Sir, wherefore, since over this place is the way from the city of Destruction to yonder gate, is it, that this plat is not mended, that poor travellers might go thither with more security?" And he said unto me, ``This miry slough is such a place as cannot be mended: it is the descent whither the scum and filth that attends conviction for sin doth continually run, and therefore it is called the Slough of Despond; for still, as the sinner is awakened about his lost condition, there arise in his soul many fears and doubts, and discouraging apprehensions, which all of them get together, and settle in this place: and this is the reason of the badness of this ground. 

``It is not the pleasure of the King that this place should remain so bad. Isa. 35:3,4. His laborers also have, by the direction of his Majesty's surveyors, been for above this sixteen hundred years employed about this patch of ground, if perhaps it might have been mended: yea, and to my knowledge," said he, ``there have been swallowed up at least twenty thousand cart loads, yea, millions of wholesome instructions, that have at all seasons been brought from all places of the King's dominions, (and they that can tell, say, they are the best materials to make good ground of the place,) if so be it might have been mended; but it is the Slough of Despond still, and so will be when they have done what they can. 

``True, there are, by the direction of the Lawgiver, certain good and substantial steps, placed even through the very midst of this slough; but at such time as this place doth much spew out its filth, as it doth against change of weather, these steps are hardly seen; or if they be, men, through the dizziness of their heads, step beside, and then they are bemired to purpose, notwithstanding the steps be there: but the ground is good when they are once got in at the gate." 1 Sam. 12:23. 

Now I saw in my dream, that by this time Pliable was got home to his house. So his neighbors came to visit him; and some of them called him wise man for coming back, and some called him fool for hazarding himself with Christian: others again did mock at his cowardliness, saying, ``Surely, since you began to venture, I would not have been so base as to have given out for a few difficulties:" so Pliable sat sneaking among them. But at last he got more confidence, and then they all turned their tales, and began to deride poor Christian behind his back. And thus much concerning Pliable. 

Now as Christian was walking solitary by himself, he espied one afar off come crossing over the field to meet him; and their hap was to meet just as they were crossing the way of each other. The gentleman's name that met him was Mr. Worldly Wiseman: he dwelt in the town of Carnal Policy, a very great town, and also hard by from whence Christian came. This man then, meeting with Christian, and having some inkling\footnote{Slight knowledge.} of him, (for Christian's setting forth from the city of Destruction was much noised abroad, not only in the town where he dwelt, but also it began to be the town-talk in some other places)--Mr. Worldly Wiseman, therefore, having some guess of him, by beholding his laborious going, by observing his sighs and groans, and the like, began thus to enter into some talk with Christian. 

Mr. Worldly Wiseman: How now, good fellow, whither away after this burdened manner? 

Christian: A burdened manner indeed, as ever I think poor creature had! And whereas you ask me, Whither away? I tell you, sir, I am going to yonder wicket-gate before me; for there, as I am informed, I shall be put into a way to be rid of my heavy burden. 

Mr. Worldly Wiseman: Hast thou a wife and children? 

Christian: Yes; but I am so laden with this burden, that I cannot take that pleasure in them as formerly: methinks I am as if I had none. 1 Cor. 7:29. 

Mr. Worldly Wiseman: Wilt thou hearken to me, if I give thee counsel? 

Christian: If it be good, I will; for I stand in need of good counsel. 

Mr. Worldly Wiseman: I would advise thee, then, that thou with all speed get thyself rid of thy burden; for thou wilt never be settled in thy mind till then: nor canst thou enjoy the benefits of the blessings which God hath bestowed upon thee till then. 

Christian: That is that which I seek for, even to be rid of this heavy burden: but get it off myself I cannot, nor is there any man in our country that can take it off my shoulders; therefore am I going this way, as I told you, that I may be rid of my burden. 

Mr. Worldly Wiseman: Who bid thee go this way to be rid of thy burden? 

Christian: A man that appeared to me to be a very great and honorable person: his name, as I remember, is Evangelist. 

Mr. Worldly Wiseman: I beshrew\footnote{Wish a curse to.} him for his counsel! there is not a more dangerous and troublesome way in the world than is that into which he hath directed thee; and that thou shalt find, if thou wilt be ruled by his counsel. Thou hast met with something, as I perceive, already; for I see the dirt of the Slough of Despond is upon thee: but that slough is the beginning of the sorrows that do attend those that go on in that way. Hear me; I am older than thou: thou art like to meet with, in the way which thou goest, wearisomeness, painfulness, hunger, perils, nakedness, sword, lions, dragons, darkness, and, in a word, death, and what not. These things are certainly true, having been confirmed by many testimonies. And should a man so carelessly cast away himself, by giving heed to a stranger? 

Christian: Why, sir, this burden on my back is more terrible to me than are all these things which you have mentioned: nay, methinks I care not what I meet with in the way, if so be I can also meet with deliverance from my burden. 

Mr. Worldly Wiseman: How camest thou by thy burden at first? 

Christian: By reading this book in my hand. 

Mr. Worldly Wiseman: I thought so; and it has happened unto thee as to other weak men, who, meddling with things too high for them, do suddenly fall into thy distractions; which distractions do not only unman men, as thine I perceive have done thee, but they run them upon desperate ventures, to obtain they know not what. 

Christian: I know what I would obtain; it is ease from my heavy burden. 

Mr. Worldly Wiseman: But why wilt thou seek for ease this way, seeing so many dangers attend it? especially since (hadst thou but patience to hear me) I could direct thee to the obtaining of what thou desirest, without the dangers that thou in this way wilt run thyself into. Yea, and the remedy is at hand. Besides, I will add, that instead of those dangers, thou shalt meet with much safety, friendship, and content. 

Christian: Sir, I pray open this secret to me. 

Mr. Worldly Wiseman: Why, in yonder village (the village is named Morality) there dwells a gentleman whose name is Legality, a very judicious man, and a man of a very good name, that has skill to help men off with such burdens as thine is from their shoulders; yea to my knowledge, he hath done a great deal of good this way; aye, and besides, he hath skill to cure those that are somewhat crazed in their wits with their burdens. To him, as I said, thou mayest go, and be helped presently. His house is not quite a mile from this place; and if he should not be at home himself, he hath a pretty young man to his son, whose name is Civility, that can do it (to speak on) as well as the old gentleman himself: there, I say, thou mayest be eased of thy burden; and if thou art not minded to go back to thy former habitation, (as indeed I would not wish thee,) thou mayest send for thy wife and children to this village, where there are houses now standing empty, one of which thou mayest have at a reasonable rate: provision is there also cheap and good; and that which will make thy life the more happy is, to be sure there thou shalt live by honest neighbors, in credit and good fashion. 

Now was Christian somewhat at a stand; but presently he concluded, If this be true which this gentleman hath said, my wisest course is to take his advice: and with that he thus farther spake. 

Christian: Sir, which is my way to this honest man's house? 

Mr. Worldly Wiseman: Do you see yonder high hill? 

Christian: Yes, very well. 

Mr. Worldly Wiseman: By that hill you must go, and the first house you come at is his. 

So Christian turned out of his way to go to Mr. Legality's house for help: but, behold, when he was got now hard by the hill, it seemed so high, and also that side of it that was next the way-side did hang so much over, that Christian was afraid to venture further, lest the hill should fall on his head; wherefore there he stood still, and wotted not what to do. Also his burden now seemed heavier to him than while he was in his way. There came also flashes of fire, Ex. 19:16, 18, out of the hill, that made Christian afraid that he should be burnt: here therefore he did sweat and quake for fear. Heb. 12:21. And now he began to be sorry that he had taken Mr. Worldly Wiseman's counsel; and with that he saw Evangelist coming to meet him, at the sight also of whom he began to blush for shame. So Evangelist drew nearer and nearer; and coming up to him, he looked upon him, with a severe and dreadful countenance, and thus began to reason with Christian. 

Evangelist: What doest thou here, Christian? said he: at which words Christian knew not what to answer; wherefore at present he stood speechless before him. Then said Evangelist farther, Art not thou the man that I found crying without the walls of the city of Destruction? 

Christian: Yes, dear sir, I am the man. 

Evangelist: Did not I direct thee the way to the little wicket-gate? 

Christian: Yes, dear sir, said Christian. 

Evangelist: How is it then thou art so quickly turned aside? For thou art now out of the way. 

Christian: I met with a gentleman so soon as I had got over the Slough of Despond, who persuaded me that I might, in the village before me, find a man that could take off my burden. 

Evangelist: What was he? 

Christian: He looked like a gentleman, and talked much to me, and got me at last to yield: so I came hither; but when I beheld this hill, and how it hangs over the way, I suddenly made a stand, lest it should fall on my head. 

Evangelist: What said that gentleman to you? 

Christian: Why, he asked me whither I was going; and I told him. 

Evangelist: And what said he then? 

Christian: He asked me if I had a family; and I told him. But, said I, I am so laden with the burden that is on my back, that I cannot take pleasure in them as formerly. 

Evangelist: And what said he then? 

Christian: He bid me with speed get rid of my burden; and I told him it was ease that I sought. And, said I, I am therefore going to yonder gate, to receive farther direction how I may get to the place of deliverance. So he said that he would show me a better way, and short, not so attended with difficulties as the way, sir, that you set me in; which way, said he, will direct you to a gentleman's house that hath skill to take off these burdens: so I believed him, and turned out of that way into this, if haply I might be soon eased of my burden. But when I came to this place, and beheld things as they are, I stopped, for fear (as I said) of danger: but I now know not what to do. 

Evangelist: Then said Evangelist, Stand still a little, that I show thee the words of God. So he stood trembling. Then said Evangelist, ``See that ye refuse not Him that speaketh; for if they escaped not who refused him that spake on earth, much more shall not we escape, if we turn away from Him that speaketh from heaven." Heb. 12:25. He said, moreover, ``Now the just shall live by faith; but if any man draw back, my soul shall have no pleasure in him." Heb. 10:38. He also did thus apply them: Thou art the man that art running into this misery; thou hast begun to reject the counsel of the Most High, and to draw back thy foot from the way of peace, even almost to the hazarding of thy perdition. 

Then Christian fell down at his feet as dead, crying, Woe is me, for I am undone! At the sight of which Evangelist caught him by the right hand, saying, ``All manner of sin and blasphemies shall be forgiven unto men." Matt. 12:31. ``Be not faithless, but believing." John 20:27. Then did Christian again a little revive, and stood up trembling, as at first, before Evangelist. 

Then Evangelist proceeded, saying, Give more earnest heed to the things that I shall tell thee of. I will now show thee who it was that deluded thee, and who it was also to whom he sent thee. The man that met thee is one Worldly Wiseman, and rightly is he so called; partly because he savoreth only the doctrine of this world, 1 John 4:5, (therefore he always goes to the town of Morality to church;) and partly because he loveth that doctrine best, for it saveth him best from the cross, Gal. 6:12: and because he is of this carnal temper, therefore he seeketh to pervert my ways, though right. Now there are three things in this man's counsel that thou must utterly abhor. 

1. His turning thee out of the way. 

2. His laboring to render the cross odious to thee. 

3. And his setting thy feet in that way that leadeth unto the administration of death. 

First, Thou must abhor his turning thee out of the way; yea, and thine own consenting thereto; because this is to reject the counsel of God for the sake of the counsel of a Worldly Wiseman. The Lord says, ``Strive to enter in at the straight gate," Luke 13:24, the gate to which I send thee; ``for strait is the gate that leadeth unto life, and few there be that find it." Matt. 7:13,14. From this little wicket-gate, and from the way thereto, hath this wicked man turned thee, to the bringing of thee almost to destruction: hate, therefore, his turning thee out of the way, and abhor thyself for hearkening to him. 

Secondly, Thou must abhor his laboring to render the cross odious unto thee; for thou art to prefer it before the treasures of Egypt. Heb. 11:25,26. Besides, the King of glory hath told thee, that he that will save his life shall lose it. And he that comes after him, and hates not his father, and mother, and wife, and children, and brethren, and sisters, yea, and his own life also, he cannot be his disciple. Mark 8:38; John 12:25; Matt. 10:39; Luke 14:26. I say, therefore, for a man to labor to persuade thee that that shall be thy death, without which, the truth hath said, thou canst not have eternal life, this doctrine thou must abhor. 

Thirdly, Thou must hate his setting of thy feet in the way that leadeth to the ministration of death. And for this thou must consider to whom he sent thee, and also how unable that person was to deliver thee from thy burden. 

He to whom thou wast sent for ease, being by name Legality, is the son of the bond-woman which now is, and is in bondage with her children, Gal. 4:21-27, and is, in a mystery, this Mount Sinai, which thou hast feared will fall on thy head. Now if she with her children are in bondage, how canst thou expect by them to be made free? This Legality, therefore, is not able to set thee free from thy burden. No man was as yet ever rid of his burden by him; no, nor ever is like to be: ye cannot be justified by the works of the law; for by the deeds of the law no man living can be rid of his burden: Therefore Mr. Worldly Wiseman is an alien, and Mr. Legality is a cheat; and for his son Civility, notwithstanding his simpering looks, he is but a hypocrite, and cannot help thee. Believe me, there is nothing in all this noise that thou hast heard of these sottish men, but a design to beguile thee of thy salvation, by turning thee from the way in which I had set thee. After this, Evangelist called aloud to the heavens for confirmation of what he had said; and with that there came words and fire out of the mountain under which poor Christian stood, which made the hair of his flesh stand up. The words were pronounced: ``As many as are of the works of the law, are under the curse; for it is written, Cursed is every one that continueth not in all things which are written in the book of the law to do them." Gal. 3:10. 

Now Christian looked for nothing but death, and began to cry out lamentably; even cursing the time in which he met with Mr. Worldly Wiseman; still calling himself a thousand fools for hearkening to his counsel. He also was greatly ashamed to think that this gentleman's arguments, flowing only from the flesh, should have the prevalency with him so far as to cause him to forsake the right way. This done, he applied himself again to Evangelist in words and sense as follows. 

Christian: Sir, what think you? Is there any hope? May I now go back, and go up to the wicket-gate? Shall I not be abandoned for this, and sent back from thence ashamed? I am sorry I have hearkened to this man's counsel; but may my sin be forgiven? 

Evangelist: Then said Evangelist to him, Thy sin is very great, for by it thou hast committed two evils: thou hast forsaken the way that is good, to tread in forbidden paths. Yet will the man at the gate receive thee, for he has good-will for men; only, said he, take heed that thou turn not aside again, lest thou ``perish from the way, when his wrath is kindled but a little." Psalm 2:12.