__________________________________________________________________

CONTENTS

[1]Author's Apology for his Book

[2]PART I.

[3]The First Stage. --Christian's deplorable condition--Evangelist
directs him--Obstinate and Pliable--Slough of Despond--Worldly
Wiseman--Mount Sinai--Conversation with Evangelist

[4]The Second Stage. --The Gate--conversation with Good-Will--the
Interpreter's House--Christian entertained--the sights there shown him

[5]The Third Stage. --Loses his burden at the Cross--Simple, Sloth,
Presumption, Formalist, Hypocrisy--hill Difficulty--the Arbor--misses
his roll--the palace Beautiful--the lions--talk with Discretion, Piety,
Prudence, and Charity--wonders shown to Christian--he is armed

[6]The Fourth Stage. --Valley of Humiliation--conflict with
Apollyon--Valley of the Shadow of Death--Giants Pope and Pagan

[7]The Fifth Stage. --Discourse with Faithful--Talkative and
Faithful--Talkative's character

[8]The Sixth Stage. --Evangelist overtakes Christian and
Faithful--Vanity Fair--the Pilgrims brought to trial--Faithful's
martyrdom

[9]The Seventh Stage. --Christian and Hopeful--By-ends and his
companions--plain of Ease--Lucre-hill--Demas--the River of
Life--Vain-Confidence--Giant Despair--the Pilgrims beaten--the
Dungeon--the Key of Promise

[10]The Eighth Stage. --The Delectable Mountains--entertained by the
Shepherds--a by-way to Hell

[11]The Ninth Stage. --Christian and Hopeful meet
Ignorance--Turn-away--Little-Faith--the Flatterer--the net--chastised
by a Shining One--Atheist--Enchanted Ground--Hopeful's account of his
conversion--discourse of Christian and Ignorance

[12]The Tenth Stage. --Talk of Christian and Hopeful--Temporary--the
backslider--the land of Beulah--Christian and Hopeful pass the
River--welcome to the Celestial city

[13]Conclusion of Part First

[14]PART II.

[15]The Author's Way

[16]To the Reader

[17]The First Stage. --Christiana and Mercy--Slough of
Despond--knocking at the gate--the Dog--talk between the Pilgrims

[18]The Second Stage. --The Devil's garden--two ill-favored ones
assault them--the Reliever--entertainment at the Interpreter's
house--the Significant Rooms--Christiana and Mercy's experience

[19]The Third Stage. --Accompanied by Great-Heart--the Cross--justified
by Christ--Sloth and his companions hung--the hill Difficulty--the
Arbor

[20]The Fourth Stage. --The Lions--Giant Grim slain by Great-Heart--the
Pilgrims entertained--the children catechized by Prudence--Mr.
Brisk--Matthew sick--the remedy--sights shown the Pilgrims

[21]The Fifth Stage. --Valley of Humiliation--Valley of the Shadow of
Death--Giant Maul slain

[22]The Sixth Stage. --Discourse with Old Honest--character and history
of Mr. Fearing--Mr. Self-will and some professors--Gaius'
house--conversation--the supper--Old Honest and Great-Heart's riddles
and discourse--Giant Slay-good killed--Mr. Feeble-mind's history--Mr.
Ready-to-halt--Vanity Fair--Mr. Mnason's house--cheering entertainment
and converse--a Monster

[23]The Seventh Stage. --Hill Lucre--River of Life--Giant Despair
killed--the Delectable Mountains--entertainment by the Shepherds

[24]The Eighth Stage. --Valiant-for-Truth's-Victory--his talk with
Great-Heart--the Enchanted Ground--Heedless and Too-bold--Mr.
Stand-fast--Madam Bubble's temptations--the land of Beulah--Christiana
summoned--her parting addresses--she passes the River--she is followed
by Ready-to-halt, Feeble-mind, Despondency and his daughter, Honest,
Valiant, Steadfast
