__________________________________________________________________

THE NINTH STAGE

And I slept, and dreamed again, and saw the same two pilgrims going
down the mountains along the highway towards the city. Now, a little
below these mountains, on the left hand, lieth the country of Conceit,
from which country there comes into the way in which the pilgrims
walked, a little crooked lane. Here, therefore, they met with a very
brisk lad that came out of that country, and his name was Ignorance. So
Christian asked him from what parts he came, and whither he was going.

Ignorance: Sir, I was born in the country that lieth off there, a
little on the left hand, and I am going to the Celestial City.

Christian: But how do you think to get in at the gate, for you may find
some difficulty there?

Ignorance: As other good people do, said he.

Christian: But what have you to show at that gate, that the gate should
be opened to you?

Ignorance: I know my Lord's will, and have been a good liver; I pay
every man his own; I pray, fast, pay tithes, and give alms, and have
left my country for whither I am going.

Christian: But thou camest not in at the wicket-gate, that is at the
head of this way; thou camest in hither through that same crooked lane,
and therefore I fear, however thou mayest think of thyself, when the
reckoning-day shall come, thou wilt have laid to thy charge, that thou
art a thief and a robber, instead of getting admittance into the city.

Ignorance: Gentlemen, ye be utter strangers to me; I know you not: be
content to follow the religion of your country, and I will follow the
religion of mine. I hope all will be well. And as for the gate that you
talk of, all the world knows that is a great way off of our country. I
cannot think that any man in all our parts doth so much as know the way
to it; nor need they matter whether they do or no, since we have, as
you see, a fine, pleasant, green lane, that comes down from our
country, the next way into the way.

When Christian saw that the man was wise in his own conceit, he said to
Hopeful whisperingly, "There is more hope of a fool than of him." Prov.
26:12. And said, moreover, "When he that is a fool walketh by the way,
his wisdom faileth him, and he saith to every one that he is a fool.
Eccles. 10:3. What, shall we talk farther with him, or outgo him at
present, and so leave him to think of what he hath heard already, and
then stop again for him afterwards, and see if by degrees we can do any
good to him? Then said Hopeful,


"Let Ignorance a little while now muse

On what is said, and let him not refuse

Good counsel to embrace, lest he remain

Still ignorant of what's the chiefest gain.

God saith, those that no understanding have,

(Although he made them,) them he will not save."

Hopeful: He further added, It is not good, I think, to say so to him
all at once; let us pass him by, if you will, and talk to him anon,
even as he is able to bear it.

So they both went on, and Ignorance he came after. Now, when they had
passed him a little way, they entered into a very dark lane, where they
met a man whom seven devils had bound with seven strong cords, and were
carrying him back to the door that they saw on the side of the hill.
Matt. 12:45; Prov. 5:22. Now good Christian began to tremble, and so
did Hopeful, his companion; yet, as the devils led away the man,
Christian looked to see if he knew him; and he thought it might be one
Turn-away, that dwelt in the town of Apostacy. But he did not perfectly
see his face, for he did hang his head like a thief that is found; but
being gone past, Hopeful looked after him, and espied on his back a
paper with this inscription, "Wanton professor, and damnable apostate."

Then said Christian to his fellow, Now I call to remembrance that which
was told me of a thing that happened to a good man hereabout. The name
of the man was Little-Faith; but a good man, and he dwelt in the town
of Sincere. The thing was this. At the entering in at this passage,
there comes down from Broadway-gate, a lane, called Dead-Man's lane; so
called because of the murders that are commonly done there; and this
Little-Faith going on pilgrimage, as we do now, chanced to sit down
there and sleep. Now there happened at that time to come down the lane
from Broadway-gate, three sturdy rogues, and their names were
Faint-Heart, Mistrust, and Guilt, three brothers; and they, espying
Little-Faith where he was, came galloping up with speed. Now the good
man was just awaked from his sleep, and was getting up to go on his
journey. So they came up all to him, and with threatening language bid
him stand. At this, Little-Faith looked as white as a sheet, and had
neither power to fight nor fly. Then said Faint-Heart, Deliver thy
purse; but he making no haste to do it, (for he was loth to lose his
money,) Mistrust ran up to him, and thrusting his hand into his pocket,
pulled out thence a bag of silver. Then he cried out, Thieves, thieves!
With that, Guilt, with a great club that was in his hand, struck
Little-Faith on the head, and with that blow felled him flat to the
ground, where he lay bleeding as one that would bleed to death. All
this while the thieves stood by. But at last, they hearing that some
were upon the road, and fearing lest it should be one Great-Grace, that
dwells in the town of Good-Confidence, they betook themselves to their
heels, and left this good man to shift for himself. Now, after a while,
Little-Faith came to himself, and getting up, made shift to scramble on
his way. This was the story.

Hopeful: But did they take from him all that ever he had?

Christian: No; the place where his jewels were they never ransacked; so
those he kept still. But, as I was told, the good man was much
afflicted for his loss; for the thieves got most of his spending-money.
That which they got not, as I said, were jewels; also, he had a little
odd money left, but scarce enough to bring him to his journey's end.
Nay, (if I was not misinformed,) he was forced to beg as he went, to
keep himself alive, for his jewels he might not sell; but beg and do
what he could, he went, as we say, with many a hungry belly the most
part of the rest of the way. 1 Pet. 4:18.

Hopeful: But is it not a wonder they got not from him his certificate,
by which he was to receive his admittance at the Celestial Gate?

Christian: It is a wonder; but they got not that, though they missed it
not through any good cunning of his; for he, being dismayed by their
coming upon him, had neither power nor skill to hide any thing; so it
was more by good providence than by his endeavor that they missed of
that good thing. 2 Tim. 1:12-14; 2 Pet. 2:9.

Hopeful: But it must needs be a comfort to him they got not this jewel
from him.

Christian: It might have been great comfort to him, had he used it as
he should; but they that told me the story said that he made but little
use of it all the rest of the way, and that because of the dismay that
he had in their taking away his money. Indeed, he forgot it a great
part of the rest of his journey; and besides, when at any time it came
into his mind, and he began to be comforted therewith, then would fresh
thoughts of his loss come again upon him, and these thoughts would
swallow up all.

Hopeful: Alas, poor man, this could not but be a great grief to him.

Christian: Grief? Aye, a grief indeed! Would it not have been so to any
of us, had we been used as he, to be robbed and wounded too, and that
in a strange place, as he was? It is a wonder he did not die with
grief, poor heart. I was told that he scattered almost all the rest of
the way with nothing but doleful and bitter complaints; telling, also,
to all that overtook him, or that he overtook in the way as he went,
where he was robbed, and how; who they were that did it, and what he
had lost; how he was wounded, and that he hardly escaped with life.

Hopeful: But it is a wonder that his necessity did not put him upon
selling or pawning some of his jewels, that he might have wherewith to
relieve himself in his journey.

Christian: Thou talkest like one upon whose head is the shell to this
very day. For what should he pawn them? or to whom should he sell them?
In all that country where he was robbed, his jewels were not accounted
of; nor did he want that relief which could from thence be administered
to him. Besides, had his jewels been missing at the gate of the
Celestial City, he had (and that he knew well enough) been excluded
from an inheritance there, and that would have been worse to him than
the appearance and villany of ten thousand thieves.

Hopeful: Why art thou so tart, my brother? Esau sold his birthright,
and that for a mess of pottage, Heb. 12:16; and that birthright was his
greatest jewel: and if he, why might not Little-Faith do so too?

Christian: Esau did sell his birthright indeed, and so do many besides,
and by so doing exclude themselves from the chief blessing, as also
that caitiff did; but you must put a difference betwixt Esau and
Little-Faith, and also betwixt their estates. Esau's birthright was
typical; but Little-Faith's jewels were not so. Esau's belly was his
god; but Little-Faith's belly was not so. Esau's want lay in his fleshy
appetite; Little-Faith's did not so. Besides, Esau could see no further
than to the fulfilling of his lusts: For I am at the point to die, said
he: and what good will this birthright do me? Gen. 25:32. But
Little-Faith, though it was his lot to have but a little faith, was by
his little faith kept from such extravagances, and made to see and
prize his jewels more than to sell them, as Esau did his birthright.
You read not any where that Esau had faith, no, not so much as a
little; therefore no marvel, where the flesh only bears sway, (as it
will in that man where no faith is to resist,) if he sells his
birthright and his soul and all, and that to the devil of hell; for it
is with such as it is with the ass, who in her occasion cannot be
turned away, Jer. 2:24: when their minds are set upon their lusts, they
will have them, whatever they cost. But Little-Faith was of another
temper; his mind was on things divine; his livelihood was upon things
that were spiritual, and from above: therefore, to what end should he
that is of such a temper sell his jewels (had there been any that would
have bought them) to fill his mind with empty things? Will a man give a
penny to fill his belly with hay? or can you persuade the turtle-dove
to live upon carrion, like the crow? Though faithless ones can, for
carnal lusts, pawn, or mortgage, or sell what they have, and themselves
outright to boot; yet they that have faith, saving faith, though but a
little of it, cannot do so. Here, therefore, my brother, is thy
mistake.

Hopeful: I acknowledge it; but yet your severe reflection had almost
made me angry.

Christian: Why, I did but compare thee to some of the birds that are of
the brisker sort, who will run to and fro in untrodden paths with the
shell upon their heads: but pass by that, and consider the matter under
debate, and all shall be well betwixt thee and me.

Hopeful: But, Christian, these three fellows, I am persuaded in my
heart, are but a company of cowards: would they have run else, think
you, as they did, at the noise of one that was coming on the road? Why
did not Little-Faith pluck up a greater heart? He might, methinks, have
stood one brush with them, and have yielded when there had been no
remedy.

Christian: That they are cowards, many have said, but few have found it
so in the time of trial. As for a great heart, Little-Faith had none;
and I perceive by thee, my brother, hadst thou been the man concerned,
thou art but for a brush, and then to yield. And verily, since this is
the height of thy stomach now they are at a distance from us, should
they appear to thee as they did to him, they might put thee to second
thoughts.

But consider again, that they are but journeymen thieves; They serve
under the king of the bottomless pit, who, if need be, will come to
their aid himself, and his voice is as the roaring of a lion. 1 Pet.
5:8. I myself have been engaged as this Little-Faith was, and I found
it a terrible thing. These three villains set upon me, and I beginning
like a Christian to resist, they gave but a call, and in came their
master. I would, as the saying is, have given my life for a penny, but
that, as God would have it, I was clothed with armor of proof. Aye, and
yet, though I was so harnessed, I found it hard work to quit myself
like a man: no man can tell what in that combat attends us, but he that
hath been in the battle himself.

Hopeful: Well, but they ran, you see, when they did but suppose that
one Great-Grace was in the way.

Christian: True, they have often fled, both they and their master, when
Great-Grace hath but appeared; and no marvel, for he is the King's
champion. But I trow you will put some difference between Little-Faith
and the King's champion. All the King's subjects are not his champions;
nor can they, when tried, do such feats of war as he. Is it meet to
think that a little child should handle Goliath as David did? or that
there should be the strength of an ox in a wren? Some are strong, some
are weak; some have great faith, some have little: this man was one of
the weak, and therefore he went to the wall.

Hopeful: I would it had been Great-Grace, for their sakes.

Christian: If it had been he, he might have had his hands full: for I
must tell you, that though Great-Grace is excellent good at his
weapons, and has, and can, so long as he keeps them at sword's point,
do well enough with them; yet if they get within him, even Faint-Heart,
Mistrust, or the other, it shall go hard but they will throw up his
heels. And when a man is down, you know, what can he do?

Whoso looks well upon Great-Grace's face, will see those scars and cuts
there that shall easily give demonstration of what I say. Yea, once I
heard that he should say, (and that when he was in the combat,) We
despaired even of life. How did these sturdy rogues and their fellows
make David groan, mourn, and roar! Yea, Heman, Psa. 88, and Hezekiah
too, though champions in their days, were forced to bestir them when by
these assaulted; and yet, notwithstanding, they had their coats soundly
brushed by them. Peter, upon a time, would go try what he could do; but
though some do say of him that he is the prince of the apostles, they
handled him so that they made him at last afraid of a sorry girl.

Besides, their king is at their whistle; he is never out of hearing;
and if at any time they be put to the worst, he, if possible, comes in
to help them; and of him it is said, "The sword of him that layeth at
him cannot hold; the spear, the dart, nor the habergeon. He esteemeth
iron as straw, and brass as rotten wood. The arrow cannot make him fly;
sling-stones are turned with him into stubble. Darts are counted as
stubble; he laugheth at the shaking of a spear." Job 41:26-29. What can
a man do in this case? It is true, if a man could at every turn have
Job's horse, and had skill and courage to ride him, he might do notable
things. "For his neck is clothed with thunder. He will not be afraid as
a grasshopper: the glory of his nostrils is terrible. He paweth in the
valley, and rejoiceth in his strength; he goeth on to meet the armed
men. He mocketh at fear, and is not affrighted; neither turneth he back
from the sword. The quiver rattleth against him, the glittering spear
and the shield. He swalloweth the ground with fierceness and rage;
neither believeth he that it is the sound of the trumpet. He saith
among the trumpets, Ha, ha! and he smelleth the battle afar off, the
thunder of the captains, and the shoutings." Job 39:19-25.

But for such footmen as thee and I are, let us never desire to meet
with an enemy, nor vaunt as if we could do better, when we hear of
others that have been foiled, nor be tickled at the thoughts of our own
manhood; for such commonly come by the worst when tried. Witness Peter,
of whom I made mention before: he would swagger, aye, he would; he
would, as his vain mind prompted him to say, do better and stand more
for his Master than all men: but who so foiled and run down by those
villains as he?

When, therefore, we hear that such robberies are done on the King's
highway, two things become us to do.

1. To go out harnessed, and be sure to take a shield with us: for it
was for want of that, that he who laid so lustily at Leviathan could
not make him yield; for, indeed, if that be wanting, he fears us not at
all. Therefore, he that had skill hath said, "Above all, take the
shield of faith, wherewith ye shall be able to quench all the fiery
darts of the wicked." Eph. 6:16.

2. It is good, also, that we desire of the King a convoy, yea, that he
will go with us himself. This made David rejoice when in the Valley of
the Shadow of Death; and Moses was rather for dying where he stood,
than to go one step without his God. Exod. 33:15.

O, my brother, if he will but go along with us, what need we be afraid
of ten thousands that shall set themselves against us? Psa. 3:5-8;
27:1-3. But without him, the proud helpers fall under the slain. Isa.
10:4.

I, for my part, have been in the fray before now; and though (through
the goodness of Him that is best) I am, as you see, alive, yet I cannot
boast of any manhood. Glad shall I be if I meet with no more such
brunts; though I fear we are not got beyond all danger. However, since
the lion and the bear have not as yet devoured me, I hope God will also
deliver us from the next uncircumcised Philistine. Then sang Christian,


"Poor Little-Faith! hast been among the thieves?

Wast robb'd? Remember this, whoso believes,

And get more faith; then shall you victors be

Over ten thousand-else scarce over three."

So they went on, and Ignorance followed. They went then till they came
at a place where they saw a way put itself into their way, and seemed
withal to lie as strait as the way which they should go; and here they
knew not which of the two to take, for both seemed strait before them:
therefore here they stood still to consider. And as they were thinking
about the way, behold a man black of flesh, but covered with a very
light robe, come to them, and asked them why they stood there. They
answered, they were going to the Celestial City, but knew not which of
these ways to take. "Follow me," said the man, "it is thither that I am
going." So they followed him in the way that but now came into the
road, which by degrees turned, and turned them so far from the city
that they desired to go to, that in a little time their faces were
turned away from it; yet they follow him. But by and by, before they
were aware, he led them both within the compass of a net, in which they
were both so entangled that they knew not what to do; and with that the
white robe fell off the black man's back. Then they saw where they
were. Wherefore there they lay crying some time, for they could not get
themselves out.

Christian: Then said Christian to his fellow, Now do I see myself in an
error. Did not the shepherds bid us beware of the Flatterer? As is the
saying of the wise man, so we have found it this day: "A man that
flattereth his neighbor, spreadeth a net for his feet." Prov. 29:5.

Hopeful: They also gave us a note of directions about the way, for our
more sure finding thereof; but therein we have also forgotten to read,
and have not kept ourselves from the paths of the destroyer. Here David
was wiser than we; for saith he, "Concerning the works of men, by the
word of thy lips I have kept me from the paths of the Destroyer." Psa.
17:4. Thus they lay bewailing themselves in the net. At last they
espied a Shining One coming towards them with a whip of small cords in
his hand. When he was come to the place where they were, he asked them
whence they came, and what they did there. They told him that they were
poor pilgrims going to Zion, but were led out of their way by a black
man clothed in white, who bid us, said they, follow him, for he was
going thither too. Then said he with the whip, It is Flatterer, a false
apostle, that hath transformed himself into an angel of light. Dan.
11:32; 2 Cor. 11:13,14. So he rent the net, and let the men out. Then
said he to them, Follow me, that I may set you in your way again. So he
led them back to the way which they had left to follow the Flatterer.
Then he asked them, saying, Where did you lie the last night? They
said, With the shepherds upon the Delectable Mountains. He asked them
then if they had not of the shepherds a note of direction for the way.
They answered, Yes. But did you not, said he, when you were at a stand,
pluck out and read your note? They answered, No. He asked them, Why?
They said they forgot. He asked, moreover, if the shepherds did not bid
them beware of the Flatterer. They answered, Yes; but we did not
imagine, said they, that this fine-spoken man had been he. Rom.
16:17,18.

Then I saw in my dream, that he commanded them to lie down; which when
they did, he chastised them sore, to teach them the good way wherein
they should walk, Deut. 25:2; 2 Chron. 6:27; and as he chastised them,
he said, "As many as I love, I rebuke and chasten; be zealous,
therefore, and repent." Rev. 3:19. This done, he bids them to go on
their way, and take good heed to the other directions of the shepherds.
So they thanked him for all his kindness, and went softly along the
right way, singing,


"Come hither, you that walk along the way,

See how the pilgrims fare that go astray:

They catched are in an entangling net,

Cause they good counsel lightly did forget:

'Tis true, they rescued were; but yet, you see,

They're scouged to boot; let this your caution be."

Now, after awhile, they perceived afar off, one coming softly, and
alone, all along the highway, to meet them. Then said Christian to his
fellow, Yonder is a man with his back towards Zion, and he is coming to
meet us.

Hopeful: I see him; let us take heed to ourselves now, lest he should
prove a Flatterer also. So he drew nearer and nearer, and at last came
up to them. His name was Atheist, and he asked them whither they were
going.

Christian: We are going to Mount Zion.

Then Atheist fell into a very great laughter.

Christian: What's the meaning of your laughter?

Atheist: I laugh to see what ignorant persons you are, to take upon you
so tedious a journey, and yet are like to have nothing but your travel
for your pains.

Christian: Why, man, do you think we shall not be received?

Atheist: Received! There is not such a place as you dream of in all
this world.

Christian: But there is in the world to come.

Atheist: When I was at home in mine own country I heard as you now
affirm, and from that hearing went out to see, and have been seeking
this city these twenty years, but find no more of it than I did the
first day I set out. Eccles. 10:15; Jer. 17:15.

Christian: We have both heard, and believe, that there is such a place
to be found.

Atheist: Had not I, when at home, believed, I had not come thus far to
seek; but finding none, (and yet I should, had there been such a place
to be found, for I have gone to seek it farther than you,) I am going
back again, and will seek to refresh myself with the things that I then
cast away for hopes of that which I now see is not.

Christian: Then said Christian to Hopeful his companion, Is it true
which this man hath said?

Hopeful: Take heed, he is one of the Flatterers. Remember what it cost
us once already for our hearkening to such kind of fellows. What! no
Mount Zion? Did we not see from the Delectable Mountains the gate of
the city? Also, are we not now to walk by faith? 2 Cor. 5:7.

Let us go on, lest the man with the whip overtake us again. You should
have taught me that lesson, which I will sound you in the ears withal:
"Cease, my son, to hear the instruction that causeth to err from the
words of knowledge." Prov. 19:27. I say, my brother, cease to hear him,
and let us believe to the saving of the soul.

Christian: My brother, I did not put the question to thee, for that I
doubted of the truth of our belief myself, but to prove thee, and to
fetch from thee a fruit of the honesty of thy heart. As for this man, I
know that he is blinded by the God of this world. Let thee and me go
on, knowing that we have belief of the truth; and no lie is of the
truth. 1 John, 5:21.

Hopeful: Now do I rejoice in hope of the glory of God. So they turned
away from the man; and he, laughing at them, went his way.

I then saw in my dream, that they went on until they came into a
certain country whose air naturally tended to make one drowsy, if he
came a stranger into it. And here Hopeful began to be very dull, and
heavy to sleep: wherefore he said unto Christian, I do now begin to
grow so drowsy that I can scarcely hold open mine eyes; let us lie down
here, and take one nap.

Christian: By no means, said the other; lest, sleeping, we never awake
more.

Hopeful: Why, my brother? sleep is sweet to the laboring man; we may be
refreshed, if we take a nap.

Christian: Do you not remember that one of the shepherds bid us beware
of the Enchanted Ground? He meant by that, that we should beware of
sleeping; wherefore "let us not sleep, as do others; but let us watch
and be sober." 1 Thess. 5:6.

Hopeful: I acknowledge myself in a fault; and had I been here alone, I
had by sleeping run the danger of death. I see it is true that the wise
man saith, "Two are better than one." Eccl. 4:9. Hitherto hath thy
company been my mercy; and thou shalt have a good reward for thy labor.

Christian: Now, then, said Christian, to prevent drowsiness in this
place, let us fall into good discourse.

Hopeful: With all my heart, said the other.

Christian: Where shall we begin?

Hopeful: Where God began with us. But do you begin, if you please.

Christian: I will sing you first this song:


"When saints do sleepy grow, let them come hither,

And hear how these two pilgrims talk together;

Yea, let them learn of them in any wise,

Thus to keep ope their drowsy, slumb'ring eyes.

Saints' fellowship, if it be managed well,

Keeps them awake, and that in spite of hell."

Then Christian began, and said, I will ask you a question. How came you
to think at first of doing what you do now?

Hopeful: Do you mean, how came I at first to look after the good of my
soul?

Christian: Yes, that is my meaning.

Hopeful: I continued a great while in the delight of those things which
were seen and sold at our fair; things which I believe now would have,
had I continued in them still, drowned me in perdition and destruction.

Christian: What things were they?

Hopeful: All the treasures and riches of the world. Also I delighted
much in rioting, reveling, drinking, swearing, lying, uncleanness,
Sabbath-breaking, and what not, that tended to destroy the soul. But I
found at last, by hearing and considering of things that are divine,
which, indeed, I heard of you, as also of beloved Faithful, that was
put to death for his faith and good living in Vanity Fair, that the end
of these things is death, Rom. 6:21-23; and that for these things'
sake, the wrath of God cometh upon the children of disobedience. Eph.
5:6.

Christian: And did you presently fall under the power of this
conviction?

Hopeful: No, I was not willing presently to know the evil of sin, nor
the damnation that follows upon the commission of it; but endeavored,
when my mind at first began to be shaken with the word, to shut mine
eyes against the light thereof.

Christian: But what was the cause of your carrying of it thus to the
first workings of God's blessed Spirit upon you?

Hopeful: The causes were, 1. I was ignorant that this was the work of
God upon me. I never thought that by awakenings for sin, God at first
begins the conversion of a sinner. 2. Sin was yet very sweet to my
flesh, and I was loth to leave it. 3. I could not tell how to part with
mine old companions, their presence and actions were so desirable unto
me. 4. The hours in which convictions were upon me, were such
troublesome and such heart-affrighting hours, that I could not bear, no
not so much as the remembrance of them upon my heart.

Christian: Then, as it seems, sometimes you got rid of your trouble?

Hopeful: Yes, verily, but it would come into my mind again; and then I
should be as bad, nay, worse than I was before.

Christian: Why, what was it that brought your sins to mind again?

Hopeful: Many things; as,

1. If I did but meet a good man in the streets; or,

2. If I have heard any read in the Bible; or,

3. If mine head did begin to ache; or,

4. If I were told that some of my neighbors were sick; or,

5. If I heard the bell toll for some that were dead; or,

6. If I thought of dying myself; or,

7. If I heard that sudden death happened to others.

8. But especially when I thought of myself, that I must quickly come to
judgment.

Christian: And could you at any time, with ease, get off the guilt of
sin, when by any of these ways it came upon you?

Hopeful: No, not I; for then they got faster hold of my conscience; and
then, if I did but think of going back to sin, (though my mind was
turned against it,) it would be double torment to me.

Christian: And how did you do then?

Hopeful: I thought I must endeavor to mend my life; for else, thought
I, I am sure to be damned.

Christian: And did you endeavor to mend?

Hopeful: Yes, and fled from, not only my sins, but sinful company too,
and betook me to religious duties, as praying, reading, weeping for
sin, speaking truth to my neighbors, etc. These things did I, with many
others, too much here to relate.

Christian: And did you think yourself well then?

Hopeful: Yes, for a while; but at the last my trouble came tumbling
upon me again, and that over the neck of all my reformations.

Christian: How came that about, since you were now reformed?

Hopeful: There were several things brought it upon me, especially such
sayings as these: "All our righteousnesses are as filthy rags." Isa.
64:6. "By the works of the law shall no flesh be justified." Gal. 2:16.
"When ye have done all these things, say, We are unprofitable," Luke
17:10; with many more such like. From whence I began to reason with
myself thus: If all my righteousnesses are as filthy rags; if by the
deeds of the law no man can be justified; and if, when we have done
all, we are yet unprofitable, then is it but a folly to think of heaven
by the law. I farther thought thus: If a man runs a hundred pounds into
the shopkeeper's debt, and after that shall pay for all that he shall
fetch; yet if his old debt stands still in the book uncrossed, the
shopkeeper may sue him for it, and cast him into prison, till he shall
pay the debt.

Christian: Well, and how did you apply this to yourself?

Hopeful: Why, I thought thus with myself: I have by my sins run a great
way into God's book, and my now reforming will not pay off that score;
therefore I should think still, under all my present amendments, But
how shall I be freed from that damnation that I brought myself in
danger of by my former transgressions?

Christian: A very good application: but pray go on.

Hopeful: Another thing that hath troubled me ever since my late
amendments, is, that if I look narrowly into the best of what I do now,
I still see sin, new sin, mixing itself with the best of that I do; so
that now I am forced to conclude, that notwithstanding my former fond
conceits of myself and duties, I have committed sin enough in one day
to send me to hell, though my former life had been faultless.

Christian: And what did you do then?

Hopeful: Do! I could not tell what to do, until I broke my mind to
Faithful; for he and I were well acquainted. And he told me, that
unless I could obtain the righteousness of a man that never had sinned,
neither mine own, nor all the righteousness of the world, could save
me.

Christian: And did you think he spake true?

Hopeful: Had he told me so when I was pleased and satisfied with my own
amendments, I had called him fool for his pains; but now, since I see
my own infirmity, and the sin which cleaves to my best performance, I
have been forced to be of his opinion.

Christian: But did you think, when at first he suggested it to you,
that there was such a man to be found, of whom it might justly be said,
that he never committed sin?

Hopeful: I must confess the words at first sounded strangely; but after
a little more talk and company with him, I had full conviction about
it.

Christian: And did you ask him what man this was, and how you must be
justified by him?

Hopeful: Yes, and he told me it was the Lord Jesus, that dwelleth on
the right hand of the Most High. Heb. 10:12-21. And thus, said he, you
must be justified by him, even by trusting to what he hath done by
himself in the days of his flesh, and suffered when he did hang on the
tree. Rom. 4:5; Col. 1:14; 1 Pet. 1:19. I asked him further, how that
man's righteousness could be of that efficacy, to justify another
before God. And he told me he was the mighty God, and did what he did,
and died the death also, not for himself, but for me; to whom his
doings, and the worthiness of them, should be imputed, if I believed on
him.

Christian: And what did you do then?

Hopeful: I made my objections against my believing, for that I thought
he was not willing to save me.

Christian: And what said Faithful to you then?

Hopeful: He bid me go to him and see. Then I said it was presumption.
He said, No; for I was invited to come. Matt. 11:28. Then he gave me a
book of Jesus' inditing, to encourage me the more freely to come; and
he said concerning that book, that every jot and tittle thereof stood
firmer than heaven and earth. Matt. 24:35. Then I asked him what I must
do when I came; and he told me I must entreat upon my knees, Psa. 95:6;
Dan. 6:10, with all my heart and soul, Jer. 29:12,13, the Father to
reveal him to me. Then I asked him further, how I must make my
supplications to him; and he said, Go, and thou shalt find him upon a
mercy-seat, where he sits all the year long to give pardon and
forgiveness to them that come. Exod. 25:22; Lev. 16:2; Num. 7:89; Heb.
4:16. I told him, that I knew not what to say when I came; and he bid
say to this effect: God be merciful to me a sinner, and make me to know
and believe in Jesus Christ; for I see, that if his righteousness had
not been, or I have not faith in that righteousness, I am utterly cast
away. Lord, I have heard that thou art a merciful God, and hast
ordained that thy Son Jesus Christ should be the Saviour of the world;
and moreover, that thou art willing to bestow him upon such a poor
sinner as I am-and I am a sinner indeed. Lord, take therefore this
opportunity, and magnify thy grace in the salvation of my soul, through
thy Son Jesus Christ. Amen.

Christian: And did you do as you were bidden?

Hopeful: Yes, over, and over, and over.

Christian: And did the Father reveal the Son to you?

Hopeful: Not at the first, nor second, nor third, nor fourth, nor
fifth, no, nor at the sixth time neither.

Christian: What did you do then?

Hopeful: What? why I could not tell what to do.

Christian: Had you not thoughts of leaving off praying?

Hopeful: Yes; an hundred times twice told.

Christian: And what was the reason you did not?

Hopeful: I believed that it was true which hath been told me, to wit,
that without the righteousness of this Christ, all the world could not
save me; and therefore, thought I with myself, if I leave off, I die,
and I can but die at the throne of grace. And withal this came into my
mind, "If it tarry, wait for it; because it will surely come, and will
not tarry." Hab. 2:3. So I continued praying until the Father showed me
his Son.

Christian: And how was he revealed unto you?

Hopeful: I did not see him with my bodily eyes, but with the eyes of my
understanding, Eph. 1:18,19; and thus it was. One day I was very sad, I
think sadder than at any one time in my life; and this sadness was
through a fresh sight of the greatness and vileness of my sins. And as
I was then looking for nothing but hell, and the everlasting damnation
of my soul, suddenly, as I thought, I saw the Lord Jesus looking down
from heaven upon me, and saying, "Believe on the Lord Jesus Christ, and
thou shalt be saved." Acts 16:31.

But I replied, Lord, I am a great, a very great sinner: and he
answered, "My grace is sufficient for thee." 2 Cor. 12:9. Then I said,
But, Lord, what is believing? And then I saw from that saying, "He that
cometh to me shall never hunger, and he that believeth on me shall
never thirst," John 6:35, that believing and coming was all one; and
that he that came, that is, that ran out in his heart and affections
after salvation by Christ, he indeed believed in Christ. Then the water
stood in mine eyes, and I asked further, But, Lord, may such a great
sinner as I am be indeed accepted of thee, and be saved by thee? And I
heard him say, "And him that cometh to me, I will in no wise cast out."
John 6:37. Then I said, But how, Lord, must I consider of thee in my
coming to thee, that my faith may be placed aright upon thee? Then he
said, "Christ Jesus came into the world to save sinners." 1 Tim. 1:15.
He is the end of the law for righteousness to every one that believes.
Rom.10:4, and chap. 4. He died for our sins, and rose again for our
justification. Rom. 4:25. He loved us, and washed us from our sins in
his own blood. Rev. 1:5. He is the Mediator between God and us. 1 Tim.
2:5. He ever liveth to make intercession for us. Heb. 7:25. From all
which I gathered, that I must look for righteousness in his person, and
for satisfaction for my sins by his blood: that what he did in
obedience to his Father's law, and in submitting to the penalty
thereof, was not for himself, but for him that will accept it for his
salvation, and be thankful. And now was my heart full of joy, mine eyes
full of tears, and mine affections running over with love to the name,
people, and ways of Jesus Christ.

Christian: This was a revelation of Christ to your soul indeed. But
tell me particularly what effect this had upon your spirit.

Hopeful: It made me see that all the world, notwithstanding all the
righteousness thereof, is in a state of condemnation. It made me see
that God the Father, though he be just, can justly justify the coming
sinner. It made me greatly ashamed of the vileness of my former life,
and confounded me with the sense of mine own ignorance; for there never
came a thought into my heart before now that showed me so the beauty of
Jesus Christ. It made me love a holy life, and long to do something for
the honor and glory of the name of the Lord Jesus. Yea, I thought that
had I now a thousand gallons of blood in my body, I could spill it all
for the sake of the Lord Jesus.

I saw then in my dream, that Hopeful looked back, and saw Ignorance,
whom they had left behind, coming after. Look, said he to Christian,
how far yonder youngster loitereth behind.

Christian: Aye, aye, I see him: he careth not for our company.

Hopeful: But I trow it would not have hurt him, had he kept pace with
us hitherto.

Christian: That is true; but I warrant you he thinketh otherwise.

Hopeful: That I think he doth; but, however, let us tarry for him. (So
they did.)

Then Christian said to him, Come away, man; why do you stay so behind?

Ignorance: I take my pleasure in walking alone, even more a great deal
than in company, unless I like it the better.

Then said Christian to Hopeful, (but softly,) Did I not tell you he
cared not for our company? But, however, said he, come up, and let us
talk away the time in this solitary place. Then, directing his speech
to Ignorance, he said, Come, how do you do? How stands it between God
and your soul now?

Ignorance: I hope, well; for I am always full of good motions, that
come into my mind to comfort me as I walk.

Christian: What good motions? Pray tell us.

Ignorance: Why, I think of God and heaven.

Christian: So do the devils and damned souls.

Ignorance: But I think of them, and desire them.

Christian: So do many that are never like to come there. "The soul of
the sluggard desireth, and hath nothing." Prov. 13:4.

Ignorance: But I think of them, and leave all for them.

Christian: That I doubt: for to leave all is a very hard matter; yea, a
harder matter than many are aware of. But why, or by what, art thou
persuaded that thou hast left all for God and heaven?

Ignorance: My heart tells me so.

Christian: The wise man says, "He that trusteth in his own heart is a
fool." Prov. 28:26.

Ignorance: That is spoken of an evil heart; but mine is a good one.

Christian: But how dost thou prove that?

Ignorance: It comforts me in hopes of heaven.

Christian: That may be through its deceitfulness; for a man's heart may
minister comfort to him in the hopes of that thing for which he has yet
no ground to hope.

Ignorance: But my heart and life agree together; and therefore my hope
is well-grounded.

Christian: Who told thee that thy heart and life agree together?

Ignorance: My heart tells me so.

Christian: "Ask my fellow if I be a thief." Thy heart tells thee so!
Except the word of God beareth witness in this matter, other testimony
is of no value.

Ignorance: But is it not a good heart that hath good thoughts? and is
not that a good life that is according to God's commandments?

Christian: Yes, that is a good heart that hath good thoughts, and that
is a good life that is according to God's commandments; but it is one
thing indeed to have these, and another thing only to think so.

Ignorance: Pray, what count you good thoughts, and a life according to
God's commandments?

Christian: There are good thoughts of divers kinds; some respecting
ourselves, some God, some Christ, and some other things.

Ignorance: What be good thoughts respecting ourselves?

Christian: Such as agree with the word of God.

Ignorance: When do our thoughts of ourselves agree with the word of
God?

Christian: When we pass the same judgment upon ourselves which the word
passes. To explain myself: the word of God saith of persons in a
natural condition, "There is none righteous, there is none that doeth
good." It saith also, that, "every imagination of the heart of man is
only evil, and that continually." Gen. 6:5; Rom. 3. And again, "The
imagination of man's heart is evil from his youth." Gen. 8:21. Now,
then, when we think thus of ourselves, having sense thereof, then are
our thoughts good ones, because according to the word of God.

Ignorance: I will never believe that my heart is thus bad.

Christian: Therefore thou never hadst one good thought concerning
thyself in thy life. But let me go on. As the word passeth a judgment
upon our hearts, so it passeth a judgment upon our ways; and when the
thoughts of our hearts and ways agree with the judgment which the word
giveth of both, then are both good, because agreeing thereto.

Ignorance: Make out your meaning.

Christian: Why, the word of God saith, that man's ways are crooked
ways, not good but perverse; it saith, they are naturally out of the
good way, that they have not known it. Psa. 125:5; Prov. 2:15; Rom.
3:12. Now, when a man thus thinketh of his ways, I say, when he doth
sensibly, and with heart-humiliation, thus think, then hath he good
thoughts of his own ways, because his thoughts now agree with the
judgment of the word of God.

Ignorance: What are good thoughts concerning God?

Christian: Even, as I have said concerning ourselves, when our thoughts
of God do agree with what the word saith of him; and that is, when we
think of his being and attributes as the word hath taught, of which I
cannot now discourse at large. But to speak of him with reference to
us: then have we right thoughts of God when we think that he knows us
better than we know ourselves, and can see sin in us when and where we
can see none in ourselves; when we think he knows our inmost thoughts,
and that our heart, with all its depths, is always open unto his eyes;
also when we think that all our righteousness stinks in his nostrils,
and that therefore he cannot abide to see us stand before him in any
confidence, even in all our best performances.

Ignorance: Do you think that I am such a fool as to think that God can
see no further than I; or that I would come to God in the best of my
performances?

Christian: Why, how dost thou think in this matter?

Ignorance: Why, to be short, I think I must believe in Christ for
justification.

Christian: How! think thou must believe in Christ, when thou seest not
thy need of him! Thou neither seest thy original nor actual
infirmities; but hast such an opinion of thyself, and of what thou
doest, as plainly renders thee to be one that did never see the
necessity of Christ's personal righteousness to justify thee before
God. How, then, dost thou say, I believe in Christ?

Ignorance: I believe well enough, for all that.

Christian: How dost thou believe?

Ignorance: I believe that Christ died for sinners; and that I shall be
justified before God from the curse, through his gracious acceptance of
my obedience to his laws. Or thus, Christ makes my duties, that are
religious, acceptable to his Father by virtue of his merits, and so
shall I be justified.

Christian: Let me give an answer to this confession of thy faith.

1. Thou believest with a fantastical faith; for this faith is nowhere
described in the word.

2. Thou believest with a false faith; because it taketh justification
from the personal righteousness of Christ, and applies it to thy own.

3. This faith maketh not Christ a justifier of thy person, but of thy
actions; and of thy person for thy action's sake, which is false.

4. Therefore this faith is deceitful, even such as will leave thee
under wrath in the day of God Almighty: for true justifying faith puts
the soul, as sensible of its lost condition by the law, upon flying for
refuge unto Christ's righteousness; (which righteousness of his is not
an act of grace by which he maketh, for justification, thy obedience
accepted with God, but his personal obedience to the law, in doing and
suffering for us what that required at our hands;) this righteousness,
I say, true faith accepteth; under the skirt of which the soul being
shrouded, and by it presented as spotless before God, it is accepted,
and acquitted from condemnation.

Ignorance: What! would you have us trust to what Christ in his own
person has done without us? This conceit would loosen the reins of our
lust, and tolerate us to live as we list: for what matter how we live,
if we may be justified by Christ's personal righteousness from all,
when we believe it?

Christian: Ignorance is thy name, and as thy name is, so art thou: even
this thy answer demonstrateth what I say. Ignorant thou art of what
justifying righteousness is, and as ignorant how to secure thy soul,
through the faith of it, from the heavy wrath of God. Yea, thou also
art ignorant of the true effects of saving faith in this righteousness
of Christ, which is to bow and win over the heart to God in Christ, to
love his name, his word, ways, and people, and not as thou ignorantly
imaginest.

Hopeful: Ask him if ever he had Christ revealed to him from heaven.

Ignorance: What! you are a man for revelations! I do believe, that what
both you and all the rest of you say about that matter, is but the
fruit of distracted brains.

Hopeful: Why, man, Christ is so hid in God from the natural
apprehensions of the flesh, that he cannot by any man be savingly
known, unless God the Father reveals him to him.

Ignorance: That is your faith, but not mine, yet mine, I doubt not, is
as good as yours, though I have not in my head so many whimsies as you.

Christian: Give me leave to put in a word. You ought not so slightly to
speak of this matter: for this I will boldly affirm, even as my good
companion hath done, that no man can know Jesus Christ but by the
revelation of the Father: yea, and faith too, by which the soul layeth
hold upon Christ, (if it be right,) must be wrought by the exceeding
greatness of his mighty power, Matt. 11:27; 1 Cor. 12:3; Eph. 1:17-19;
the working of which faith, I perceive, poor Ignorance, thou art
ignorant of. Be awakened, then, see thine own wretchedness, and fly to
the Lord Jesus; and by his righteousness, which is the righteousness of
God, (for he himself is God,) thou shalt be delivered from
condemnation.

Ignorance: You go so fast I cannot keep pace with you; do you go on
before: I must stay a while behind.

Then they said,


"Well, Ignorance, wilt thou yet foolish be,

To slight good counsel, ten times given thee?

And if thou yet refuse it, thou shalt know,

Ere long, the evil of thy doing so.

Remember, man, in time: stoop, do not fear:

Good counsel, taken well, saves; therefore hear.

But if thou yet shalt slight it, thou wilt be

The loser, Ignorance, I'll warrant thee."
