\chapter[THE FOURTH STAGE]{}

Now when they had eaten and drank, and had chatted a little longer, their guide said to them, The day wears away; if you think good, let us prepare to be going. So they got up to go, and the little boys went before; But Christiana forgot to take her bottle of spirits with her, so she sent her little boy back to fetch it. Then said Mercy, I think this is a losing place: here Christian lost his roll, and here Christiana left her bottle behind her. Sir, what is the cause of this? So their guide made answer, and said, The cause is sleep, or forgetfulness: some sleep when they should keep awake, and some forget when they should remember; and this is the very cause why often, at the resting-places, some pilgrims in some things come off losers. Pilgrims should watch, and remember what they have already received, under their greatest enjoyments; but for want of doing so, oftentimes their rejoicing ends in tears, and their sunshine in a cloud: witness the story of Christian at this place.

When they were come to the place where Mistrust and Timorous met Christian, to persuade him to go back for fear of the lions, they perceived as it were a stage, and before it, towards the road, a broad plate with a copy of verses written thereon, and underneath the reason of raising up that stage in that place rendered. The verses were,
\begin{verse}
``Let him that sees this stage, take heed\\
Unto his heart and tongue;\\
Lest, if he do not, here he speed\\
As some have long agone."\\
\end{verse}

The words underneath the verses were, ``This stage was built to punish those upon, who, through timorousness or mistrust, shall be afraid to go further on pilgrimage. Also, on this stage both Mistrust and Timorous were burned through the tongue with a hot iron, for endeavoring to hinder Christian on his journey."

Then said Mercy, This is much like to the saying of the Beloved: ``What shall be given unto thee, or what shall be done unto thee, thou false tongue? Sharp arrows of the mighty, with coals of juniper. Psa. 120:3,4.

So they went on till they came within sight of the lions. Now Mr. Great-Heart was a strong man, so he was not afraid of a lion: But yet when they were come up to the place where the lions were, the boys, that went before, were now glad to cringe behind, for they were afraid of the lions; so they stepped back, and went behind. At this their guide smiled, and said, How now, my boys; do you love to go before when no danger doth approach, and love to come behind so soon as the lions appear?

Now, as they went on, Mr. Great-heart drew his sword, with intent to make a way for the pilgrims in spite of the lions. Then there appeared one that, it seems, had taken upon him to back the lions; and he said to the pilgrims' guide, What is the cause of your coming hither? Now the name of that man was Grim, or Bloody-man because of his slaying of pilgrims; and he was of the race of the giants.

Mr. Great-Heart: Then said the pilgrims' guide, These women and children are going on pilgrimage, and this is the way they must go; and go it they shall, in spite of thee and the lions.

Grim: This is not their way, neither shall they go therein. I am come forth to withstand them, and to that end will back the lions.

Now, to say the truth, by reason of the fierceness of the lions, and of the grim carriage of him that did back them, this way had of late lain much unoccupied, and was almost grown over with grass.

Christiana: Then said Christiana, Though the highways have been unoccupied heretofore, and though the travellers have been made in times past to walk through by-paths, it must not be so now I am risen, now I am risen a mother in Israel. Judges 5:6,7.

Grim: Then he swore by the lions that it should; and therefore bid them turn aside, for they should not have passage there.

But Great-Heart their guide made first his approach unto Grim, and laid so heavily on him with his sword that he forced him to retreat.

Grim: Then said he that attempted to back the lions, Will you slay me upon mine own ground?

Mr. Great-Heart: It is the King's highway that we are in, and in this way it is that thou hast placed the lions; but these women, and these children, though weak, shall hold on their way in spite of thy lions. And with that he gave him again a downright blow, and brought him upon his knees. With this blow also he broke his helmet, and with the next he cut off an arm. Then did the giant roar so hideously that his voice frightened the women, and yet they were glad to see him lie sprawling upon the ground. Now the lions were chained, and so of themselves could do nothing. Wherefore, when old Grim, that intended to back them, was dead, Mr. Great-Heart said to the pilgrims, Come now, and follow me, and no hurt shall happen to you from the lions. They therefore went on, but the women trembled as they passed by them; the boys also looked as if they would die; but they all got by without further hurt.

Now, when they were within sight of the Porter's lodge, they soon came up unto it; but they made the more haste after this to go thither, because it is dangerous traveling there in the night. So when they were come to the gate, the guide knocked, and the Porter cried, Who is there? But as soon as the guide had said, It is I, he knew his voice, and came down, for the guide had oft before that come thither as a conductor of pilgrims. When he was come down, he opened the gate; and seeing the guide standing just before it, (for he saw not the women, for they were behind him,) he said unto him, How now, Mr. Great-Heart, what is your business here so late at night? I have brought, said he, some pilgrims hither, where, by my Lord's commandment, they must lodge: I had been here some time ago, had I not been opposed by the giant that did use to back the lions. But I, after a long and tedious combat with him, have cut him off, and have brought the pilgrims hither in safety.

The Porter: Will you not go in, and stay till morning?

Mr. Great-Heart: No, I will return to my Lord to-night.

Christiana: O, sir, I know not how to be willing you should leave us in our pilgrimage: you have been so faithful and loving to us, you have fought so stoutly for us, you have been so hearty in counselling of us, that I shall never forget your favor towards us.

Mercy: Then said Mercy, O that we might have thy company to our journey's end! How can such poor women as we hold out in a way so full of troubles as this way is, without a friend and defender?

James: Then said James, the youngest of the boys, Pray, sir, be persuaded to go with us, and help us, because we are so weak, and the way so dangerous as it is.

Mr. Great-Heart: I am at my Lord's commandment; if he shall allot me to be your guide quite through, I will willingly wait upon you. But here you failed at first; for when he bid me come thus far with you, then you should have begged me of him to have gone quite through with you, and he would have granted your request. However, at present I must withdraw; and so, good Christiana, Mercy, and my brave children, adieu.

Then the Porter, Mr. Watchful, asked Christiana of her country, and of her kindred. And she said, I came from the city of Destruction. I am a widow woman, and my husband is dead, his name was Christian, the pilgrim. How! said the Porter, was he your husband? Yes, said she, and these are his children and this, pointing to Mercy, is one of my town's-women. Then the Porter rang his bell, as at such times he is wont, and there come to the door one of the damsels, whose name was Humble-Mind; and to her the Porter said, Go tell it within, that Christiana, the wife of Christian, and her children, are come hither on pilgrimage. She went in, therefore, and told it. But oh, what noise for gladness was there within when the damsel did but drop that out of her mouth!

So they came with haste to the Porter, for Christana stood still at the door. Then some of the most grave said unto her, Come in, Christiana, come in, thou wife of that good man; come in, thou blessed woman, come in, with all that are with thee. So she went in, and they followed her that were her children and companions. Now when they were gone in, they were had into a large room, where they were bidden to sit down: so they sat down, and the chief of the house were called to see and welcome the guests. Then they came in, and understanding who they were, did salute each other with a kiss, and said, Welcome, ye vessels of the grace of God; welcome to us, your friends.

Now, because it was somewhat late, and because the pilgrims were weary with their journey, and also made faint with the sight of the fight, and of the terrible lions, they desired, as soon as might be, to prepare to go to rest. Nay, said those of the family, refresh yourselves first with a morsel of meat; for they had prepared for them a lamb, with the accustomed sauce belonging thereto, Exod. 12:21; John 1:29; for the Porter had heard before of their coming, and had told it to them within. So when they had supped, and ended their prayer with a psalm, they desired they might go to rest.

But let us, said Christiana, if we may be so bold as to choose, be in that chamber that was my husband's when he was here; so they had them up thither, and they all lay in a room. When they were at rest, Christiana and Mercy entered into discourse about things that were convenient.

Christiana: Little did I think once, when my husband went on pilgrimage, that I should ever have followed him.

Mercy: And you as little thought of lying in his bed, and in his chamber to rest, as you do now.

Christiana: And much less did I ever think of seeing his face with comfort, and of worshiping the Lord the King with him; and yet now I believe I shall.

Mercy: Hark, don't you hear a noise?

Christiana: Yes, it is, as I believe, a noise of music, for joy that we are here.

Mercy: Wonderful! Music in the house, music in the heart, and music also in heaven, for joy that we are here! Thus they talked a while, and then betook themselves to sleep.

So in the morning when they were awake, Christiana said to Mercy, What was the matter that you did laugh in your sleep to-night? I suppose you were in a dream.

Mercy: So I was, and a sweet dream it was; but are you sure I laughed?

Christiana: Yes, you laughed heartily; but prithee, Mercy, tell me thy dream.

Mercy: I was a dreaming that I sat all alone in a solitary place, and was bemoaning of the hardness of my heart. Now I had not sat there long but methought many were gathered about me to see me, and to hear what it was that I said. So they hearkened, and I went on bemoaning the hardness of my heart. At this, some of them laughed at me, some called me fool, and some began to thrust me about. With that, methought I looked up and saw one coming with wings towards me. So he came directly to me, and said, Mercy, what aileth thee? Now when he had heard me make my complaint, he said, Peace be to thee; he also wiped my eyes with his handkerchief, and clad me in silver and gold. Ezek. 16:8-11. He put a chain about my neck, and ear-rings in mine ears, and a beautiful crown upon my head. Then he took me by the hand, and said, Mercy, come after me. So he went up, and I followed till we came at a golden gate. Then he knocked; and when they within had opened, the man went in, and I followed him up to a throne, upon which one sat; and he said to me, Welcome, daughter. The place looked bright and twinkling, like the stars, or rather like the sun, and I thought that I saw your husband there; so I awoke from my dream. But did I laugh?

Christiana: Laugh! aye, and well you might to see yourself so well. For you must give me leave to tell you that it was a good dream; and that, as you have begun to find the first part true, so you shall find the second at last. ``God speaks once, yea, twice, yet man perceiveth it not; in a dream, in a vision of the night, when deep sleep falleth upon men, in slumberings upon the bed." Job 33:14,15. We need not, when abed, to lie awake to talk with God; he can visit us while we sleep, and cause us then to hear his voice. Our heart oftentimes wakes when we sleep, and God can speak to that, either by words, by proverbs, by signs and similitudes, as well as if one was awake.

Mercy: Well, I am glad of my dream; for I hope ere long to see it fulfilled, to the making me laugh again.

Christiana: I think it is now high time to rise, and to know what we must do.

Mercy: Pray, if they invite us to stay a while, let us willingly accept of the proffer. I am the more willing to stay a while here, to grow better acquainted with these maids: methinks Prudence, Piety, and Charity, have very comely and sober countenances.

Christiana: We shall see what they will do.

So when they were up and ready, they came down, and they asked one another of their rest, and if it was comfortable or not.

Mercy: Very good, said Mercy: it was one of the best night's lodgings that ever I had in my life.

Then said Prudence and Piety, If you will be persuaded to stay here a while, you shall have what the house will afford.

Charity: Aye, and that with a very good will, said Charity. So they consented, and stayed there about a month or above, and became very profitable one to another. And because Prudence would see how Christiana had brought up her children, she asked leave of her to catechise them. So she gave her free consent. Then she began with her youngest, whose name was James.

Prudence: And she said, Come, James, canst thou tell me who made thee?

James: God the Father, God the Son, and God the Holy Ghost.

Prudence: Good boy. And canst thou tell who saved thee?

James: God the Father, God the Son, and God the Holy Ghost.

Prudence: Good boy still. But how doth God the Father save thee?

James: By his grace.

Prudence: How doth God the Son save thee?

James: By his righteousness, death and blood, and life.

Prudence: And how doth God the Holy Ghost save thee?

James: By his illumination, by his renovation, and by his preservation.

Then said Prudence to Christiana, You are to be commended for thus bringing up your children. I suppose I need not ask the rest these questions, since the youngest of them can answer them so well. I will therefore now apply myself to the next youngest.

Prudence: Then she said, Come, Joseph, (for his name was Joseph,) will you let me catechise you?

Joseph: With all my heart.

Prudence: What is man?

Joseph: A reasonable creature, so made by God, as my brother said.

Prudence: What is supposed by this word, saved?

Joseph: That man, by sin, has brought himself into a state of captivity and misery.

Prudence: What is supposed by his being saved by the Trinity?

Joseph: That sin is so great and mighty a tyrant that none can pull us out of its clutches but God; and that God is so good and loving to man, as to pull him indeed out of this miserable state.

Prudence: What is God's design in saving poor men?

Joseph: The glorifying of his name, of his grace, and justice, etc., and the everlasting happiness of his creature.

Prudence: Who are they that will be saved?

Joseph: They that accept of his salvation.

Prudence: Good boy, Joseph; thy mother hath taught thee well, and thou hast hearkened unto what she has said unto thee.

Then said Prudence to Samuel, who was the eldest but one,

Prudence: Come, Samuel, are you willing that I should catechise you?

Samuel: Yes, forsooth, if you please.

Prudence: What is heaven?

Samuel: A place and state most blessed, because God dwelleth there.

Prudence: What is hell?

Samuel: A place and state most woful, because it is the dwelling-place of sin, the devil, and death.

Prudence: Why wouldst thou go to heaven?

Samuel: That I may see God, and serve him without weariness; that I may see Christ, and love him everlastingly; that I may have that fullness of the Holy Spirit in me which I can by no means here enjoy.

Prudence: A very good boy, and one that has learned well.

Then she addressed herself to the eldest, whose name was Matthew; and she said to him, Come, Matthew, shall I also catechise you?

Matthew: With a very good will.

Prudence: I ask then, if there was ever any thing that had a being antecedent to or before God?

Matthew: No, for God is eternal; nor is there any thing, excepting himself, that had a being until the beginning of the first day. For in six days the Lord made heaven and earth, the sea, and all that in them is.

Prudence: What do you think of the Bible?

Matthew: It is the holy word of God.

Prudence: Is there nothing written therein but what you understand?

Matthew: Yes, a great deal.

Prudence: What do you do when you meet with places therein that you do not understand?

Matthew: I think God is wiser than I. I pray also that he will please to let me know all therein that he knows will be for my good.

Prudence: How believe you as touching the resurrection of the dead?

Matthew: I believe they shall rise the same that was buried; the same in nature, though not in corruption. And I believe this upon a double account: first, because God has promised it; secondly, because he is able to perform it.

Then said Prudence to the boys, You must still hearken to your mother; for she can teach you more. You must also diligently give ear to what good talk you shall hear from others: for your sakes do they speak good things. Observe also, and that with carefulness, what the heavens and the earth do teach you; but especially be much in the meditation of that book which was the cause of your father's becoming a pilgrim. I, for my part, my children, will teach you what I can while you are here, and shall be glad if you will ask me questions that tend to godly edifying.

Now by that these pilgrim's had been at this place a week, Mercy had a visitor that pretended some good-will unto her, and his name was Mr. Brisk; a man of some breeding, and that pretended to religion, but a man that stuck very close to the world. So he came once or twice, or more, to Mercy, and offered love unto her. Now Mercy was of a fair countenance, and therefore the more alluring.

Her mind also was to be always busying of herself in doing; for when she had nothing to do for herself, she would be making hose and garments for others, and would bestow them upon those that had need. And Mr. Brisk not knowing where or how she disposed of what she made, seemed to be greatly taken, for that he found her never idle. I will warrant her a good housewife, quoth he to himself.

Mercy then revealed the business to the maidens that were of the house, and inquired of them concerning him, for they did know him better than she. So they told her that he was a very busy young man, and one who pretended to religion, but was, as they feared, a stranger to the power of that which is good.

Nay then, said Mercy, I will look no more on him; for I purpose never to have a clog to my soul.

Prudence then replied, that there needed no matter of great discouragement to be given to him; her continuing so as she had begun to do for the poor, would quickly cool his courage.

So the next time he comes he finds her at her old work, making things for the poor. Then said he, What, always at it? Yes, said she, either for myself or for others. And what canst thou earn a day? said he. I do these things, said she, that I may be rich in good works, laying up in store for myself a good foundation against the time to come, that I may lay hold on eternal life. 1 Tim. 6:17-19. Why, prithee, what doest thou with them? said he. Clothe the naked, said she. With that his countenance fell. So he forbore to come at her again. And when he was asked the reason why, he said, that Mercy was a pretty lass, but troubled with ill conditions.

When he had left her, Prudence said, Did I not tell thee that Mr. Brisk would soon forsake thee? yea, he will rise up an ill report of thee; for, notwithstanding his pretence to religion, and his seeming love to Mercy, yet Mercy and he are of tempers so different that I believe they will never come together.

Mercy: I might have had husbands before now, though I spoke not of it to any; but they were such as did not like my conditions, though never did any of them find fault with my person. So they and I could not agree.

Prudence: Mercy in our days is but little set by any further than as to its name: the practice which is set forth by thy conditions, there are but few that can abide.

Mercy: Well, said Mercy, if nobody will have me, I will die unmarried, or my conditions shall be to me as a husband: for I cannot change my nature; and to have one who lies cross to me in this, that I purpose never to admit of as long as I live. I had a sister named Bountiful, that was married to one of these churls, but he and she could never agree; but because my sister was resolved to do as she had begun, that is, to show kindness to the poor, therefore her husband first cried her down at the cross, and then turned her out of his doors.

Prudence: And yet he was a professor, I warrant you?

Mercy: Yes, such a one as he was, and of such as he the world is now full: but I am for none of them all.

Now Matthew, the eldest son of Christiana, fell sick, and his sickness was sore upon him, for he was much pained in his bowels, so that he was with it at times pulled, as it were, both ends together. There dwelt also not far from thence one Mr. Skill, an ancient and well-approved physician. So Christiana desired it, and entered the room, and had a little observed the boy, he concluded that he was sick of the gripes. Then he said to his mother, What diet has Matthew of late fed upon? Diet! said Christiana, nothing but what is wholesome. The physician answered, This boy has been tampering with something that lies in his stomach undigested, and that will not away without means. And I tell you he must be purged, or else he will die.

Samuel: Then said Samuel, Mother, what was that which my brother did gather up and eat as soon as we were come from the gate that is at the head of this way? You know that there was an orchard on the left hand, on the other side of the wall, and some of the trees hung over the wall, and my brother did pluck and eat.

Christiana: True, my child, said Christiana, he did take thereof, and did eat: naughty boy as he was, I chid him, and yet he would eat thereof.

Mr. Skill: I knew he had eaten something that was not wholesome food; and that food, to wit, that fruit, is even the most hurtful of all. It is the fruit of Beelzebub's orchard. I do marvel that none did warn you of it; many have died thereof.

Christiana: Then Christiana began to cry; and she said, Oh, naughty boy! and Oh, careless mother! what shall I do for my son?

Mr. Skill: Come, do not be too much dejected; the boy may do well again, but he must purge and vomit.

Christiana: Pray, sir, try the utmost of your skill with him, whatever it costs.

Mr. Skill: Nay, I hope I shall be reasonable. So he made him a purge, but it was too weak; it was said it was made of the blood of a goat, the ashes of a heifer, and some of the juice of hyssop. Heb. 9:13, 19; 10: 1-4. When Mr. Skill had seen that that purge was too weak, he made one to the purpose. It was made ex carne et sanguine Christi,\footnote{Of the flesh and blood of Christ} John 6:54-57; Heb. 9:14; (you know physicians give strange medicines to their patients:) and it was made into pills, with a promise or two, and a proportionable quantity of salt. Mark 9:49. Now, he was to take them three at a time, fasting, in half a quarter of a pint of the tears of repentance. Zech. 12:10.

When this potion was prepared, and brought to the boy, he was loth to take it, though torn with the gripes as if he should be pulled in pieces. Come, come, said the physician, you must take it. It goes against my stomach, said the boy. I must have you take it, said his mother. I shall vomit it up again, said the boy. Pray, sir, said Christiana to Mr. Skill, how does it taste? It has no ill taste, said the doctor; and with that she touched one of the pills with the tip of her tongue. Oh, Matthew, said she, this potion is sweeter than honey. If thou lovest thy mother, if thou lovest thy brothers, if thou lovest Mercy, if thou lovest thy life, take it. So, with much ado, after a short prayer for the blessing of God upon it, he took it, and it wrought kindly with him. It caused him to purge; it caused him to sleep, and to rest quietly; it put him into a fine heat and breathing sweat, and did quite rid him of his gripes. So in a little time he got up, and walked about with a staff, and would go from room to room, and talk with Prudence, Piety, and Charity, of his distemper, and how he was healed.

So when the boy was healed, Christiana asked Mr. Skill, saying, Sir, what will content you for your pains and care to and of my child? And he said, You must pay the master of the College of Physicians, Heb. 13:11-15, according to rules made in that case and provided.

Christiana: But, sir, said she, what is this pill good for else?

Mr. Skill: It is a universal pill; it is good against all the diseases that pilgrims are incident to; and when it is well prepared, it will keep good, time out of mind.

Christiana: Pray, sir, make me up twelve boxes of them; for if I can get these, I will never take other physic.

Mr. Skill: These pills are good to prevent diseases, as well as to cure when one is sick. Yea, I dare say it, and stand to it, that if a man will but use this physic as he should, it will make him live for ever. John 6:51. But, good Christiana, thou must give these pills no other way but as I have prescribed; for if you do, they will do no good. So he gave unto Christiana physic for herself, and her boys, and for Mercy; and bid Matthew take heed how he ate any more green plums; and kissed them, and went his way.

It was told you before, that Prudence bid the boys, that if at any time they would, they should ask her some questions that might be profitable and she would say something to them.

Matthew: Then Matthew, who had been sick, asked her, why for the most part physic should be bitter to our palates.

Prudence: To show how unwelcome the word of God and the effects thereof are to a carnal heart.

Matthew: Why does physic, if it does good, purge, and cause to vomit?

Prudence: To show that the word, when it works effectually, cleanseth the heart and mind. For look, what the one doth to the body, the other doth to the soul.

Matthew: What should we learn by seeing the flame of our fire go upwards, and by seeing the beams and sweet influences of the sun strike downwards?

Prudence: By the going up of the fire, we are taught to ascend to heaven by fervent and hot desires. And by the sun sending his heat, beams, and sweet influences downwards, we are taught the Saviour of the world, though high, reaches down with his grace and love to us below.

Matthew: Whence have the clouds their water?

Prudence: Out of the sea.

Matthew: What may we learn from that?

Prudence: That ministers should fetch their doctrine from God.

Matthew: Why do they empty themselves upon the earth?

Prudence: To show that ministers should give out what they know of God to the world.

Matthew: Why is the rainbow caused by the sun?

Prudence: To show that the covenant of God's grace is confirmed to us in Christ.

Matthew: Why do the springs come from the sea to us through the earth?

Prudence: To show that the grace of God comes to us through the body of Christ.

Matthew: Why do some of the springs rise out of the tops of high hills?

Prudence: To show that the Spirit of grace shall spring up in some that are great and mighty, as well as in many that are poor and low.

Matthew: Why doth the fire fasten upon the candle-wick?

Prudence: To show that unless grace doth kindle upon the heart, there will be no true light of life in us.

Matthew: Why are the wick, and tallow and all, spent to maintain the light of the candle?

Prudence: To show that body and soul, and all, should be at the service of, and spend themselves to maintain in good condition that grace of God that is in us.

Matthew: Why doth the pelican pierce her own breast with her bill?

Prudence: To nourish her young ones with her blood, and thereby to show that Christ the blessed so loved his young, (his people,) as to save them from death by his blood.

Matthew: What may one learn by hearing the cock to crow?

Prudence: Learn to remember Peter's sin, and Peter's repentance. The cock's crowing shows also, that day is coming on: let, then, the crowing of the cock put thee in mind of that last and terrible day of judgment.

Now about this time their month was out; wherefore they signified to those of the house, that it was convenient for them to up and be going. Then said Joseph to his mother, It is proper that you forget not to send to the house of Mr. Interpreter, to pray him to grant that Mr. Great-Heart should be sent unto us, that he may be our conductor for the rest of the way. Good boy, said she, I had almost forgot. So she drew up a petition, and prayed Mr. Watchful the porter to send it by some fit man to her good friend Mr. Interpreter; who, when it was come, and he had seen the contents of the petition, said to the messenger, Go, tell them that I will send him.

When the family where Christiana was, saw that they had a purpose to go forward, they called the whole house together, to give thanks to their King for sending of them such profitable guests as these. Which done, they said unto Christiana, And shall we not show thee something, as our custom is to do to pilgrims, on which thou mayest meditate when thou art upon the way? So they took Christiana, her children, and Mercy, into the closet, and showed them one of the apples that Eve ate of, and that she also did give to her husband, and that for the eating of which they were both turned out of paradise, and asked her what she thought that was. Then Christiana said, It is food or poison, I know not which. So they opened the matter to her, and she held up her hands and wondered. Gen. 3:6; Rom. 7:24.

Then they had her to a place, and showed her Jacob's ladder. Gen. 28:12. Now at that time there were some angels ascending upon it. So Christiana looked and looked to see the angels go up: so did the rest of the company. Then they were going into another place, to show them something else; but James said to his mother, Pray, bid them stay here a little longer, for this is a curious sight. So they turned again, and stood feeding their eyes with this so pleasant a prospect.

After this, they had them into a place where did hang up a golden anchor. So they bid Christiana take it down; for said they, You shall have it with you, for it is of absolute necessity that you should, that you may lay hold of that within the veil, Heb. 6:19, and stand stedfast in case you should meet with turbulent weather, Joel 3:16: so they were glad thereof.

Then they took them, and had them to the mount upon which Abraham our father offered up Isaac his son, and showed them the altar, the wood, the fire, and the knife, for they remain to be seen to this very day. Gen. 22:9. When they had seen it, they held up their hands, and blessed themselves, and said, Oh, what a man for love to his Master, and for denial to himself, was Abraham!

After they had showed them all these things, Prudence took them into a dining room, where stood a pair of excellent virginals;\footnote{A musical instrument.} so she played upon them, and turned what she had showed them into this excellent song, saying,
\begin{verse}
``Eve's apple we have showed you;\\
Of that be you aware:\\
You have seen Jacob's ladder too,\\
Upon which angels are.\\
An anchor you received have;\\
But let not these suffice,\\
Until with Abra'm you have gave\\
Your best, a sacrifice."\\
\end{verse}

Now, about this time, one knocked at the door; so the Porter opened, and behold, Mr. Great-Heart was there. But when he was come in, what joy was there! for it came now afresh again into their minds, how but a while ago he had slain old Grim Bloody-man the giant, and had delivered them from the lions.

Then said Mr. Great-Heart to Christiana and to Mercy, My Lord has sent each of you a bottle of wine, and also some parched corn, together with a couple of pomegranates; he has also sent the boys some figs and raisins; to refresh you in your way.

Then they addressed themselves to their journey, and Prudence and Piety went along with them. When they came to the gate, Christiana asked the Porter if any of late went by. He said, No; only one, some time since, who also told me, that of late there had been a great robbery committed on the King's highway as you go. But, said he, the thieves are taken, and will shortly be tried for their lives. Then Christiana and Mercy were afraid; but Matthew said, Mother, fear nothing, as long as Mr. Great-Heart is to go with us, and to be our conductor.

Then said Christiana to the Porter, Sir, I am much obliged to you for all the kindnesses that you have showed to me since I came hither; and also for that you have been so loving and kind to my children. I know not how to gratify your kindness; wherefore, pray, as a token of my respect to you, accept of this small mite. So she put a gold angel\footnote{A gold angel was a coin of the value of ten shillings sterling} in his hand; and he made her a low obeisance, and said, ``Let thy garments be always white; and let thy head want no ointment." Eccles. 9:8. Let Mercy live and not die, and let not her works be few. Deut. 33:6. And to the boys he said, Do you fly youthful lusts, and follow after godliness with them that are grave and wise, 2 Tim. 2:22: so shall you put gladness into your mother's heart, and obtain praise of all that are sober-minded. So they thanked the Porter, and departed. 
