\chapter[THE THIRD STAGE]{}

Now I saw in my dream, that the highway up which Christian was to go, was fenced on either side with a wall, and that wall was called Salvation. Isaiah 26:1. Up this way, therefore, did burdened Christian run, but not without great difficulty, because of the load on his back.

He ran thus till he came at a place somewhat ascending; and upon that place stood a cross, and a little below, in the bottom, a sepulchre. So I saw in my dream, that just as Christian came up with the cross, his burden loosed from off his shoulders, and fell from off his back, and began to tumble, and so continued to do till it came to the mouth of the sepulchre, where it fell in, and I saw it no more.

Then was Christian glad and lightsome, and said with a merry heart, ``He hath given me rest by his sorrow, and life by his death." Then he stood still a while, to look and wonder; for it was very surprising to him that the sight of the cross should thus ease him of his burden. He looked, therefore, and looked again, even till the springs that were in his head sent the waters down his cheeks. Zech. 12:10. Now as he stood looking and weeping, behold, three Shining Ones came to him, and saluted him with, ``Peace be to thee." So the first said to him, ``Thy sins be forgiven thee," Mark 2:5; the second stripped him of his rags, and clothed him with change of raiment, Zech. 3:4; the third also set a mark on his forehead, Eph. 1:13, and gave him a roll with a seal upon it, which he bid him look on as he ran, and that he should give it in at the celestial gate: so they went their way. Then Christian gave three leaps for joy, and went on singing,
\begin{verse}
``Thus far did I come laden with my sin,\\
Nor could aught ease the grief that I was in,\\
Till I came hither. What a place is this!\\
Must here be the beginning of my bliss?\\
Must here the burden fall from off my back?\\
Must here the strings that bound it to me crack?\\
Blest cross! blest sepulchre! blest rather be\\
The Man that there was put to shame for me!"\\
\end{verse}
I saw then in my dream, that he went on thus, even until he came at the bottom, where he saw, a little out of the way, three men fast asleep, with fetters upon their heels. The name of the one was Simple, of another Sloth, and of the third Presumption.

Christian then seeing them lie in this case, went to them, if peradventure he might awake them, and cried, you are like them that sleep on the top of a mast, Prov. 23:34, for the Dead Sea is under you, a gulf that hath no bottom: awake, therefore, and come away; be willing also, and I will help you off with your irons. He also told them, If he that goeth about like a roaring lion, 1 Pet. 5:8, comes by, you will certainly become a prey to his teeth. With that they looked upon him, and began to reply in this sort: Simple said, I see no danger; Sloth said, Yet a little more sleep; and Presumption said, Every tub must stand upon its own bottom. And so they lay down to sleep again, and Christian went on his way.

Yet he was troubled to think that men in that danger should so little esteem the kindness of him that so freely offered to help them, both by awakening of them, counselling of them, and proffering to help them off with their irons. And as he was troubled thereabout, he espied two men come tumbling over the wall, on the left hand of the narrow way; and they made up apace to him. The name of the one was Formalist, and the name of the other Hypocrisy. So, as I said, they drew up unto him, who thus entered with them into discourse.

Christian: Gentlemen, whence came you, and whither do you go?

Formalist and Hypocrisy: We were born in the land of Vain-glory, and are going, for praise, to Mount Zion.

Christian: Why came you not in at the gate which standeth at the beginning of the way? Know ye not that it is written, that ``he that cometh not in by the door, but climbeth up some other way, the same is a thief and a robber?" John 10:1.

Formalist and Hypocrisy: They said, that to go to the gate for entrance was by all their countrymen counted too far about; and that therefore their usual way was to make a short cut of it, and to climb over the wall, as they had done.

Christian: But will it not be counted a trespass against the Lord of the city whither we are bound, thus to violate his revealed will?

Formalist and Hypocrisy: They told him, that as for that, he needed not to trouble his head thereabout: for what they did they had custom for, and could produce, if need were, testimony that would witness it for more than a thousand years.

Christian: But, said Christian, will you stand a trial at law?

Formalist and Hypocrisy: They told him, that custom, it being of so long standing as above a thousand years, would doubtless now be admitted as a thing legal by an impartial judge: and besides, said they, if we get into the way, what matter is it which way we get in? If we are in, we are in: thou art but in the way, who, as we perceive, came in at the gate; and we also are in the way, that came tumbling over the wall: wherein now is thy condition better than ours?

Christian: I walk by the rule of my Master: you walk by the rude working of your fancies. You are counted thieves already by the Lord of the way: therefore I doubt you will not be found true men at the end of the way. You come in by yourselves without his direction, and shall go out by yourselves without his mercy.

To this they made him but little answer; only they bid him look to himself. Then I saw that they went on, every man in his way, without much conference one with another, save that these two men told Christian, that as to laws and ordinances, they doubted not but that they should as conscientiously do them as he. Therefore, said they, we see not wherein thou differest from us, but by the coat that is on thy back, which was, as we trow, given thee by some of thy neighbors, to hide the shame of thy nakedness.

Christian: By laws and ordinances you will not be saved, since you came not in by the door. Gal. 2:16. And as for this coat that is on my back, it was given me by the Lord of the place whither I go; and that, as you say, to cover my nakedness with. And I take it as a token of kindness to me; for I had nothing but rags before. And besides, thus I comfort myself as I go. Surely, think I, when I come to the gate of the city, the Lord thereof will know me for good, since I have his coat on my back; a coat that he gave me freely in the day that he stripped me of my rags. I have, moreover, a mark in my forehead, of which perhaps you have taken no notice, which one of my Lord's most intimate associates fixed there in the day that my burden fell off my shoulders. I will tell you, moreover, that I had then given me a roll sealed, to comfort me by reading as I go on the way; I was also bid to give it in at the celestial gate, in token of my certain going in after it: all which things I doubt you want, and want them because you came not in at the gate.

To these things they gave him no answer; only they looked upon each other, and laughed. Then I saw that they went all on, save that Christian kept before, who had no more talk but with himself, and that sometimes sighingly, and sometimes comfortably: also he would be often reading in the roll that one of the Shining Ones gave him, by which he was refreshed.

I beheld then, that they all went on till they came to the foot of the hill Difficulty, at the bottom of which there was a spring. There were also in the same place two other ways besides that which came straight from the gate: one turned to the left hand, and the other to the right, at the bottom of the hill; but the narrow way lay right up the hill, and the name of the going up the side of the hill is called Difficulty. Christian now went to the spring, Isa. 49:10, and drank thereof to refresh himself, and then began to go up the hill, saying,
\begin{verse}
``The hill, though high, I covet to ascend;\\
The difficulty will not me offend;\\
For I perceive the way to life lies here:\\
Come, pluck up heart, let's neither faint nor fear.\\
Better, though difficult, the right way to go,\\
Than wrong, though easy, where the end is woe."\\
\end{verse}

The other two also came to the foot of the hill. But when they saw that the hill was steep and high, and that there were two other ways to go; and supposing also that these two ways might meet again with that up which Christian went, on the other side of the hill; therefore they were resolved to go in those ways. Now the name of one of those ways was Danger, and the name of the other Destruction. So the one took the way which is called Danger, which led him into a great wood; and the other took directly up the way to Destruction, which led him into a wide field, full of dark mountains, where he stumbled and fell, and rose no more.

I looked then after Christian, to see him go up the hill, where I perceived he fell from running to going, and from going to clambering upon his hands and his knees, because of the steepness of the place. Now about the midway to the top of the hill was a pleasant Arbor, made by the Lord of the hill for the refreshment of weary travellers. Thither, therefore, Christian got, where also he sat down to rest him: then he pulled his roll out of his bosom, and read therein to his comfort; he also now began afresh to take a review of the coat or garment that was given to him as he stood by the cross. Thus pleasing himself awhile, he at last fell into a slumber, and thence into a fast sleep, which detained him in that place until it was almost night; and in his sleep his roll fell out of his hand. Now, as he was sleeping, there came one to him, and awaked him, saying, ``Go to the ant, thou sluggard; consider her ways, and be wise." Prov. 6:6. And with that, Christian suddenly started up, and sped him on his way, and went apace till he came to the top of the hill.

Now when he was got up to the top of the hill, there came two men running amain; the name of the one was Timorous, and of the other Mistrust: to whom Christian said, Sirs, what's the matter? you run the wrong way. Timorous answered, that they were going to the city of Zion, and had got up that difficult place: but, said he, the farther we go, the more danger we meet with; wherefore we turned, and are going back again.

Yes, said Mistrust, for just before us lie a couple of lions in the way, whether sleeping or waking we know not; and we could not think, if we came within reach, but they would presently pull us in pieces.

Christian: Then said Christian, You make me afraid; but whither shall I fly to be safe? If I go back to mine own country, that is prepared for fire and brimstone, and I shall certainly perish there; if I can get to the celestial city, I am sure to be in safety there: I must venture. To go back is nothing but death: to go forward is fear of death, and life everlasting beyond it: I will yet go forward. So Mistrust and Timorous ran down the hill, and Christian went on his way. But thinking again of what he had heard from the men, he felt in his bosom for his roll, that he might read therein and be comforted; but he felt, and found it not. Then was Christian in great distress, and knew not what to do; for he wanted that which used to relieve him, and that which should have been his pass into the celestial city. Here, therefore, he began to be much perplexed, and knew not what to do. At last he bethought himself that he had slept in the arbor that is on the side of the hill; and falling down upon his knees, he asked God forgiveness for that foolish act, and then went back to look for his roll. But all the way he went back, who can sufficiently set forth the sorrow of Christian's heart? Sometimes he sighed, sometimes he wept, and oftentimes he chid himself for being so foolish to fall asleep in that place, which was erected only for a little refreshment from his weariness. Thus, therefore, he went back, carefully looking on this side and on that, all the way as he went, if happily he might find his roll, that had been his comfort so many times in his journey. He went thus till he came again in sight of the arbor where he sat and slept; but that sight renewed his sorrow the more, by bringing again, even afresh, his evil of sleeping unto his mind. Rev. 2:4; 1 Thess. 5:6-8. Thus, therefore, he now went on, bewailing his sinful sleep, saying, O wretched man that I am, that I should sleep in the daytime! that I should sleep in the midst of difficulty! that I should so indulge the flesh as to use that rest for ease to my flesh which the Lord of the hill hath erected only for the relief of the spirits of pilgrims! How many steps have I taken in vain! Thus it happened to Israel; for their sin they were sent back again by the way of the Red Sea; and I am made to tread those steps with sorrow, which I might have trod with delight, had it not been for this sinful sleep. How far might I have been on my way by this time! I am made to tread those steps thrice over, which I needed not to have trod but once: yea, now also I am like to be benighted, for the day is almost spent. O that I had not slept!

Now by this time he was come to the arbor again, where for a while he sat down and wept; but at last, (as Providence would have it,) looking sorrowfully down under the settle, there he espied his roll, the which he with trembling and haste catched up, and put it into his bosom. But who can tell how joyful this man was when he had gotten his roll again? For this roll was the assurance of his life, and acceptance at the desired haven. Therefore he laid it up in his bosom, gave thanks to God for directing his eye to the place where it lay, and with joy and tears betook himself again to his journey. But O how nimbly did he go up the rest of the hill! Yet before he got up, the sun went down upon Christian; and this made him again recall the vanity of his sleeping to his remembrance; and thus he again began to condole with himself: Oh thou sinful sleep! how for thy sake am I like to be benighted in my journey! I must walk without the sun, darkness must cover the path of my feet, and I must hear the noise of the doleful creatures, because of my sinful sleep! Now also he remembered the story that Mistrust and Timorous told him of, how they were frighted with the sight of the lions. Then said Christian to himself again, These beasts range in the night for their prey; and if they should meet with me in the dark, how should I shift them? how should I escape being by them torn in pieces? Thus he went on his way. But while he was bewailing his unhappy miscarriage, he lift up his eyes, and behold there was a very stately palace before him, the name of which was Beautiful, and it stood by the highway-side.

So I saw in my dream that he made haste, and went forward, that if possible he might get lodging there. Now before he had gone far, he entered into a very narrow passage, which was about a furlong off the Porter's lodge, and looking very narrowly before him as he went, he espied two lions in the way. Now, thought he, I see the dangers that Mistrust and Timorous were driven back by. (The lions were chained, but he saw not the chains.) Then he was afraid, and thought also himself to go back after them; for he thought nothing but death was before him. But the Porter at the lodge, whose name is Watchful, perceiving that Christian made a halt, as if he would go back, cried unto him, saying, Is thy strength so small? Mark 4:40. Fear not the lions, for they are chained, and are placed there for trial of faith where it is, and for discovery of those that have none: keep in the midst of the path, and no hurt shall come unto thee.

Then I saw that he went on, trembling for fear of the lions, but taking good heed to the directions of the Porter; he heard them roar, but they did him no harm. Then he clapped his hands, and went on till he came and stood before the gate where the Porter was. Then said Christian to the Porter, Sir, what house is this? and may I lodge here to-night? The Porter answered, This house was built by the Lord of the hill, and he built it for the relief and security of pilgrims. The Porter also asked whence he was, and whither he was going.

Christian: I am come from the city of Destruction, and am going to Mount Zion: but because the sun is now set, I desire, if I may, to lodge here to-night.

The Porter: What is your name?

Christian: My name is now Christian, but my name at the first was Graceless: I came of the race of Japheth, whom God will persuade to dwell in the tents of Shem. Gen. 9:27.

The Porter: But how does it happen that you come so late? The sun is set.

Christian: I had been here sooner, but that, wretched man that I am, I slept in the arbor that stands on the hill-side! Nay, I had, notwithstanding that, been here much sooner, but that in my sleep I lost my evidence, and came without it to the brow of the hill; and then feeling for it, and not finding it, I was forced with sorrow of heart to go back to the place where I slept my sleep, where I found it; and now I am come.

The Porter: Well, I will call out one of the virgins of this place, who will, if she likes your talk, bring you in to the rest of the family, according to the rules of the house. So Watchful the Porter rang a bell, at the sound of which came out of the door of the house a grave and beautiful damsel, named Discretion, and asked why she was called.

The Porter answered, This man is on a journey from the city of Destruction to Mount Zion; but being weary and benighted, he asked me if he might lodge here to-night: so I told him I would call for thee, who, after discourse had with him, mayest do as seemeth thee good, even according to the law of the house.

Then she asked him whence he was, and whither he was going; and he told her. She asked him also how he got into the way; and he told her. Then she asked him what he had seen and met with in the way, and he told her. And at last she asked his name. So he said, It is Christian; and I have so much the more a desire to lodge here to-night, because, by what I perceive, this place was built by the Lord of the hill for the relief and security of pilgrims. So she smiled, but the water stood in her eyes; and after a little pause she said, I will call forth two or three more of the family. So she ran to the door, and called out Prudence, Piety, and Charity, who, after a little more discourse with him, had him into the family; and many of them meeting him at the threshold of the house, said, Come in, thou blessed of the Lord; this house was built by the Lord of the hill on purpose to entertain such pilgrims in. Then he bowed his head, and followed them into the house. So when he was come in and sat down, they gave him something to drink, and consented together that, until supper was ready, some of them should have some particular discourse with Christian, for the best improvement of time; and they appointed Piety, Prudence, and Charity to discourse with him: and thus they began.

Piety: Come, good Christian, since we have been so loving to you as to receive you into our house this night, let us, if perhaps we may better ourselves thereby, talk with you of all things that have happened to you in your pilgrimage.

Christian: With a very good will; and I am glad that you are so well disposed.

Piety: What moved you at first to betake yourself to a pilgrim's life?

Christian: I was driven out of my native country by a dreadful sound that was in mine ears; to wit, that unavoidable destruction did attend me, if I abode in that place where I was.

Piety: But how did it happen that you came out of your country this way?

Christian: It was as God would have it; for when I was under the fears of destruction, I did not know whither to go; but by chance there came a man, even to me, as I was trembling and weeping, whose name is Evangelist, and he directed me to the Wicket-gate, which else I should never have found, and so set me into the way that hath led me directly to this house.

Piety: But did you not come by the house of the Interpreter?

Christian: Yes, and did see such things there, the remembrance of which will stick by me as long as I live, especially three things: to wit, how Christ, in despite of Satan, maintains his work of grace in the heart; how the man had sinned himself quite out of hopes of God's mercy; and also the dream of him that thought in his sleep the day of judgment was come.

Piety: Why, did you hear him tell his dream?

Christian: Yes, and a dreadful one it was, I thought; it made my heart ache as he was telling of it, but yet I am glad I heard it.

Piety: Was this all you saw at the house of the Interpreter?

Christian: No; he took me, and had me where he showed me a stately palace, and how the people were clad in gold that were in it; and how there came a venturous man, and cut his way through the armed men that stood in the door to keep him out; and how he was bid to come in, and win eternal glory. Methought those things did ravish my heart. I would have stayed at that good man's house a twelvemonth, but that I knew I had farther to go.

Piety: And what saw you else in the way?

Christian: Saw? Why, I went but a little farther, and I saw One, as I thought in my mind, hang bleeding upon a tree; and the very sight of him made my burden fall off my back; for I groaned under a very heavy burden, but then it fell down from off me. It was a strange thing to me, for I never saw such a thing before: yea, and while I stood looking up, (for then I could not forbear looking,) three Shining Ones came to me. One of them testified that my sins were forgiven me; another stripped me of my rags, and gave me this broidered coat which you see; and the third set the mark which you see in my forehead, and gave me this sealed roll, (and with that he plucked it out of his bosom.)

Piety: But you saw more than this, did you not?

Christian: The things that I have told you were the best: yet some other I saw, as, namely, I saw three men, Simple, Sloth, and Presumption, lie asleep, a little out of the way, as I came, with irons upon their heels; but do you think I could awake them? I also saw Formality and Hypocrisy come tumbling over the wall, to go, as they pretended, to Zion; but they were quickly lost, even as I myself did tell them, but they would not believe. But, above all, I found it hard work to get up this hill, and as hard to come by the lions' mouths; and, truly, if it had not been for the good man, the porter that stands at the gate, I do not know but that, after all, I might have gone back again; but I thank God I am here, and thank you for receiving me.

Then Prudence thought good to ask him a few questions, and desired his answer to them.

Prudence: Do you not think sometimes of the country from whence you came?

Christian: Yea, but with much shame and detestation. Truly, if I had been mindful of that country from whence I came out, I might have had opportunity to have returned; but now I desire a better country, that is, a heavenly one. Heb. 11:15,16.

Prudence: Do you not yet bear away with you some of the things that then you were conversant withal?

Christian: Yes, but greatly against my will; especially my inward and carnal cogitations, with which all my countrymen, as well as myself, were delighted. But now all those things are my grief; and might I but choose mine own things, I would choose never to think of those things more: but when I would be a doing that which is best, that which is worst is with me. Rom. 7:15, 21.

Prudence: Do you not find sometimes as if those things were vanquished, which at other times are your perplexity?

Christian: Yes, but that is but seldom; but they are to me golden hours in which such things happen to me.

Prudence: Can you remember by what means you find your annoyances at times as if they were vanquished?

Christian: Yes: when I think what I saw at the cross, that will do it; and when I look upon my broidered coat, that will do it; and when I look into the roll that I carry in my bosom, that will do it; and when my thoughts wax warm about whither I am going, that will do it.

Prudence: And what is it that makes you so desirous to go to Mount Zion?

Christian: Why, there I hope to see Him alive that did hang dead on the cross; and there I hope to be rid of all those things that to this day are in me an annoyance to me: there they say there is no death, Isa. 25:8; Rev. 21:4; and there I shall dwell with such company as I like best. For, to tell you the truth, I love Him because I was by Him eased of my burden; and I am weary of my inward sickness. I would fain be where I shall die no more, and with the company that shall continually cry, Holy, holy, holy.

Then said Charity to Christian, Have you a family; Are you a married man?

Christian: I have a wife and four small children.

Charity: And why did you not bring them along with you?

Christian: Then Christian wept, and said, Oh, how willingly would I have done it! but they were all of them utterly averse to my going on pilgrimage.

Charity: But you should have talked to them, and have endeavored to show them the danger of staying behind.

Christian: So I did; and told them also what God had shown to me of the destruction of our city; but I seemed to them as one that mocked, and they believed me not. Gen. 19:14.

Charity: And did you pray to God that he would bless your counsel to them?

Christian: Yes, and that with much affection; for you must think that my wife and poor children were very dear to me.

Charity: But did you tell them of your own sorrow, and fear of destruction? for I suppose that destruction was visible enough to you.

Christian: Yes, over, and over, and over. They might also see my fears in my countenance, in my tears, and also in my trembling under the apprehension of the judgment that did hang over our heads; but all was not sufficient to prevail with them to come with me.

Charity: But what could they say for themselves, why they came not?

Christian: Why, my wife was afraid of losing this world, and my children were given to the foolish delights of youth; so, what by one thing, and what by another, they left me to wander in this manner alone.

Charity: But did you not, with your vain life, damp all that you, by words, used by way of persuasion to bring them away with you?

Christian: Indeed, I cannot commend my life, for I am conscious to myself of many failings therein. I know also, that a man, by his conversation, may soon overthrow what, by argument or persuasion, he doth labor to fasten upon others for their good. Yet this I can say, I was very wary of giving them occasion, by any unseemly action, to make them averse to going on pilgrimage. Yea, for this very thing, they would tell me I was too precise, and that I denied myself of things (for their sakes) in which they saw no evil. Nay, I think I may say, that if what they saw in me did hinder them, it was my great tenderness in sinning against God, or of doing any wrong to my neighbor.

Charity: Indeed, Cain hated his brother, because his own works were evil, and his brother's righteous, 1 John, 3:12; and if thy wife and children have been offended with thee for this, they thereby show themselves to be implacable to good; thou hast delivered thy soul from their blood. Ezek. 3:19.

Now I saw in my dream, that thus they sat talking together until supper was ready. So when they had made ready, they sat down to meat. Now the table was furnished with fat things, and with wine that was well refined; and all their talk at the table was about the Lord of the hill; as, namely, about what he had done, and wherefore he did what he did, and why he had builded that house; and by what they said, I perceived that he had been a great warrior, and had fought with and slain him that had the power of death, Heb. 2:14,15; but not without great danger to himself, which made me love him the more.

For, as they said, and as I believe, said Christian, he did it with the loss of much blood. But that which put the glory of grace into all he did, was, that he did it out of pure love to his country. And besides, there were some of them of the household that said they had been and spoke with him since he did die on the cross; and they have attested that they had it from his own lips, that he is such a lover of poor pilgrims, that the like is not to be found from the east to the west. They, moreover, gave an instance of what they affirmed; and that was, he had stripped himself of his glory that he might do this for the poor; and that they heard him say and affirm, that he would not dwell in the mountain of Zion alone. They said, moreover, that he had made many pilgrims princes, though by nature they were beggars born, and their original had been the dunghill. 1 Sam. 2:8; Psa. 113:7.

Thus they discoursed together till late at night; and after they had committed themselves to their Lord for protection, they betook themselves to rest. The pilgrim they laid in a large upper chamber, whose window opened towards the sun-rising. The name of the chamber was Peace, where he slept till break of day, and then he awoke and sang,
\begin{verse}
``Where am I now? Is this the love and care\\
Of Jesus, for the men that pilgrims are,\\
Thus to provide that I should be forgiven,\\
And dwell already the next door to heaven!"\\
\end{verse}
So in the morning they all got up; and, after some more discourse, they told him that he should not depart till they had shown him the rarities of that place. And first they had him into the study, where they showed him records of the greatest antiquity; in which, as I remember my dream, they showed him the pedigree of the Lord of the hill, that he was the Son of the Ancient of days, and came by eternal generation. Here also was more fully recorded the acts that he had done, and the names of many hundreds that he had taken into his service; and how he had placed them in such habitations that could neither by length of days, nor decays of nature, be dissolved.

Then they read to him some of the worthy acts that some of his servants had done; as how they had subdued kingdoms, wrought righteousness, obtained promises, stopped the mouths of lions, quenched the violence of fire, escaped the edge of the sword, out of weakness were made strong, waxed valiant in fight, and turned to flight the armies of the aliens. Heb. 11:33,34.

Then they read again another part of the records of the house, where it was shown how willing their Lord was to receive into his favor any, even any, though they in time past had offered great affronts to his person and proceedings. Here also were several other histories of many other famous things, of all which Christian had a view; as of things both ancient and modern, together with prophecies and predictions of things that have their certain accomplishment, both to the dread and amazement of enemies, and the comfort and solace of pilgrims.

The next day they took him, and had him into the armory, where they showed him all manner of furniture which their Lord had provided for pilgrims, as sword, shield, helmet, breastplate, all-prayer, and shoes that would not wear out. And there was here enough of this to harness out as many men for the service of their Lord as there be stars in the heaven for multitude.

They also showed him some of the engines with which some of his servants had done wonderful things. They showed him Moses' rod; the hammer and nail with which Jael slew Sisera; the pitchers, trumpets, and lamps too, with which Gideon put to flight the armies of Midian. Then they showed him the ox-goad wherewith Shamgar slew six hundred men. They showed him also the jawbone with which Samson did such mighty feats. They showed him moreover the sling and stone with which David slew Goliath of Gath; and the sword also with which their Lord will kill the man of sin, in the day that he shall rise up to the prey. They showed him besides many excellent things, with which Christian was much delighted. This done, they went to their rest again.

Then I saw in my dream, that on the morrow he got up to go forward, but they desired him to stay till the next day also; and then, said they, we will, if the day be clear, show you the Delectable Mountains; which, they said, would yet farther add to his comfort, because they were nearer the desired haven than the place where at present he was; so he consented and stayed. When the morning was up, they had him to the top of the house, and bid him look south. So he did, and behold, at a great distance, he saw a most pleasant mountainous country, beautified with woods, vineyards, fruits of all sorts, flowers also, with springs and fountains, very delectable to behold. Isa. 33:16,17. Then he asked the name of the country. They said it was Immanuel's land; and it is as common, said they, as this hill is, to and for all the pilgrims. And when thou comest there, from thence thou mayest see to the gate of the celestial city, as the shepherds that live there will make appear.

Now he bethought himself of setting forward, and they were willing he should. But first, said they, let us go again into the armory. So they did; and when he came there, they harnessed him from head to foot with what was of proof, lest perhaps he should meet with assaults in the way. He being therefore thus accoutred, walked out with his friends to the gate; and there he asked the Porter if he saw any pilgrim pass by. Then the Porter answered, Yes.

Christian: Pray, did you know him? said he.

The Porter: I asked his name, and he told me it was Faithful.

Christian: O, said Christian, I know him; he is my townsman, my near neighbor; he comes from the place where I was born. How far do you think he may be before?

The Porter: He is got by this time below the hill.

Christian: Well, said Christian, good Porter, the Lord be with thee, and add to all thy plain blessings much increase for the kindness that thou hast showed me. 
