\chapter[THE SECOND STAGE]{}

Then did Christian address himself to go back; and Evangelist, after he had kissed him, gave him one smile, and bid him God speed; So he went on with haste, neither spake he to any man by the way; nor if any asked him, would he vouchsafe them an answer. He went like one that was all the while treading on forbidden ground, and could by no means think himself safe, till again he was got into the way which he had left to follow Mr. Worldly Wiseman's counsel. So, in process of time, Christian got up to the gate. Now, over the gate there was written, ``Knock, and it shall be opened unto you." Matt. 7:7.

He knocked, therefore, more than once or twice, saying,
\begin{verse}
``May I now enter here? Will he within\\
Open to sorry me, though I have been\\
An undeserving rebel? Then shall I\\
Not fail to sing his lasting praise on high."\\
\end{verse}
At last there came a grave person to the gate, named Goodwill, who asked who was there, and whence he came, and what he would have.

Christian: Here is a poor burdened sinner. I come from the city of Destruction, but am going to Mount Zion, that I may be delivered from the wrath to come; I would therefore, sir, since I am informed that by this gate is the way thither, know if you are willing to let me in.

Goodwill: I am willing with all my heart, said he; and with that he opened the gate.

So when Christian was stepping in, the other gave him a pull. Then said Christian, What means that? The other told him, A little distance from this gate there is erected a strong castle, of which Beelzebub is the captain: from thence both he and they that are with him, shoot arrows at those that come up to this gate, if haply they may die before they can enter in. Then said Christian, I rejoice and tremble. So when he was got in, the man of the Gate asked him who directed him thither.

Christian: Evangelist bid me come hither and knock, as I did: and he said, that you, sir, would tell me what I must do.

Goodwill: An open door is set before thee, and no man can shut it.

Christian: Now I begin to reap the benefits of my hazards.

Goodwill: But how is it that you came alone?

Christian: Because none of my neighbors saw their danger as I saw mine.

Goodwill: Did any of them know of your coming?

Christian: Yes, my wife and children saw me at the first, and called after me to turn again: also, some of my neighbors stood crying and calling after me to return; but I put my fingers in my ears, and so came on my way.

Goodwill: But did none of them follow you, to persuade you to go back?

Christian: Yes, both Obstinate and Pliable; but when they saw that they could not prevail, Obstinate went railing back; but Pliable came with me a little way.

Goodwill: But why did he not come through?

Christian: We indeed came both together until we came to the Slough of Despond, into the which we also suddenly fell. And then was my neighbor Pliable discouraged, and would not venture farther. Wherefore, getting out again on the side next to his own house, he told me I should possess the brave country alone for him: so he went his way, and I came mine; he after Obstinate, and I to this gate.

Goodwill: Then said Goodwill, Alas, poor man; is the celestial glory of so little esteem with him, that he counteth it not worth running the hazard of a few difficulties to obtain it?

Christian: Truly, said Christian, I have said the truth of Pliable; and if I should also say all the truth of myself, it will appear there is no betterment betwixt him and myself. It is true, he went back to his own house, but I also turned aside to go in the way of death, being persuaded thereto by the carnal arguments of one Mr. Worldly Wiseman.

Goodwill: Oh, did he light upon you? What, he would have had you seek for ease at the hands of Mr. Legality! They are both of them a very cheat. But did you take his counsel?

Christian: Yes, as far as I durst. I went to find out Mr. Legality, until I thought that the mountain that stands by his house would have fallen upon my head; wherefore there I was forced to stop.

Goodwill: That mountain has been the death of many, and will be the death of many more: it is well you escaped being by it dashed in pieces.

Christian: Why truly I do not know what had become of me there, had not Evangelist happily met me again as I was musing in the midst of my dumps; but it was God's mercy that he came to me again, for else I had never come hither. But now I am come, such a one as I am, more fit indeed for death by that mountain, than thus to stand talking with my Lord. But O, what a favor is this to me, that yet I am admitted entrance here!

Goodwill: We make no objections against any, notwithstanding all that they have done before they come hither; they in no wise are cast out. John 6:37. And therefore good Christian, come a little way with me, and I will teach thee about the way thou must go. Look before thee; dost thou see this narrow way? That is the way thou must go. It was cast up by the patriarchs, prophets, Christ, and his apostles, and it is as strait as a rule can make it; this is the way thou must go.

Christian: But, said Christian, are there no turnings nor windings, by which a stranger may lose his way?

Goodwill: Yes, there are many ways butt down upon this, and they are crooked and wide: but thus thou mayest distinguish the right from the wrong, the right only being strait and narrow. Matt. 7:14.

Then I saw in my dream, that Christian asked him further, if he could not help him off with his burden that was upon his back. For as yet he had not got rid thereof; nor could he by any means get it off without help.

He told him, ``As to thy burden, be content to bear it until thou comest to the place of deliverance; for there it will fall from thy back of itself."

Then Christian began to gird up his loins, and to address himself to his journey. So the other told him, that by that he was gone some distance from the gate, he would come to the house of the Interpreter, at whose door he should knock, and he would show him excellent things. Then Christian took his leave of his friend, and he again bid him God speed.

Then he went on till he came at the house of the Interpreter, [6] where he knocked over and over. At last one came to the door, and asked who was there.

Christian: Sir, here is a traveller, who was bid by an acquaintance of the good man of this house to call here for my profit; I would therefore speak with the master of the house.

So he called for the master of the house, who, after a little time, came to Christian, and asked him what he would have.

Christian: Sir, said Christian, I am a man that am come from the city of Destruction, and am going to the Mount Zion; and I was told by the man that stands at the gate at the head of this way, that if I called here you would show me excellent things, such as would be helpful to me on my journey.

Interpreter: Then said Interpreter, Come in; I will show thee that which will be profitable to thee. So he commanded his man to light the candle, and bid Christian follow him; so he had him into a private room, and bid his man open a door; the which when he had done, Christian saw the picture a very grave person hang up against the wall; and this was the fashion of it: It had eyes lifted up to heaven, the best of books in his hand, the law of truth was written upon its lips, the world was behind its back; it stood as if it pleaded with men, and a crown of gold did hang over its head.

Christian: Then said Christian, What means this?

Interpreter: The man whose picture this is, is one of a thousand: he can beget children, 1 Cor. 4:15, travail in birth with children, Gal. 4:19, and nurse them himself when they are born. And whereas thou seest him with his eyes lift up to heaven, the best of books in his hand, and the law of truth writ on his lips: it is to show thee, that his work is to know, and unfold dark things to sinners; even as also thou seest him stand as if he pleaded with men. And whereas thou seest the world as cast behind him, and that a crown hangs over his head; that is to show thee, that slighting and despising the things that are present, for the love that he hath to his Master's service, he is sure in the world that comes next, to have glory for his reward. Now, said the Interpreter, I have showed thee this picture first, because the man whose picture this is, is the only man whom the Lord of the place whither thou art going hath authorized to be thy guide in all difficult places thou mayest meet with in the way: wherefore take good heed to what I have showed thee, and bear well in thy mind what thou hast seen, lest in thy journey thou meet with some that pretend to lead thee right, but their way goes down to death.

Then he took him by the hand, and led him into a very large parlor that was full of dust, because never swept; the which after he had reviewed it a little while, the Interpreter called for a man to sweep. Now, when he began to sweep, the dust began so abundantly to fly about, that Christian had almost therewith been choked. Then said the Interpreter to a damsel that stood by, ``Bring hither water, and sprinkle the room;" the which when she had done, it was swept and cleansed with pleasure.

Christian: Then said Christian, What means this?

Interpreter: The Interpreter answered, This parlor is the heart of a man that was never sanctified by the sweet grace of the Gospel. The dust is his original sin, and inward corruptions, that have defiled the whole man. He that began to sweep at first, is the law; but she that brought water, and did sprinkle it, is the Gospel. Now whereas thou sawest, that so soon as the first began to sweep, the dust did so fly about that the room by him could not be cleansed, but that thou wast almost choked therewith; this is to show thee, that the law, instead of cleansing the heart (by its working) from sin, doth revive, Rom. 7:9, put strength into, 1 Cor. 15:56, and increase it in the soul, Rom. 5:20, even as it doth discover and forbid it; for it doth not give power to subdue. Again, as thou sawest the damsel sprinkle the room with water, upon which it was cleansed with pleasure, this is to show thee, that when the Gospel comes in the sweet and precious influences thereof to the heart, then, I say, even as thou sawest the damsel lay the dust by sprinkling the floor with water, so is sin vanquished and subdued, and the soul made clean, through the faith of it, and consequently fit for the King of glory to inhabit. John 15:3; Eph. 5:26; Acts 15:9; Rom. 16:25,26.

I saw moreover in my dream, that the Interpreter took him by the hand, and had him into a little room, where sat two little children, each one in his chair. The name of the eldest was Passion, and the name of the other Patience. Passion seemed to be much disconted, but Patience was very quiet. Then Christian asked, ``What is the reason of the discontent of Passion?" The Interpreter answered, ``The governor of them would have him stay for his best things till the beginning of the next year, but he will have all now; but Patience is willing to wait."

Then I saw that one came to Passion, and brought him a bag of treasure, and poured it down at his feet: the which he took up, and rejoiced therein, and withal laughed Patience to scorn. But I beheld but a while, and he had lavished all away, and had nothing left him but rags.

Christian: Then said Christian to the Interpreter, Expound this matter more fully to me.

Interpreter: So he said, These two lads are figures; Passion of the men of this world, and Patience of the men of that which is to come; for, as here thou seest, passion will have all now, this year, that is to say, in this world; so are the men of this world: They must have all their good things now; they cannot stay till the next year, that is, until the next world, for their portion of good. That proverb, ``A bird in the hand is worth two in the bush," is of more authority with them than are all the divine testimonies of the good of the world to come. But as thou sawest that he had quickly lavished all away, and had presently left him nothing but rags, so will it be with all such men at the end of this world.

Christian: Then said Christian, Now I see that Patience has the best wisdom, and that upon many accounts. 1. Because he stays for the best things. 2. And also because he will have the glory of his, when the other has nothing but rags.

Interpreter: Nay, you may add another, to wit, the glory of the next world will never wear out; but these are suddenly gone. Therefore Passion had not so much reason to laugh at Patience because he had his good things first, as Patience will have to laugh at Passion because he had his best things last; for first must give place to last, because last must have his time to come: but last gives place to nothing, for there is not another to succeed. He, therefore, that hath his portion first, must needs have a time to spend it; but he that hath his portion last, must have it lastingly: therefore it is said of Dives, ``In thy lifetime thou receivedst thy good things, and likewise Lazarus evil things; but now he is comforted, and thou art tormented." Luke 16:25.

Christian: Then I perceive it is not best to covet things that are now, but to wait for things to come.

Interpreter: You say truth: for the things that are seen are temporal, but the things that are not seen are eternal. 2 Cor. 4:18. But though this be so, yet since things present and our fleshly appetite are such near neighbors one to another; and again, because things to come and carnal sense are such strangers one to another; therefore it is, that the first of these so suddenly fall into amity, and that distance is so continued between the second.

Then I saw in my dream, that the Interpreter took Christian by the hand, and led him into a place where was a fire burning against a wall, and one standing by it, always casting much water upon it, to quench it; yet did the fire burn higher and hotter.

Then said Christian, What means this?

The Interpreter answered, This fire is the work of grace that is wrought in the heart; he that casts water upon it, to extinguish and put it out, is the devil: but in that thou seest the fire, notwithstanding, burn higher and hotter, thou shalt also see the reason of that. So he had him about to the back side of the wall, where he saw a man with a vessel of oil in his hand, of the which he did also continually cast (but secretly) into the fire.

Then said Christian, What means this?

The Interpreter answered, This is Christ, who continually, with the oil of his grace, maintains the work already begun in the heart; by the means of which, notwithstanding what the devil can do, the souls of his people prove gracious still. 2 Cor. 12:9. And in that thou sawest that the man stood behind the wall to maintain the fire; this is to teach thee, that it is hard for the tempted to see how this work of grace is maintained in the soul.

I saw also, that the Interpreter took him again by the hand, and led him into a pleasant place, where was built a stately palace, beautiful to behold; at the sight of which Christian was greatly delighted. He saw also upon the top thereof certain persons walking, who were clothed all in gold.

Then said Christian may we go in thither?

Then the Interpreter took him, and led him up towards the door of the palace; and behold, at the door stood a great company of men, as desirous to go in, but durst not. There also sat a man at a little distance from the door, at a table-side, with a book and his inkhorn before him, to take the names of them that should enter therein; he saw also that in the doorway stood many men in armor to keep it, being resolved to do to the men that would enter, what hurt and mischief they could. Now was Christian somewhat in amaze. At last, when every man started back for fear of the armed men, Christian saw a man of a very stout countenance come up to the man that sat there to write, saying, ``Set down my name, sir;" the which when he had done, he saw the man draw his sword, and put a helmet on his head, and rush towards the door upon the armed men, who laid upon him with deadly force; but the man, not at all discouraged, fell to cutting and hacking most fiercely. So after he had received and given many wounds to those that attempted to keep him out, Matt. 11:12; Acts 14:22; he cut his way through them all, and pressed forward into the palace; at which there was a pleasant voice heard from those that were within, even of those that walked upon the top of the palace, saying,

 ``Come in, come in,

Eternal glory thou shalt win."

So he went in, and was clothed with such garments as they. Then Christian smiled, and said, I think verily I know the meaning of this.

Now, said Christian, let me go hence. Nay, stay, said the Interpreter, till I have showed thee a little more, and after that thou shalt go on thy way. So he took him by the hand again, and led him into a very dark room, where there sat a man in an iron cage.

Now the man, to look on, seemed very sad; he sat with his eyes looking down to the ground, his hands folded together, and he sighed as if he would break his heart. Then said Christian, What means this? At which the Interpreter bid him talk with the man.

Then said Christian to the man, What art thou? The man answered, I am what I was not once.

Christian: What wast thou once?

The Man: The man said, I was once a fair and flourishing professor, Luke 8:13, both in mine own eyes, and also in the eyes of others: I once was, as I thought, fair for the celestial city, and had then even joy at the thoughts that I should get thither.

Christian: Well, but what art thou now?

The Man: I am now a man of despair, and am shut up in it, as in this iron cage. I cannot get out; Oh now I cannot!

Christian: But how camest thou into this condition?

The Man: I left off to watch and be sober: I laid the reins upon the neck of my lusts; I sinned against the light of the word, and the goodness of God; I have grieved the Spirit, and he is gone; I tempted the devil, and he is come to me; I have provoked God to anger, and he has left me: I have so hardened my heart, that I cannot repent.

Then said Christian to the Interpreter, But is there no hope for such a man as this? Ask him, said the Interpreter.

Christian: Then said Christian, Is there no hope, but you must be kept in the iron cage of despair?

The Man: No, none at all.

Christian: Why, the Son of the Blessed is very pitiful.

The Man: I have crucified him to myself afresh, Heb. 6:6; I have despised his person, Luke 19:14; I have despised his righteousness; I have counted his blood an unholy thing; I have done despite to the spirit of grace, Heb. 10:29: therefore I have shut myself out of all the promises and there now remains to me nothing but threatenings, dreadful threatenings, faithful threatenings of certain judgment and fiery indignation, which shall devour me as an adversary.

Christian: For what did you bring yourself into this condition?

The Man: For the lusts, pleasures, and profits of this world; in the enjoyment of which I did then promise myself much delight: but now every one of those things also bite me, and gnaw me like a burning worm.

Christian: But canst thou not now repent and turn?

The Man: God hath denied me repentance. His word gives me no encouragement to believe; yea, himself hath shut me up in this iron cage: nor can all the men in the world let me out. Oh eternity! eternity! how shall I grapple with the misery that I must meet with in eternity?

Interpreter: Then said the Interpreter to Christian, Let this man's misery be remembered by thee, and be an everlasting caution to thee.

Christian: Well, said Christian, this is fearful! God help me to watch and to be sober, and to pray that I may shun the cause of this man's misery. Sir, is it not time for me to go on my way now?

Interpreter: Tarry till I shall show thee one thing more, and then thou shalt go on thy way.

So he took Christian by the hand again and led him into a chamber where there was one rising out of bed; and as he put on his raiment, he shook and trembled. Then said Christian, Why doth this man thus tremble? The Interpreter then bid him tell to Christian the reason of his so doing.

So he began, and said, ``This night, as I was in my sleep, I dreamed, and behold the heavens grew exceeding black; also it thundered and lightened in most fearful wise, that it put me into an agony. So I looked up in my dream, and saw the clouds rack at an unusual rate; upon which I heard a great sound of a trumpet, and saw also a man sitting upon a cloud, attended with the thousands of heaven: they were all in flaming fire; also the heavens were in a burning flame. I heard then a voice, saying, Arise, ye dead, and come to judgment.' And with that the rocks rent, the graves opened, and the dead that were therein came forth: some of them were exceeding glad, and looked upward; and some sought to hide themselves under the mountains. Then I saw the man that sat upon the cloud open the book, and bid the world draw near. Yet there was, by reason of a fierce flame that issued out and came from before him, a convenient distance between him and them, as between the judge and the prisoners at the bar. 1 Cor. 15; 1 Thess. 4:16; Jude 15; John 5: 28,29; 2 Thess. 1:8-10; Rev. 20:11-14; Isa. 26:21; Micah 7:16,17; Psa. 5:4; 50:1-3; Mal. 3:2,3; Dan. 7:9,10. I heard it also proclaimed to them that attended on the man that sat on the cloud, Gather together the tares, the chaff, and stubble, and cast them into the burning lake.' Matt. 3:12; 18:30; 24:30; Mal. 4:1. And with that the bottomless pit opened, just whereabout I stood; out of the mouth of which there came, in an abundant manner, smoke, and coals of fire, with hideous noises. It was also said to the same persons, Gather my wheat into the garner.' Luke 3:17. And with that I saw many catched up and carried away into the clouds, but I was left behind. 1 Thess. 4:16,17. I also sought to hide myself, but I could not, for the man that sat upon the cloud still kept his eye upon me; my sins also came into my mind, and my conscience did accuse me on every side. Rom. 2:14,15. Upon this I awakened from my sleep."

Christian: But what was it that made you so afraid of this sight?

The Man: Why, I thought that the day of judgment was come, and that I was not ready for it: but this frightened me most, that the angels gathered up several, and left me behind; also the pit of hell opened her mouth just where I stood. My conscience too afflicted me; and, as I thought, the Judge had always his eye upon me, showing indignation in his countenance.

Then said the Interpreter to Christian, ``Hast thou considered all these things?"

Christian: Yes, and they put me in hope and fear.

Interpreter: Well, keep all things so in thy mind, that they may be as a goad in thy sides, to prick thee forward in the way thou must go. Then Christian began to gird up his loins, and to address himself to his journey. Then said the Interpreter, ``The Comforter be always with thee, good Christian, to guide thee in the way that leads to the city." So Christian went on his way, saying,

\begin{verse}
``Here I have seen things rare and profitable,\\
Things pleasant, dreadful, things to make me stable\\
In what I have begun to take in hand:\\
Then let me think on them, and understand\\
Wherefore they showed me were, and let me be\\
Thankful, O good Interpreter, to thee."\\
\end{verse}