\chapter[THE THIRD STAGE]{}

The Interpreter then called for a man-servant of his, one Great-heart, and bid him take A sword, and helmet, and shield; and, Take these my daughters, said he, conduct them to the house called Beautiful, at which place they will rest next. So he took his weapons, and went before them; and the Interpreter said, God speed. Those also that belonged to the family, sent them away with many a good wish. So they went on their way, and sang,
\begin{verse}
This place hath been our second stage:\\
Here we have heard, and seen\\
Those good things, that from age to age\\
To others hid have been.\\
\end{verse}
\begin{verse}
The dunghill-raker, spider, hen,\\
The chicken, too, to me\\
Have taught a lesson: let me then\\
Conformed to it be.\\
\end{verse}
\begin{verse}
The butcher, garden, and the field,\\
The robin and his bait,\\
Also the rotten tree, doth yield\\
Me argument of weight,\\
\end{verse}
\begin{verse}
To move me for to watch and pray,\\
To strive to be sincere;\\
To take my cross up day by day,\\
And serve the Lord with fear.\\
\end{verse}
Now I saw in my dream, that they went on, and Great-Heart before them. So they went, and came to the place where Christian's burden fell off his back and tumbled into a sepulchre. Here then they made a pause; and here also they blessed God. Now, said Christiana, it comes to my mind what was said to us at the gate, to wit, that we should have pardon by word and deed: by word, that is, by the promise; by deed, to wit, in the way it was obtained. What the promise is, of that I know something; but what is it to have pardon by deed, or in the way that it was obtained, Mr. Great-Heart, I suppose you know; wherefore, if you please, let us hear your discourse thereof.

Mr. Great-Heart: Pardon by the deed done, is pardon obtained by some one for another that hath need thereof; not by the person pardoned, but in the way, saith another, in which I have obtained it. So then, to speak to the question more at large, the pardon that you, and Mercy, and these boys have attained, was obtained by another; to wit, by him that let you in at the gate. And he hath obtained it in this double way; he hath performed righteousness to cover you, and spilt his blood to wash you in.

Christiana: But if he parts with his righteousness to us, what will he have for himself?

Mr. Great-Heart: He has more righteousness than you have need of, or than he needeth himself.

Christiana: Pray make that appear.

Mr. Great-Heart: With all my heart: but first I must premise, that he of whom we are now about to speak, is one that has not his fellow: He has two natures in one person, plain to be distinguished, impossible to be divided. Unto each of these natures a righteousness belongeth, and each righteousness is essential to that nature; so that one may as easily cause that nature to be extinct, as to separate its justice or righteousness from it. Of these righteousnesses therefore, we are not made partakers, so as that they, or any of them, should be put upon us, that we might be made just, and live thereby. Besides these, there is a righteousness which this person has, as these two natures are joined in one. And this is not the righteousness of the Godhead, as distinguished from the manhood; nor the righteousness of the manhood, as distinguished from the Godhead; but a righteousness which standeth in the union of both natures, and may properly be called the righteousness that is essential to his being prepared of God to the capacity of the mediatory office, which he was to be entrusted with. If he parts with his first righteousness, he parts with his Godhead; if he parts with his second righteousness, he parts with the purity of his manhood; if he parts with his third, he parts with that perfection that capacitates him to the office of mediation. He has therefore another righteousness, which standeth in performance, or obedience to a revealed will; and that is what he puts upon sinners, and that by which their sins are covered. Wherefore he saith, ``As by one man's disobedience many were made sinners, so by the obedience of one shall many be made righteous." Rom. 5:19.

Christiana: But are the other righteousnesses of no use to us?

Mr. Great-Heart: Yes; for though they are essential to his natures and office, and cannot be communicated unto another, yet it is by virtue of them that the righteousness that justifies is for that purpose efficacious. The righteousness of his Godhead gives virtue to his obedience; the righteousness of his manhood giveth capability to his obedience to justify; and the righteousness that standeth in the union of these two natures to his office, giveth authority to that righteousness to do the work for which it was ordained.

So then here is a righteousness that Christ, as God, has no need of; for he is God without it: Here is a righteousness that Christ, as man, has no need of to make him so; for he is perfect man without it. Again, here is a righteousness that Christ, as God-man, has no need of; for he is perfectly so without it. Here then is a righteousness that Christ, as God, and as God-man, has no need of, with reference to himself, and therefore he can spare it; a justifying righteousness, that he for himself wanteth not, and therefore giveth it away: Hence it is called the gift of righteousness. This righteousness, since Christ Jesus the Lord has made himself under the law, must be given away; for the law doth not only bind him that is under it, to do justly, but to use charity. Rom. 5:17. Wherefore he must, or ought by the law, if he hath two coats, to give one to him that hath none. Now, our Lord indeed hath two coats, one for himself, and one to spare; wherefore he freely bestows one upon those that have none. And thus, Christiana and Mercy, and the rest of you that are here, doth your pardon come by deed, or by the work of another man. Your Lord Christ is he that worked, and hath given away what he wrought for, to the next poor beggar he meets.

But again, in order to pardon by deed, there must something be paid to God as a price, as well as something prepared to cover us withal. Sin has delivered us up to the just curse of a righteous law: now from this curse we must be justified by way of redemption, a price being paid for the harms we have done; and this is by the blood of your Lord, who came and stood in your place and stead, and died your death for your transgressions: Thus has he ransomed you from your transgressions by blood, and covered your polluted and deformed souls with righteousness, Rom. 8:34; for the sake of which, God passeth by you and will not hurt you when he comes to judge the world. Gal. 3:13.

Christiana: This is brave! Now I see that there was something to be learned by our being pardoned by word and deed. Good Mercy, let us labor to keep this in mind: and, my children, do you remember it also. But, sir, was not this it that made my good Christian's burden fall from off his shoulder, and that made him give three leaps for joy?

Mr. Great-Heart: Yes, it was the belief of this that cut those strings that could not be cut by other means; and it was to give him a proof of the virtue of this, that he was suffered to carry his burden to the cross.

Christiana: I thought so; for though my heart was lightsome and joyous before, yet it is ten times more lightsome and joyous now. And I am persuaded by what I have felt, though I have felt but little as yet, that if the most burdened man in the world was here, and did see and believe as I now do, it would make his heart the more merry and blithe.

Mr. Great-Heart: There is not only comfort and the ease of a burden brought to us by the sight and consideration of these, but an endeared affection begot in us by it: for who can, if he doth but once think that pardon comes not only by promise but thus, but be affected with the way and means of his redemption, and so with the man that hath wrought it for him?

Christiana: True; methinks it makes my heart bleed to think that he should bleed for me. Oh, thou loving One: Oh, thou blessed One. Thou deservest to have me; thou hast bought me. Thou deservest to have me all: thou hast paid for me ten thousand times more than I am worth. No marvel that this made the tears stand in my husband's eyes, and that it made him trudge so nimbly on. I am persuaded he wished me with him: but, vile wretch that I was, I let him come all alone. Oh, Mercy, that thy father and mother were here; yea, and Mrs. Timorous also: nay, I wish now with all my heart that here was Madam Wanton too. Surely, surely, their hearts would be affected; nor could the fear of the one, nor the powerful lusts of the other, prevail with them to go home again, and to refuse to become good pilgrims.

Mr. Great-Heart: You speak now in the warmth of your affections; will it, think you, be always thus with you? Besides, this is not communicated to every one, nor to every one that did see your Jesus bleed. There were that stood by, and that saw the blood run from the heart to the ground, and yet were so far off this, that instead of lamenting, they laughed at him, and, instead of becoming his disciples, did harden their hearts against him. So that all that you have, my daughters, you have by peculiar impression made by a divine contemplating upon what I have spoken to you. Remember, that twas told you, that the hen, by her common call, gives no meat to her chickens. This you have therefore by a special grace.

Now I saw in my dream, that they went on until they were come to the place that Simple, and Sloth, and Presumption, lay and slept in when Christian went by on pilgrimage: and behold, they were hanged up in irons a little way off on the other side.

Mercy: Then said Mercy to him that was their guide and conductor, what are these three men; and for what are they hanged there?

Mr. Great-Heart: These three men were men of very bad qualities; they had no mind to be pilgrims themselves, and whomsoever they could, they hindered. They were sloth and folly themselves, and whomsoever they could persuade they made so too, and withal taught them to presume that they should do well at last. They were asleep when Christian went by; and now you go by, they are hanged.

Mercy: But could they persuade any to be of their opinion?

Mr. Great-Heart: Yes, they turned several out of the way. There was Slow-pace that they persuaded to do as they. They also prevailed with one Short-wind, with one No-heart, with one Linger-after-Lust, and with one Sleepy-head, and with a young woman, her name was Dull, to turn out of the way and become as they. Besides, they brought up an ill report of your Lord, persuading others that he was a hard taskmaster. They also brought up an evil report of the good Land, saying, it was not half so good as some pretended it was. They also began to vilify his servants, and to count the best of them meddlesome, troublesome busybodies. Further, they would call the bread of God husks; the comforts of his children, fancies; the travel and labor of pilgrims, things to no purpose.

Christiana: Nay, said Christiana, if they were such, they shall never be bewailed by me: they have but what they deserve; and I think it is well that they stand so near the highway, that others may see and take warning. But had it not been well if their crimes had been engraven in some plate of iron or brass, and left here where they did their mischiefs, for a caution to other bad men?

Mr. Great-Heart: So it is, as you may well perceive, if you will go a little to the wall.

Mercy: No, no; let them hang, and their names rot, and their crimes live forever against them. I think it a high favor that they were hanged before we came hither: who knows else what they might have done to such poor women as we are? Then she turned it into a song, saying,
\begin{verse}
``Now then you three hang there, and be a sign\\
To all that shall against the truth combine.\\
And let him that comes after, fear this end,\\
If unto pilgrims he is not a friend.\\
And thou, my soul, of all such men beware,\\
That unto holiness opposers are."\\
\end{verse}

Thus they went on till they came to the foot of the hill Difficulty, where again the good Mr. Great-Heart took an occasion to tell them what happened there when Christian himself went by. So he had them first to the spring. Lo, saith he, this is the spring that Christian drank of before he went up this hill: and then it was clear and good; but now it is dirty with the feet of some that are not desirous that pilgrims here should quench their thirst. Ezek. 34:18,19. Thereat Mercy said, And why so envious, trow? But, said their guide, it will do, if taken up and put into a vessel that is sweet and good; for then the dirt will sink to the bottom, and the water come out by itself more clear. Thus therefore Christiana and her companions were compelled to do. They took it up, and put it into an earthen pot, and so let it stand till the dirt was gone to the bottom, and then they drank thereof.

Next he showed them the two by-ways that were at the foot of the hill, where Formality and Hypocrisy lost themselves. And, said he, these are dangerous paths. Two were here cast away when Christian came by; and although, as you see these ways are since stopped up with chains, posts, and a ditch, yet there are those that will choose to adventure here rather than take the pains to go up this hill.

Christiana: ``The way of transgressors is hard." Prov. 13:15. It is a wonder that they can get into these ways without danger of breaking their necks.

Mr. Great-Heart: They will venture: yea, if at any time any of the King's servants do happen to see them, and do call upon them, and tell them that they are in the wrong way, and do bid them beware of the danger, then they railingly return them answer, and say, ``As for the word that thou hast spoken unto us in the name of the King, we will not hearken unto thee; but we will certainly do whatsoever thing goeth out of our own mouths." Jer. 44:16,17. Nay, if you look a little further, you shall see that these ways are made cautionary enough, not only by these posts, and ditch, and chain, but also by being hedged up: yet they will choose to go there.

Christiana: They are idle; they love not to take pains; up-hill way is unpleasant to them. So it is fulfilled unto them as it is written, ``The way of the slothful man is full of thorns." Prov. 15:19. Yea, they will rather choose to walk upon a snare than to go up this hill, and the rest of this way to the city.

Then they set forward, and began to go up the hill, and up the hill they went. But before they got to the top, Christiana began to pant, and said, I dare say this is a breathing hill; no marvel if they that love their ease more than their souls choose to themselves a smoother way.

Then said Mercy, I must sit down: also the least of the children began to cry. Come, come, said Great-Heart, sit not down here; for a little above is the Prince's arbor. Then he took the little boy by the hand, and led him up thereto.

When they were come to the arbor, they were very willing to sit down, for they were all in a pelting heat. Then said Mercy, ``How sweet is rest to them that labor." Matt. 11:28; and how good is the Prince of pilgrims to provide such resting-places for them! Of this arbor I have heard much; but I never saw it before. But here let us beware of sleeping; for, as I have heard, it cost poor Christian dear.

Then said Mr. Great-Heart to the little ones, Come, my pretty boys, how do you do? What think you now of going on pilgrimage? Sir, said the least, I was almost beat out of heart; but I thank you for lending me a hand at my need. And I remember now what my mother hath told me, namely, that the way to heaven is as a ladder, and the way to hell is as down a hill. But I had rather go up the ladder to life, than down the hill to death.

Then said Mercy, But the proverb is, ``To go down the hill is easy." But James said, (for that was his name,) The day is coming when, in my opinion, when going down the hill will be the hardest of all. Tis a good boy, said his master; thou hast given her a right answer. Then Mercy smiled, but the little boy did blush.

Christiana: Come, said Christiana, will you eat a bit to sweeten your mouths, while you sit here to rest your legs? for I have here a piece of pomegranate which Mr. Interpreter put into my hand just when I came out of his door; he gave me also a piece of an honeycomb, and a little bottle of spirits. I thought he gave you something, said Mercy, because he called you aside. Yes, so he did, said the other; but, said Christiana, it shall be still as I said it should, when at first we came from home; thou shalt be a sharer in all the good that I have, because thou so willingly didst become my companion. Then she gave to them, and they did eat, both Mercy and the boys. And said Christiana to Mr. Great-Heart, Sir, will you do as we? But he answered, You are going on pilgrimage, and presently I shall return; much good may what you have do you: at home I eat the same every day. 
