\chapter[THE EIGHTH STAGE]{}

They went then till they came to the Delectable Mountains, which mountains belong to the Lord of that hill of which we have spoken before. So they went up to the mountains, to behold the gardens and orchards, the vineyards and fountains of water; where also they drank and washed themselves, and did freely eat of the vineyards. Now, there were on the tops of these mountains shepherds feeding their flocks, and they stood by the highway-side. The pilgrims, therefore, went to them, and leaning upon their staffs, (as is common with weary pilgrims when they stand to talk with any by the way,) they asked, Whose Delectable Mountains are these; and whose be the sheep that feed upon them?

The Shepherds: These mountains are Emmanuel's land, and they are within sight of his city; and the sheep also are his, and he laid down his life for them. John 10:11,15.

Christian: Is this the way to the Celestial City?

The Shepherds: You are just in your way.

Christian: How far is it thither?

The Shepherds: Too far for any but those who shall get thither indeed.

Christian: Is the way safe or dangerous?

The Shepherds: Safe for those for whom it is to be safe; but transgressors shall fall therein. Hos. 14:9.

Christian: Is there in this place any relief for pilgrims that are weary and faint in the way?

The Shepherds: The Lord of these mountains hath given us a charge not to be forgetful to entertain strangers, Heb. 13:2; therefore the good of the place is before you .

I saw also in my dream, that when the shepherds perceived that they were wayfaring men, they also put questions to them, (to which they made answer as in other places,) as, Whence came you? and, How got you into the way? and, By what means have you so persevered therein? for but few of them that begin to come hither, do show their face on these mountains. But when the shepherds heard their answers, being pleased therewith, they looked very lovingly upon them, and said, Welcome to the Delectable Mountains.

The shepherds, I say, whose names were Knowledge, Experience, Watchful, and Sincere, took them by the hand, and had them to their tents, and made them partake of that which was ready at present. They said moreover, We would that you should stay here a while, to be acquainted with us, and yet more to solace yourselves with the good of these Delectable Mountains. Then they told them that they were content to stay. So they went to their rest that night, because it was very late.

Then I saw in my dream, that in the morning the shepherds called up Christian and Hopeful to walk with them upon the mountains. So they went forth with them, and walked a while, having a pleasant prospect on every side. Then said the shepherds one to another, Shall we show these pilgrims some wonders? So when they had concluded to do it, they had them first to the top of a hill called Error, which was very steep on the farthest side, and bid them look down to the bottom. So Christian and Hopeful looked down, and saw at the bottom several men dashed all to pieces by a fall that they had had from the top. Then said Christian, What meaneth this? The shepherds answered, Have you not heard of them that were made to err, by hearkening to Hymenius and Philetus, as concerning the faith of the resurrection of the body? 2 Tim. 2:17,18. They answered, Yes. Then said the shepherds, Those that you see lie dashed in pieces at the bottom of this mountain are they; and they have continued to this day unburied, as you see, for an example to others to take heed how they clamber too high, or how they come too near the brink of this mountain.

Then I saw that they had them to the top of another mountain, and the name of that is Caution, and bid them look afar off; which, when they did, they perceived, as they thought, several men walking up and down among the tombs that were there; and they perceived that the men were blind, because they stumbled sometimes upon the tombs, and because they could not get out from among them. Then said Christian, What means this?

The shepherds then answered, Did you not see, a little below these mountains, a stile that led into a meadow, on the left hand of this way? They answered, Yes. Then said the shepherds, From that stile there goes a path that leads directly to Doubting Castle, which is kept by Giant Despair; and these men (pointing to them among the tombs) came once on pilgrimage, as you do now, even until they came to that same stile. And because the right way was rough in that place, they chose to go out of it into that meadow, and there were taken by Giant Despair, and cast into Doubting Castle; where after they had a while been kept in the dungeon, he at last did put out their eyes, and led them among those tombs, where he has left them to wander to this very day, that the saying of the wise man might be fulfilled, ``He that wandereth out of the way of understanding shall remain in the congregation of the dead." Prov. 21:16. Then Christian and Hopeful looked upon one another, with tears gushing out, but yet said nothing to the shepherds.

Then I saw in my dream, that the shepherds had them to another place in a bottom, where was a door on the side of a hill; and they opened the door, and bid them look in. They looked in, therefore, and saw that within it was very dark and smoky; they also thought that they heard there a rumbling noise, as of fire, and a cry of some tormented, and that they smelt the scent of brimstone. Then said Christian, What means this? The shepherds told them, This is a by-way to hell, a way that hypocrites go in at; namely, such as sell their birthright, with Esau; such as sell their Master, with Judas; such as blaspheme the Gospel, with Alexander; and that lie and dissemble, with Ananias and Sapphira his wife.

Then said Hopeful to the shepherds, I perceive that these had on them, even every one, a show of pilgrimage, as we have now; had they not?

The Shepherds: Yes, and held it a long time, too.

Hopeful: How far might they go on in pilgrimage in their day, since they, notwithstanding, were miserably cast away?

The Shepherds: Some farther, and some not so far as these mountains.

Then said the pilgrims one to the other, We had need to cry to the Strong for strength.

The Shepherds: Aye, and you will have need to use it, when you have it, too.

By this time the pilgrims had a desire to go forward, and the shepherds a desire they should; so they walked together towards the end of the mountains. Then said the shepherds one to another, Let us here show the pilgrims the gates of the Celestial City, if they have skill to look through our perspective glass. The pilgrims lovingly accepted the motion: so they had them to the top of a high hill, called Clear, and gave them the glass to look.

Then they tried to look; but the remembrance of that last thing that the shepherds had shown them made their hands shake, by means of which impediment they could not look steadily through the glass; yet they thought they saw something like the gate, and also some of the glory of the place. Then they went away, and sang,
\begin{verse}
 ``Thus by the shepherds secrets are reveal'd,\\
Which from all other men are kept concealed:\\
Come to the shepherds then, if you would see\\
Things deep, things hid, and that mysterious be."\\
\end{verse}
When they were about to depart, one of the shepherds gave them a note of the way. Another of them bid them beware of the Flatterer. The third bid them take heed that they slept not upon Enchanted Ground. And the fourth bid them God speed. So I awoke from my dream. 
