\chapter[THE FIRST STAGE]{}

By this time Christiana was got on her way, and Mercy went along with her: so as they went, her children being there also, Christiana began to discourse. And, Mercy, said Christiana, I take this as an unexpected favor, that thou shouldest set forth out of doors with me to accompany me a little in the way.

Mercy: Then said young Mercy, (for she was but young,) If I thought it would be to purpose to go with you, I would never go near the town any more.

Christiana: Well, Mercy, said Christiana, cast in thy lot with me: I well know what will be the end of our pilgrimage: my husband is where he would not but be for all the gold in the Spanish mines. Nor shalt thou be rejected, though thou goest but upon my invitation. The King, who hath sent for me and my children, is one that delighteth in mercy. Besides, if thou wilt, I will hire thee, and thou shalt go along with me as my servant. Yet we will have all things in common betwixt thee and me: only go along with me.

Mercy: But how shall I be ascertained that I also should be entertained? Had I this hope but from one that can tell, I would make no stick at all, but would go, being helped by Him that can help, though the way was never so tedious.

Christiana: Well, loving Mercy, I will tell thee what thou shalt do: go with me to the Wicket-gate, and there I will further inquire for thee; and if there thou shalt not meet with encouragement, I will be content that thou return to thy place: I will also pay thee for thy kindness which thou showest to me and my children, in the accompanying of us in the way that thou dost.

Mercy: Then will I go thither, and will take what shall follow; and the Lord grant that my lot may there fall, even as the King of heaven shall have his heart upon me.

Christiana then was glad at heart, not only that she had a companion, but also for that she had prevailed with this poor maid to fall in love with her own salvation. So they went on together, and Mercy began to weep. Then said Christiana, Wherefore weepeth my sister so?

Mercy: Alas! said she, who can but lament, that shall but rightly consider what a state and condition my poor relations are in, that yet remain in our sinful town? And that which makes my grief the more heavy is, because they have no instructor, nor any to tell them what is to come.

Christiana: Pity becomes pilgrims; and thou dost weep for thy friends, as my good Christian did for me when he left me: he mourned for that I would not heed nor regard him; but his Lord and ours did gather up his tears, and put them into his bottle; and now both I and thou, and these my sweet babes, are reaping the fruit and benefit of them. I hope, Mercy, that these tears of thine will not be lost; for the truth hath said, that ``they that sow in tears shall reap in joy." And ``he that goeth forth and weepeth, bearing precious seed, shall doubtless come again with rejoicing, bringing his sheaves with him." Psa. 126:5,6.

Then said Mercy,
\begin{verse} 
``Let the Most Blessed be my guide,\\
If it be his blessed will,\\
Unto his gate, into his fold,\\
Up to his holy hill.\\ 
And let him never suffer me\\
To swerve, or turn aside\\
From his free-grace and holy ways,\\
Whate'er shall me betide.\\
And let him gather them of mine\\
That I have left behind;\\
Lord, make them pray they may be thine,\\
With all their heart and mind."\\
\end{verse}

Now my old friend proceeded, and said, But when Christiana came to the Slough of Despond, she began to be at a stand; For, said she, this is the place in which my dear husband had like to have been smothered with mud. She perceived, also, that notwithstanding the command of the King to make this place for pilgrims good, yet it was rather worse than formerly. So I asked if that was true. Yes, said the old gentleman, too true; for many there be that pretend to be the King's laborers, and that say they are for mending the King's highways, who bring dirt and dung instead of stones, and so mar instead of mending. Here Christiana therefore, with her boys, did make a stand. But said Mercy, Come, let us venture; only let us be wary. Then they looked well to their steps, and made a shift to get staggering over.

Yet Christiana had like to have been in, and that not once or twice. Now they had no sooner got over, but they thought they heard words that said unto them, ``Blessed is she that believeth; for there shall be a performance of those things which were told her from the Lord." Luke 1:45.

Then they went on again; and said Mercy to Christiana, had I as good ground to hope for a loving reception at the Wicket-gate as you, I think no Slough of Despond would discourage me.

Well, said the other, you know your sore, and I know mine; and, good friend, we shall all have enough evil before we come to our journey's end. For can it be imagined that the people who design to attain such excellent glories as we do, and who are so envied that happiness as we are, but that we shall meet with what fears and snares, with what troubles and afflictions they can possibly assault us with that hate us?

And now Mr. Sagacity left me to dream out my dream by myself. Wherefore, methought I saw Christiana, and Mercy, and the boys, go all of them up to the gate: to which, when they were come, they betook themselves to a short debate about how they must manage their calling at the gate, and what should be said unto him that did open to them: so it was concluded, since Christiana was the eldest, that she should knock for entrance, and that she should speak to him that did open, for the rest. So Christiana began to knock, and as her poor husband did, she knocked and knocked again. But instead of any that answered, they all thought they heard as if a dog came barking upon them; a dog, and a great one too; and this made the women and children afraid. Nor durst they for a while to knock any more, for fear the mastiff should fly upon them. Now, therefore, they were greatly tumbled up and down in their minds, and knew not what to do: knock they durst not, for fear of the dog; go back they durst not, for fear the keeper of that gate should espy them as they so went, and should be offended with them; at last they thought of knocking again, and knocked more vehemently than they did at first. Then said the keeper of the gate, Who is there? So the dog left off to bark, and he opened unto them.

Then Christiana made low obeisance, and said, Let not our Lord be offended with his handmaidens, for that we have knocked at his princely gate. Then said the keeper, Whence come ye? And what is it that you would have?

Christiana answered, We are come from whence Christian did come, and upon the same errand as he; to wit, to be, if it shall please you, graciously admitted by this gate into the way that leads unto the Celestial City. And I answer, my Lord, in the next place, that I am Christiana, once the wife of Christian, that now is gotten above.

With that the keeper of the gate did marvel, saying, What, is she now become a pilgrim that but a while ago abhorred that life? Then she bowed her head, and said, Yea; and so are these my sweet babes also.

Then he took her by the hand and led her in, and said also, Suffer little children to come unto me; and with that he shut up the gate. This done, he called to a trumpeter that was above, over the gate, to entertain Christiana with shouting, and the sound of trumpet for joy. So he obeyed, and sounded, and filled the air with his melodious notes.

Now all this while poor Mercy did stand without, trembling and crying, for fear that she was rejected. But when Christiana had got admittance for herself and her boys, then she began to make intercession for Mercy.

Christiana: And she said, My Lord, I have a companion that stands yet without, that is come hither upon the same account as myself: one that is much dejected in her mind, for that she comes, as she thinks, without sending for; whereas I was sent for by my husband's King to come.

Now Mercy began to be very impatient, and each minute was as long to her as an hour; wherefore she prevented Christiana from a fuller interceding for her, by knocking at the gate herself. And she knocked then so loud that she made Christiana to start. Then said the keeper of the gate, Who is there? And Christiana said, It is my friend.

So he opened the gate, and looked out, but Mercy was fallen down without in a swoon, for she fainted, and was afraid that no gate should be opened to her.

Then he took her by the hand, and said, Damsel, I bid thee arise.

Oh, sir, said she, I am faint; there is scarce life left in me. But he answered, that one once said, ``When my soul fainted within me I remembered the Lord: and my prayer came unto thee, into thy holy temple." Jonah 2:7. Fear not, but stand upon thy feet, and tell me wherefore thou art come.

Mercy: I am come for that unto which I was never invited, as my friend Christiana was. Hers was from the King, and mine was but from her. Wherefore I fear I presume.

Keep: Did she desire thee to come with her to this place?

Mercy: Yes; and, as my Lord sees, I am come. And if there is any grace and forgiveness of sins to spare, I beseech that thy poor handmaid may be a partaker thereof.

Then he took her again by the hand, and led her gently in, and said, I pray for all them that believe on me, by what means soever they come unto me. Then said he to those that stood by, Fetch something and give it to Mercy to smell on, thereby to stay her faintings; so they fetched her a bundle of myrrh, and a while after she was revived.

And now were Christiana and her boys, and Mercy, received of the Lord at the head of the way, and spoken kindly unto by him. Then said they yet further unto him, We are sorry for our sins, and beg of our Lord his pardon, and further information what we must do.

I grant pardon, said he, by word and deed; by word in the promise of forgiveness, by deed in the way I obtained it. Take the first from my lips with a kiss, and the other as it shall be revealed. Song 1:2; John 20:20.

Now I saw in my dream, that he spake many good words unto them, whereby they were greatly gladdened. He also had them up to the top of the gate, and showed them by what deed they were saved; and told them withal, that that sight they would have again as they went along in the way, to their comfort.

So he left them awhile in a summer parlor below, where they entered into talk by themselves; and thus Christiana began. O how glad am I that we are got in hither.

Mercy: So you well may; but I, of all, have cause to leap for joy.

Christiana: I thought one time, as I stood at the gate, because I had knocked and none did answer, that all our labor had been lost, especially when that ugly cur made such a heavy barking against us.

Mercy: But my worst fear was after I saw that you was taken into his favor, and that I was left behind. Now, thought I, it is fulfilled which is written, ``Two women shall be grinding at the mill; the one shall be taken, and the other left." Matt. 24:41. I had much ado to forbear crying out, Undone! And afraid I was to knock any more; but when I looked up to what was written over the gate, I took courage. I also thought that I must either knock again, or die; so I knocked, but I cannot tell how, for my spirit now struggled between life and death.

Christiana: Can you not tell how you knocked? I am sure your knocks were so earnest that the very sound of them made me start; I thought I never heard such knocking in all my life; I thought you would come in by a violent hand, or take the kingdom by storm. Matt. 11:12.

Mercy: Alas! to be in my case, who that so was could but have done so? You saw that the door was shut upon me, and there was a most cruel dog thereabout. Who, I say, that was so faint-hearted as I, would not have knocked with all their might? But pray, what said my Lord to my rudeness? Was he not angry with me?

Christiana: When he heard your lumbering noise, he gave a wonderful innocent smile; I believe what you did pleased him well, for he showed no sign to the contrary. But I marvel in my heart why he keeps such a dog: had I known that before, I should not have had heart enough to have ventured myself in this manner. But now we are in, we are in, and I am glad with all my heart.

Mercy: I will ask, if you please, next time he comes down, why he keeps such a filthy cur in his yard; I hope he will not take it amiss.

Do so, said the children, and persuade him to hang him; for we are afraid he will bite us when we go hence.

So at last he came down to them again, and Mercy fell to the ground on her face before him, and worshiped, and said, ``Let my Lord accept the sacrifice of praise which I now offer unto him with the calves of my lips."

So he said unto her, Peace be to thee; stand up. But she continued upon her face, and said, ``Righteous art thou, O Lord, when I plead with thee; yet let me talk with thee of thy judgments." Jer. 12:1. Wherefore dost thou keep so cruel a dog in thy yard, at the sight of which such women and children as we are ready to fly from thy gate for fear?

He answered and said, That dog has another owner; he also is kept close in another man's ground, only my pilgrims hear his barking; he belongs to the castle which you see there at a distance, but can come up to the walls of this place. He has frighted many an honest pilgrim from worse to better, by the great voice of his roaring. Indeed, he that owneth him doth not keep him out of any good-will to me or mine, but with intent to keep the pilgrims from coming to me, and that they may be afraid to come and knock at this gate for entrance. Sometimes also he has broken out, and has worried some that I loved; but I take all at present patiently. I also give my pilgrims timely help, so that they are not delivered to his power, to do with them what his doggish nature would prompt him to. But what my purchased one, I trow, hadst thou known never so much beforehand, thou wouldest not have been afraid of a dog. The beggars that go from door to door, will, rather than lose a supposed alms, run the hazard of the bawling, barking, and biting too of a dog; and shall a dog, a dog in another man's yard, a dog whose barking I turn to the profit of pilgrims, keep any from coming to me? I deliver them from the lions, and my darling from the power of the dog. Psa. 22:21,22.

Mercy: Then said Mercy, I confess my ignorance; I spake what I understood not; I acknowledge that thou doest all things well.

Christiana: Then Christiana began to talk of their journey, and to inquire after the way. So he fed them and washed their feet, and set them in the way of his steps, according as he had dealt with her husband before. 
